---
title: "Session 2: Working with Data Structures"
subtitle: "R Programming Workbook"
author: "R Programming Course"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-tools: true
    code-overflow: wrap
    theme: cosmo
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-note title="Learning Objectives"}
By the end of this session, you will be able to:

**Part 1: Indexing and Subsetting**

- Subset vectors and data frames by position, logic, and name
- Combine conditions with `&`, `|`, and `%in%`
- Replace values conditionally

**Part 2: Control Flow**

- Write if/else statements for scalar conditions
- Use `ifelse()` for vectorized conditions
- Build for and while loops with proper pre-allocation

**Part 3: Functions**

- Write functions with default arguments and input validation
- Understand lexical scoping
- Use lambda syntax `\(x)` and closures
:::

```{r}
students <- data.frame(
  name = c("Alice", "Bob", "Carlos", "Diana", "Elena", "Frank"),
  age = c(21, 23, 22, 20, 24, 21),
  major = c("CS", "Math", "CS", "Stats", "CS", "Math"),
  gpa = c(3.8, 3.2, 3.5, 3.9, 3.1, 3.6),
  year = c(3, 4, 3, 2, 4, 3)
)
```

## Part 1: Indexing and Subsetting

### 1.1 Numeric Indexing

Select by position. R is 1-indexed.

```{r}
x <- c(10, 20, 30, 40, 50)
x[1]           # first element
x[c(1, 3, 5)] # multiple positions
x[2:4]        # range
x[-1]          # all except first
x[-c(2, 4)]   # all except 2nd and 4th
```

Negative indices exclude. You cannot mix positive and negative in the same bracket.

### 1.2 Logical Indexing

Select by condition:

```{r}
x <- c(10, 20, 30, 40, 50)
x[x > 25]
x[x >= 20 & x <= 40]
```

The condition produces a TRUE/FALSE vector the same length as `x`.

### 1.3 Name-Based Indexing

```{r}
grades <- c(alice = 90, bob = 85, carlos = 92)
grades["alice"]
grades[c("bob", "carlos")]
```

### 1.4 Data Frame Subsetting

```{r}
# Rows by condition
students[students$major == "CS", ]

# Specific columns
students[, c("name", "gpa")]

# Both
students[students$gpa > 3.5, c("name", "gpa", "major")]

# Using %in% for multiple values
students[students$major %in% c("CS", "Stats"), ]
```

### 1.5 Conditional Replacement

```{r}
x <- c(1, -2, 3, -4, 5)
x[x < 0] <- 0
x
```

### Your Turn: Exercise 1.1

```{r}
temps <- c(72, 65, 80, 55, 90, 78, 62, 85, 70, 95)
temps[3]
temps[5]
temps[8]
temps[temps>75]
temps[temps<65] <- 65
temps[temps == 80]
```

1. Extract the 3rd, 5th, and 8th temperatures
2. Get all temperatures above 75
3. Replace any temperature below 65 with 65 (a floor)
4. Find the positions of temperatures above 80


```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
temps[c(3, 5, 8)]
temps[temps > 75]
temps[temps < 65] <- 65
which(temps > 80)
```

### Your Turn: Exercise 1.2

Using the `students` data frame:

1. Find students who are juniors (year 3) with GPA above 3.4
2. Get just the names of all Math majors
3. Find the student with the lowest GPA
4. Create a subset of only CS majors, sorted by GPA descending

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
# 1
students[students$year == 3 & students$gpa > 3.4, ]

# 2
students$name[students$major == "Math"]

# 3
students[which.min(students$gpa), ]

# 4
cs <- students[students$major == "CS", ]
cs[order(cs$gpa, decreasing = TRUE), ]
```

::: {.callout-warning title="Common Mistake"}
**Forgetting the comma in data frame subsetting.**

```{r}
#| error: true
students[students$major == "CS"]  # missing comma!
```

For data frames, you need `[rows, columns]`. Without the comma, R tries to subset columns by the logical vector, which gives unexpected results or errors.

Correct: `students[students$major == "CS", ]`
:::

### Your Turn: Exercise 1.3

```{r}
products <- data.frame(
  item = c("laptop", "mouse", "keyboard", "monitor", "headset",
           "webcam", "speaker", "cable"),
  price = c(999, 25, 75, 400, 150, 80, 120, 10),
  category = c("computer", "accessory", "accessory", "computer",
               "accessory", "accessory", "audio", "accessory")
)
```

1. Find all accessories that cost more than $50
2. What is the average price of items in the "computer" category?
3. Replace any price below $20 with $20 (minimum price policy)
4. Find items that are NOT in the "accessory" category

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
# 1
products[products$category == "accessory" & products$price > 50, ]

# 2
mean(products$price[products$category == "computer"])

# 3
products$price[products$price < 20] <- 20

# 4
products[products$category != "accessory", ]
# or: products[!products$category %in% "accessory", ]
```

---

## Part 2: Control Flow

### 2.1 if/else

```{r}
score <- 78

if (score >= 90) {
  grade <- "A"
} else if (score >= 80) {
  grade <- "B"
} else if (score >= 70) {
  grade <- "C"
} else {
  grade <- "F"
}
grade
```

`if` evaluates a single TRUE/FALSE value. It does not work on vectors.

### 2.2 Vectorized ifelse()

```{r}
scores <- c(85, 92, 67, 73, 95, 58)
ifelse(scores >= 70, "pass", "fail")
```

Nested `ifelse()` for multiple categories:

```{r}
ifelse(scores >= 90, "A",
ifelse(scores >= 80, "B",
ifelse(scores >= 70, "C", "F")))
```

### 2.3 for Loops

```{r}
for (i in 1:5) {
  cat("Square of", i, "is", i^2, "\n")
}
```

Use `seq_along()` instead of `1:length()`:

```{r}
x <- c(10, 20, 30)
for (i in seq_along(x)) {
  cat("Element", i, ":", x[i], "\n")
}
```

### 2.4 Pre-allocating Output

```{r}
# Good: pre-allocate
n <- 10
result <- numeric(n)
for (i in 1:n) {
  result[i] <- i^2 + i
}
result
```

```{r}
# Bad: growing a vector (slow for large n)
result_bad <- c()
for (i in 1:n) {
  result_bad <- c(result_bad, i^2 + i)
}
result_bad
```

Both produce the same result, but the pre-allocated version is much faster for large n.

### 2.5 while Loops

```{r}
# Find the smallest power of 2 greater than 1000
x <- 1
while (x <= 1000) {
  x <- x * 2
}
x
```

### 2.6 break and next

```{r}
# Print odd numbers up to 15, stop at 11
for (i in 1:15) {
  if (i %% 2 == 0) next    # skip even numbers
  if (i > 11) break        # stop after 11
  cat(i, "")
}
```

### Your Turn: Exercise 2.1

Write a loop that iterates through numbers 1 to 20 and prints:

- "fizz" for multiples of 3
- "buzz" for multiples of 5
- "fizzbuzz" for multiples of both
- the number itself otherwise

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
for (i in 1:20) {
  if (i %% 15 == 0) {
    cat("fizzbuzz\n")
  } else if (i %% 3 == 0) {
    cat("fizz\n")
  } else if (i %% 5 == 0) {
    cat("buzz\n")
  } else {
    cat(i, "\n")
  }
}
```

### Your Turn: Exercise 2.2

Write a `while` loop that simulates a random walk. Start at position 0. Each step, add a random value from `sample(c(-1, 1), 1)`. Stop when you reach +10 or -10. Print the final position and how many steps it took.

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
set.seed(42)
position <- 0
steps <- 0

while (abs(position) < 10) {
  position <- position + sample(c(-1, 1), 1)
  steps <- steps + 1
}

cat("Final position:", position, "\n")
cat("Steps taken:", steps, "\n")
```

### Your Turn: Exercise 2.3

Use `ifelse()` to add a `gpa_category` column to the `students` data frame:

- "high" if GPA >= 3.7
- "medium" if GPA >= 3.3
- "low" otherwise

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
students$gpa_category <- ifelse(students$gpa >= 3.7, "high",
                         ifelse(students$gpa >= 3.3, "medium", "low"))
students[, c("name", "gpa", "gpa_category")]
```

::: {.callout-warning title="Common Mistake"}
**Using `if` instead of `ifelse()` on a vector.**

```{r}
scores <- c(85, 92, 67)
if (scores > 80) "high" else "low"  # only checks first element!
```

`if` looks at only the first element. Use `ifelse()` for element-wise checks:

```{r}
ifelse(scores > 80, "high", "low")
```
:::

::: {.callout-tip title="Challenge"}
Write a for loop that computes the Fibonacci sequence up to the 20th term. Store results in a pre-allocated vector.

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
fib <- numeric(20)
fib[1] <- 1
fib[2] <- 1
for (i in 3:20) {
  fib[i] <- fib[i - 1] + fib[i - 2]
}
fib
```
:::

---

## Part 3: Writing Functions

### 3.1 Basic Function Structure

```{r}
square <- function(x) {
  x^2
}
square(4)
square(1:5)  # works on vectors because ^2 is vectorized
```

The last expression is the return value. No explicit `return()` needed (but you can use it).

### 3.2 Multiple Arguments and Defaults

```{r}
bmi <- function(weight_kg, height_m = 1.75) {
  weight_kg / height_m^2
}
bmi(70)
bmi(70, 1.80)
```

### 3.3 Input Validation with stop()

```{r}
safe_divide <- function(a, b) {
  if (!is.numeric(a) || !is.numeric(b)) {
    stop("Both arguments must be numeric")
  }
  if (any(b == 0)) {
    stop("Cannot divide by zero")
  }
  a / b
}
safe_divide(10, 3)
```

```{r}
#| error: true
safe_divide(10, 0)
```

`stop()` throws an error with your message. Always validate inputs in functions you plan to reuse.

### 3.4 Returning Multiple Values

```{r}
stats <- function(x) {
  list(
    mean = mean(x, na.rm = TRUE),
    sd = sd(x, na.rm = TRUE),
    min = min(x, na.rm = TRUE),
    max = max(x, na.rm = TRUE),
    n = sum(!is.na(x))
  )
}
result <- stats(c(10, 20, NA, 40, 50))
result$mean
result$n
```

### 3.5 Lambda Syntax

For short, one-off functions:

```{r}
sapply(1:5, \(x) x^2 + x)
```

`\(x) x^2 + x` is equivalent to `function(x) x^2 + x`.

### 3.6 Scope and Closures

```{r}
make_counter <- function() {
  count <- 0
  list(
    increment = function() {
      count <<- count + 1  # <<- modifies parent scope
      count
    },
    get = function() count
  )
}

counter <- make_counter()
counter$increment()
counter$increment()
counter$increment()
counter$get()
```

The `<<-` operator assigns to the parent environment. Use it sparingly.

### Your Turn: Exercise 3.1

Write a function called `temp_convert` that converts between Fahrenheit and Celsius. It should:

- Take a temperature value and a `to` argument ("C" or "F")
- Default `to` to "C"
- Return the converted value
- Stop with an error if `to` is not "C" or "F"

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
temp_convert <- function(temp, to = "C") {
  if (!to %in% c("C", "F")) {
    stop("'to' must be 'C' or 'F'")
  }
  if (to == "C") {
    (temp - 32) * 5/9
  } else {
    temp * 9/5 + 32
  }
}
temp_convert(100)      # 100F to C
temp_convert(37, "F")  # 37C to F
```

### Your Turn: Exercise 3.2

Write a function `describe_column` that takes a data frame and a column name (as a string), and returns a named list with: count, mean, min, max, and number of NAs. Test it on the `students` data frame with the "gpa" column.

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
describe_column <- function(df, col_name) {
  x <- df[[col_name]]
  if (!is.numeric(x)) stop("Column must be numeric")
  
  list(
    count = length(x),
    mean = mean(x, na.rm = TRUE),
    min = min(x, na.rm = TRUE),
    max = max(x, na.rm = TRUE),
    na_count = sum(is.na(x))
  )
}
describe_column(students, "gpa")
```

### Your Turn: Exercise 3.3

Write a function `create_multiplier` that takes a number `n` and returns a function that multiplies its input by `n`. This is a closure.

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
create_multiplier <- function(n) {
  \(x) x * n
}

double <- create_multiplier(2)
triple <- create_multiplier(3)

double(5)   # 10
triple(5)   # 15
double(1:5) # 2 4 6 8 10
```

::: {.callout-warning title="Common Mistake"}
**Forgetting that R copies on modify.**

```{r}
add_column <- function(df) {
  df$new_col <- 1
  df
}

result <- add_column(students)
"new_col" %in% names(students)  # FALSE - original unchanged
"new_col" %in% names(result)    # TRUE - modified copy returned
```

Passing a data frame to a function does not modify the original. The function works on a copy. You must assign the result back if you want the changes.
:::

::: {.callout-tip title="Challenge"}
Write a function `apply_to_numeric` that takes a data frame and a function, and applies that function to every numeric column. Return a named list of results.

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
apply_to_numeric <- function(df, func) {
  numeric_cols <- names(df)[sapply(df, is.numeric)]
  result <- list()
  for (col in numeric_cols) {
    result[[col]] <- func(df[[col]])
  }
  result
}

apply_to_numeric(students, mean)
apply_to_numeric(students, \(x) round(sd(x), 2))
```
:::

---

## Session Synthesis

### Connect the Dots

Build a complete grade analysis workflow:

1. Create a data frame with 10 students: name, exam1 (0-100), exam2 (0-100), homework (0-100)
2. Write a function `weighted_average` that computes: 40% exam1 + 40% exam2 + 20% homework
3. Add a `final_score` column using your function
4. Write a function `assign_letter` that converts a score to a letter grade (A/B/C/D/F)
5. Add a `grade` column using `sapply()` and your letter grade function
6. Find all students who passed (grade C or above)
7. Calculate the class average final score

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
# 1
set.seed(100)
class_data <- data.frame(
  name = paste0("Student_", 1:10),
  exam1 = sample(50:100, 10, replace = TRUE),
  exam2 = sample(50:100, 10, replace = TRUE),
  homework = sample(60:100, 10, replace = TRUE)
)

# 2
weighted_average <- function(e1, e2, hw) {
  0.4 * e1 + 0.4 * e2 + 0.2 * hw
}

# 3
class_data$final_score <- weighted_average(
  class_data$exam1, class_data$exam2, class_data$homework
)

# 4
assign_letter <- function(score) {
  if (score >= 90) "A"
  else if (score >= 80) "B"
  else if (score >= 70) "C"
  else if (score >= 60) "D"
  else "F"
}

# 5
class_data$grade <- sapply(class_data$final_score, assign_letter)

# 6
class_data[class_data$grade %in% c("A", "B", "C"), ]

# 7
mean(class_data$final_score)
```

### Key Takeaways

1. Logical indexing is the most powerful way to filter data: `df[condition, ]`
2. Use `ifelse()` for vectors, `if`/`else` for single values
3. Pre-allocate output vectors before filling them in loops
4. Functions return their last expression. Use named lists for multiple return values.
5. `\(x)` is the modern lambda syntax for short anonymous functions.

### Prepare for Next Session

**Session 3: Patterns and Output**

We will cover the apply family (replacing loops with cleaner code), string handling, and base R plotting. Make sure you are comfortable writing functions — the apply functions pass functions as arguments.
