---
title: "Session 10: Real-World Data Work"
subtitle: "R Programming Workbook"
author: "R Programming Course"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-tools: true
    code-overflow: wrap
    theme: cosmo
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-note title="Learning Objectives"}
By the end of this session you will be able to:

1. Scrape data from HTML pages using rvest
2. Use CSS selectors to target specific page elements
3. Extract text, attributes, and tables from web pages
4. Apply the full tidyverse toolkit to solve multi-step analysis problems
5. Build an end-to-end data pipeline from messy input to clean output
6. Handle common data quality issues (inconsistent text, mixed date formats, type mismatches)
7. Wrap a pipeline into a reusable function
8. Combine dplyr, tidyr, stringr, lubridate, forcats, purrr, and ggplot2 in a single workflow
:::

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(lubridate)
library(purrr)
library(ggplot2)
library(rvest)
library(jsonlite)

set.seed(42)
sales <- tibble(
  date = rep(seq(as.Date("2024-01-01"), as.Date("2024-12-31"), by = "day"), 3),
  region = rep(c("North", "South", "West"), each = 366),
  product = sample(c("Widget", "Gadget", "Doohickey"), 366 * 3, replace = TRUE),
  units = rpois(366 * 3, lambda = 50),
  revenue = round(runif(366 * 3, 10, 500), 2),
  returned = rbinom(366 * 3, 1, 0.05)
)
```

## Part 1: Web Scraping with rvest

### HTML and CSS Selectors

rvest works by reading HTML and extracting elements using CSS selectors:

| Selector | Meaning |
|----------|---------|
| `h1` | All `<h1>` tags |
| `.classname` | Elements with that class |
| `#idname` | Element with that ID |
| `table td` | `<td>` elements inside `<table>` |
| `div > p` | `<p>` that are direct children of `<div>` |

### Reading and Selecting

```{r}
html_page <- read_html('
<html><body>
  <h1 class="main-title">Course Catalog</h1>
  <div class="course">
    <h2>Statistics 101</h2>
    <span class="credits">3 credits</span>
    <p class="desc">Introduction to statistical thinking.</p>
  </div>
  <div class="course">
    <h2>Data Science 200</h2>
    <span class="credits">4 credits</span>
    <p class="desc">Applied data analysis with R and Python.</p>
  </div>
  <div class="course">
    <h2>Machine Learning 300</h2>
    <span class="credits">4 credits</span>
    <p class="desc">Supervised and unsupervised learning methods.</p>
  </div>
</body></html>')

# Extract course names
html_page |> html_elements(".course h2") |> html_text2()
```

### Building a Tibble from Scraped Data

```{r}
courses <- tibble(
  name = html_page |> html_elements(".course h2") |> html_text2(),
  credits = html_page |> html_elements(".credits") |> html_text2() |>
    str_extract("\\d+") |> as.integer(),
  description = html_page |> html_elements(".desc") |> html_text2()
)

courses
```

Extract each field separately, then combine into a tibble.

### Extracting Tables

```{r}
table_html <- read_html('
<table id="grades">
  <tr><th>Student</th><th>Midterm</th><th>Final</th></tr>
  <tr><td>Alice</td><td>92</td><td>88</td></tr>
  <tr><td>Bob</td><td>78</td><td>85</td></tr>
  <tr><td>Carlos</td><td>95</td><td>91</td></tr>
</table>')

table_html |> html_table() |> pluck(1)
```

`html_table()` converts HTML tables to data frames directly.

### Extracting Links

```{r}
link_html <- read_html('
<ul class="resources">
  <li><a href="https://r-project.org">R Project</a></li>
  <li><a href="https://tidyverse.org">Tidyverse</a></li>
  <li><a href="https://quarto.org">Quarto</a></li>
</ul>')

tibble(
  text = link_html |> html_elements(".resources a") |> html_text2(),
  url = link_html |> html_elements(".resources a") |> html_attr("href")
)
```

::: {.callout-note title="Base R vs Tidyverse"}
Base R does not have built-in web scraping tools. The xml2 package (which rvest builds on) provides the low-level XML/HTML parsing. rvest adds the user-friendly `html_elements()`, `html_text2()`, and `html_table()` wrappers.
:::

### Your Turn: Exercise 1

Extract the product names and prices from this HTML:

```{r}
shop_html <- read_html('
<div class="shop">
  <div class="item"><span class="name">Laptop</span><span class="price">$999</span></div>
  <div class="item"><span class="name">Mouse</span><span class="price">$29</span></div>
  <div class="item"><span class="name">Keyboard</span><span class="price">$79</span></div>
  <div class="item"><span class="name">Monitor</span><span class="price">$349</span></div>
</div>')
```

Build a tibble with columns `name` (character) and `price` (numeric).

```{r}
#| code-fold: true
#| code-summary: "Solution"
tibble(
  name = shop_html |> html_elements(".name") |> html_text2(),
  price = shop_html |> html_elements(".price") |> html_text2() |>
    str_remove("\\$") |> as.numeric()
)
```

### Your Turn: Exercise 2

Extract the table from this HTML and compute each student's average score:

```{r}
score_html <- read_html('
<table>
  <tr><th>Name</th><th>Quiz1</th><th>Quiz2</th><th>Quiz3</th></tr>
  <tr><td>Diana</td><td>85</td><td>90</td><td>88</td></tr>
  <tr><td>Elena</td><td>92</td><td>87</td><td>95</td></tr>
  <tr><td>Frank</td><td>78</td><td>82</td><td>80</td></tr>
</table>')
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
scores <- score_html |> html_table() |> pluck(1)

scores |>
  mutate(across(starts_with("Quiz"), as.numeric)) |>
  rowwise() |>
  mutate(average = mean(c_across(starts_with("Quiz")))) |>
  ungroup()
```

### Your Turn: Exercise 3

Extract all links from this HTML and filter to only external links (those starting with "http"):

```{r}
nav_html <- read_html('
<nav>
  <a href="/">Home</a>
  <a href="/about">About</a>
  <a href="https://github.com/example">GitHub</a>
  <a href="/contact">Contact</a>
  <a href="https://docs.example.com">Documentation</a>
</nav>')
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
all_links <- tibble(
  text = nav_html |> html_elements("a") |> html_text2(),
  url = nav_html |> html_elements("a") |> html_attr("href")
)

all_links |> filter(str_starts(url, "http"))
```

::: {.callout-warning title="Common Mistake"}
Using `html_text()` instead of `html_text2()`:

```r
# html_text() preserves internal whitespace/newlines (often messy)
# html_text2() simulates how text looks in a browser (cleaner)
```

Always prefer `html_text2()` unless you need the raw whitespace.
:::

::: {.callout-tip title="Challenge"}
Scrape this nested HTML and build a tibble with team name, wins, and losses. Compute win percentage.

```{r}
standings_html <- read_html('
<div class="standings">
  <div class="team">
    <h3>Eagles</h3>
    <span class="record"><span class="wins">10</span>-<span class="losses">4</span></span>
  </div>
  <div class="team">
    <h3>Hawks</h3>
    <span class="record"><span class="wins">8</span>-<span class="losses">6</span></span>
  </div>
  <div class="team">
    <h3>Bears</h3>
    <span class="record"><span class="wins">12</span>-<span class="losses">2</span></span>
  </div>
</div>')
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
tibble(
  team = standings_html |> html_elements(".team h3") |> html_text2(),
  wins = standings_html |> html_elements(".wins") |> html_text2() |> as.integer(),
  losses = standings_html |> html_elements(".losses") |> html_text2() |> as.integer()
) |>
  mutate(win_pct = round(wins / (wins + losses), 3))
```
:::

## Part 2: Tidyverse Cumulative Practice

### Practice Problem 1: Seasonal Analysis

Find the season (Spring: Mar-May, Summer: Jun-Aug, Fall: Sep-Nov, Winter: Dec-Feb) with the highest average daily revenue.

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  mutate(
    month_num = month(date),
    season = case_when(
      month_num %in% 3:5 ~ "Spring",
      month_num %in% 6:8 ~ "Summer",
      month_num %in% 9:11 ~ "Fall",
      TRUE ~ "Winter"
    )
  ) |>
  summarize(avg_daily_revenue = mean(revenue), .by = season) |>
  arrange(desc(avg_daily_revenue))
```

### Practice Problem 2: Return Rate Heatmap

Create a heatmap of return rate by month and region.

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| fig-width: 8
#| fig-height: 4
sales |>
  mutate(month_name = month(date, label = TRUE)) |>
  summarize(return_rate = mean(returned), .by = c(month_name, region)) |>
  ggplot(aes(x = month_name, y = region, fill = return_rate)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red", labels = scales::percent) +
  labs(title = "Return Rate by Month and Region", x = NULL, y = NULL, fill = "Return Rate") +
  theme_minimal()
```

### Practice Problem 3: Product Performance Ranking

Rank products by total revenue within each region. Show the rank as a new column.

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  summarize(total_revenue = sum(revenue), .by = c(region, product)) |>
  mutate(rank = min_rank(desc(total_revenue)), .by = region) |>
  arrange(region, rank)
```

### Practice Problem 4: Rolling 7-Day Average

Compute a 7-day rolling average of daily revenue for the North region.

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| fig-width: 10
#| fig-height: 4
north_daily <- sales |>
  filter(region == "North") |>
  summarize(daily_revenue = sum(revenue), .by = date) |>
  arrange(date) |>
  mutate(
    rolling_avg = slider::slide_dbl(daily_revenue, mean, .before = 6, .complete = TRUE)
  )

ggplot(north_daily, aes(x = date)) +
  geom_line(aes(y = daily_revenue), alpha = 0.3) +
  geom_line(aes(y = rolling_avg), color = "blue", linewidth = 1) +
  labs(title = "North Region: Daily Revenue with 7-Day Rolling Average",
       x = NULL, y = "Revenue ($)") +
  theme_minimal()
```

### Practice Problem 5: Per-Group Models

Fit a linear model of revenue ~ units for each product. Report the slope and R-squared.

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  nest(.by = product) |>
  mutate(
    model = map(data, \(df) lm(revenue ~ units, data = df)),
    slope = map_dbl(model, \(m) coef(m)["units"]),
    r_squared = map_dbl(model, \(m) summary(m)$r.squared)
  ) |>
  select(product, slope, r_squared) |>
  arrange(desc(r_squared))
```

### Your Turn: Exercise 4

Find the day of the week with the highest total sales (units). Create a bar chart showing total units by day of week.

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| fig-width: 8
#| fig-height: 4
sales |>
  mutate(weekday = wday(date, label = TRUE)) |>
  summarize(total_units = sum(units), .by = weekday) |>
  ggplot(aes(x = weekday, y = total_units)) +
  geom_col(fill = "steelblue") +
  labs(title = "Total Units Sold by Day of Week", x = NULL, y = "Total Units") +
  theme_minimal()
```

### Your Turn: Exercise 5

For each region, find the month with the highest revenue and the month with the lowest revenue. Return a tibble with columns: region, best_month, worst_month.

```{r}
#| code-fold: true
#| code-summary: "Solution"
monthly <- sales |>
  mutate(month_name = month(date, label = TRUE)) |>
  summarize(total_rev = sum(revenue), .by = c(region, month_name))

best <- monthly |> slice_max(total_rev, n = 1, by = region) |> select(region, best_month = month_name)
worst <- monthly |> slice_min(total_rev, n = 1, by = region) |> select(region, worst_month = month_name)

inner_join(best, worst, by = "region")
```

::: {.callout-warning title="Common Mistake"}
Using `group_by()` without `ungroup()`:

```r
sales |>
  group_by(region) |>
  mutate(pct = revenue / sum(revenue))
# Next operation is still grouped! Can cause unexpected behavior.
```

Prefer `.by` in `summarize()` and `mutate()` to avoid forgotten groups.
:::

## Part 3: End-to-End Pipeline

### The Messy Dataset

```{r}
messy <- tibble(
  order_date = c("2024-01-15", "2024/02/20", "March 5, 2024", "2024-04-10",
                 "2024-05-22", "2024-06-30", NA, "2024-08-15"),
  customer = c("alice smith", "BOB JONES", "Carlos M.", " diana Lee",
               "Elena Wu", "FRANK o'brien", "alice smith", "Bob Jones"),
  item = c("Widget_Pro", "gadget-basic", "WIDGET PRO", "Doohickey Ultra",
           "widget_pro", "Gadget Basic", "DOOHICKEY_ULTRA", "Widget Pro"),
  quantity = c("5", "three", "2", "10", "7", "4", "6", "8"),
  amount = c(149.95, 29.97, NA, 199.90, 209.65, 59.96, 179.94, 239.92),
  status = c("shipped", "DELIVERED", "Shipped", "pending", "SHIPPED",
             "Delivered", "cancelled", "PENDING")
)

messy
```

### Your Turn: Exercise 6 (Pipeline Project)

Clean this dataset. Your final tibble should have:

- `order_date` as a proper Date
- `customer` in title case with trimmed whitespace
- `item` cleaned to consistent format (title case, spaces instead of underscores/hyphens)
- `quantity` as numeric (handle the text "three" -> NA or 3)
- `amount` with NAs handled
- `status` as a factor with levels: pending, shipped, delivered, cancelled

```{r}
#| code-fold: true
#| code-summary: "Solution"
word_to_num <- function(x) {
  words <- c(one = 1, two = 2, three = 3, four = 4, five = 5,
             six = 6, seven = 7, eight = 8, nine = 9, ten = 10)
  x_lower <- str_to_lower(x)
  ifelse(x_lower %in% names(words), words[x_lower], suppressWarnings(as.numeric(x)))
}

clean <- messy |>
  mutate(
    order_date = parse_date_time(order_date, orders = c("ymd", "y/m/d", "mdy")) |> as_date(),
    customer = str_trim(customer) |> str_to_title(),
    item = str_replace_all(item, "[_-]", " ") |> str_to_title(),
    quantity = word_to_num(quantity),
    amount = coalesce(amount, median(amount, na.rm = TRUE)),
    status = str_to_lower(status) |>
      factor(levels = c("pending", "shipped", "delivered", "cancelled"))
  ) |>
  filter(!is.na(order_date))

clean
```

### Your Turn: Exercise 7

Using your cleaned data from Exercise 6, answer:

1. How many orders per status?
2. Which customer has the highest total amount?
3. Which item is most popular by quantity?

```{r}
#| code-fold: true
#| code-summary: "Solution"
# 1. Orders per status
clean |> count(status)

# 2. Highest total by customer
clean |>
  summarize(total = sum(amount), .by = customer) |>
  slice_max(total, n = 1)

# 3. Most popular item
clean |>
  summarize(total_qty = sum(quantity, na.rm = TRUE), .by = item) |>
  slice_max(total_qty, n = 1)
```

### Your Turn: Exercise 8

Wrap your cleaning logic from Exercise 6 into a function called `clean_orders` that accepts a raw tibble and returns the cleaned version. Test it on the original `messy` data.

```{r}
#| code-fold: true
#| code-summary: "Solution"
clean_orders <- function(raw) {
  word_to_num <- function(x) {
    words <- c(one = 1, two = 2, three = 3, four = 4, five = 5,
               six = 6, seven = 7, eight = 8, nine = 9, ten = 10)
    x_lower <- str_to_lower(x)
    ifelse(x_lower %in% names(words), words[x_lower], suppressWarnings(as.numeric(x)))
  }

  raw |>
    mutate(
      order_date = parse_date_time(order_date, orders = c("ymd", "y/m/d", "mdy")) |> as_date(),
      customer = str_trim(customer) |> str_to_title(),
      item = str_replace_all(item, "[_-]", " ") |> str_to_title(),
      quantity = word_to_num(quantity),
      amount = coalesce(amount, median(amount, na.rm = TRUE)),
      status = str_to_lower(status) |>
        factor(levels = c("pending", "shipped", "delivered", "cancelled"))
    ) |>
    filter(!is.na(order_date))
}

clean_orders(messy)
```

### Think About It

You receive a CSV file every week with the same structure but different data. Your boss wants a weekly report with summary stats and charts. How would you structure your R project to handle this? What would you put in functions vs. scripts vs. Quarto documents?

::: {.callout-tip title="Challenge"}
Build a complete mini-pipeline: scrape the product data below, clean it, join it with order data, and produce a summary.

```{r}
product_html <- read_html('
<table id="catalog">
  <tr><th>Product</th><th>Category</th><th>Base Price</th></tr>
  <tr><td>Widget Pro</td><td>Hardware</td><td>$29.99</td></tr>
  <tr><td>Gadget Basic</td><td>Electronics</td><td>$9.99</td></tr>
  <tr><td>Doohickey Ultra</td><td>Accessories</td><td>$29.99</td></tr>
</table>')

order_data <- tibble(
  product = c("Widget Pro", "Gadget Basic", "Widget Pro",
              "Doohickey Ultra", "Gadget Basic", "Widget Pro"),
  quantity = c(5, 3, 2, 10, 7, 4),
  date = as.Date(c("2024-01-15", "2024-01-20", "2024-02-05",
                    "2024-02-10", "2024-03-01", "2024-03-15"))
)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
# Step 1: Scrape product catalog
catalog <- product_html |>
  html_table() |>
  pluck(1) |>
  as_tibble() |>
  mutate(
    `Base Price` = str_remove(`Base Price`, "\\$") |> as.numeric()
  ) |>
  rename(product = Product, category = Category, base_price = `Base Price`)

# Step 2: Join with orders
enriched_orders <- order_data |>
  left_join(catalog, by = "product") |>
  mutate(
    total = quantity * base_price,
    month = floor_date(date, "month")
  )

# Step 3: Summary
enriched_orders |>
  summarize(
    total_revenue = sum(total),
    total_units = sum(quantity),
    avg_order_value = mean(total),
    .by = c(category, month)
  ) |>
  arrange(month, category)
```
:::

## Session Synthesis

### Connect the Dots

Build a full pipeline that:

1. Scrapes this employee directory HTML
2. Cleans and standardizes the data
3. Joins with performance data
4. Produces a summary report with a visualization

```{r}
directory_html <- read_html('
<div class="directory">
  <div class="employee">
    <h3>Alice Chen</h3>
    <span class="dept">Engineering</span>
    <span class="hire-date">2020-03-15</span>
  </div>
  <div class="employee">
    <h3>Bob Martinez</h3>
    <span class="dept">Marketing</span>
    <span class="hire-date">2019-08-01</span>
  </div>
  <div class="employee">
    <h3>Carlos Lee</h3>
    <span class="dept">Engineering</span>
    <span class="hire-date">2021-01-10</span>
  </div>
  <div class="employee">
    <h3>Diana Patel</h3>
    <span class="dept">Marketing</span>
    <span class="hire-date">2022-06-20</span>
  </div>
</div>')

performance <- tibble(
  name = c("Alice Chen", "Bob Martinez", "Carlos Lee", "Diana Patel"),
  q1_score = c(92, 78, 88, 95),
  q2_score = c(88, 82, 91, 90),
  q3_score = c(95, 80, 85, 93),
  q4_score = c(90, 85, 89, 97)
)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| fig-width: 8
#| fig-height: 5

# 1. Scrape directory
employees <- tibble(
  name = directory_html |> html_elements(".employee h3") |> html_text2(),
  department = directory_html |> html_elements(".dept") |> html_text2(),
  hire_date = directory_html |> html_elements(".hire-date") |> html_text2() |> as_date()
)

# 2. Join with performance
full_data <- employees |>
  left_join(performance, by = "name") |>
  mutate(tenure_years = as.numeric(Sys.Date() - hire_date) / 365.25)

# 3. Compute yearly average
full_data <- full_data |>
  rowwise() |>
  mutate(avg_score = mean(c_across(starts_with("q")))) |>
  ungroup()

# 4. Summary
full_data |>
  summarize(
    n_employees = n(),
    avg_performance = mean(avg_score),
    avg_tenure = mean(tenure_years),
    .by = department
  )

# 5. Visualization
full_data |>
  pivot_longer(cols = starts_with("q"), names_to = "quarter", values_to = "score") |>
  mutate(quarter = str_remove(quarter, "_score") |> str_to_upper()) |>
  ggplot(aes(x = quarter, y = score, color = name, group = name)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  facet_wrap(~department) +
  labs(title = "Quarterly Performance by Department",
       x = "Quarter", y = "Score", color = "Employee") +
  theme_minimal()
```

### Key Takeaways

- rvest follows a read -> select -> extract pattern for web scraping.
- Real data requires cleaning. Expect inconsistent text, dates, types, and missing values.
- A good data pipeline is a function: raw data in, analysis-ready data out.
- You now have the full tidyverse toolkit: dplyr, tidyr, stringr, forcats, lubridate, purrr, ggplot2, rvest, httr2, jsonlite.

### Prepare for Next Session

Session 11 introduces Quarto. You will learn to turn your R analysis into polished HTML documents, PDFs, and Word files. Review how code chunks work (you have been using them in every workbook). Bring a short analysis you would like to turn into a report.
