---
title: "Session 14: Professional Practices"
subtitle: "R Programming Workbook — Bioinformatics Edition"
author: "R Programming Course"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-tools: true
    code-overflow: wrap
    theme: cosmo
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(here)
library(fs)
library(profvis)
```

## Learning Objectives

::: {.callout-note title="Learning Objectives"}
By the end of this session you will be able to:

1. Create and organize bioinformatics R projects with standard directory structure using the `fs` package
2. Use the `here` package to build portable file paths
3. Use `renv` to manage project-specific package dependencies, including Bioconductor packages
4. Orchestrate multi-step bioinformatics pipelines with `targets`
5. Debug R code with `traceback()`, `browser()`, and Positron breakpoints
6. Profile code performance with `system.time()` and `profvis`, and optimize with vectorization
7. Apply the tidyverse style guide for naming, spacing, and formatting
8. Check code style with `lintr` and auto-format with `styler`
9. Write clear comments and document functions with `roxygen2`
:::

## Part 1: Project Organization

### R Projects and Working Directories

An R project is a directory with a `.Rproj` file. Opening the `.Rproj` file sets the working directory to the project root.

```{r}
# Check your current working directory
getwd()
```

When you work inside a project, all relative paths resolve from the project root. This makes your code portable across machines.

### The `here` Package

`here()` builds file paths relative to the project root:

```{r}
library(here)

# Show the project root
here()

# Build a path to a gene counts data file
here("data", "raw", "gene_counts.csv")
```

`here()` finds the project root by looking for markers like `.Rproj`, `.here`, or `.git`. It works regardless of where the calling script lives in the directory tree.

Compare the approaches:

```{r}
#| eval: false

# Fragile: depends on working directory
read.csv("data/raw/gene_counts.csv")

# Fragile: absolute path, breaks on other machines
read.csv("C:/Users/alice/project/data/raw/gene_counts.csv")

# Robust: works from any script location
read.csv(here("data", "raw", "gene_counts.csv"))
```

### Your Turn: Project Structure

:::: {.callout-caution title="Your Turn — Project Structure"}
Design the directory structure for a bioinformatics project that:

- Processes RNA-seq data: raw FASTQ files, alignment, and gene counts
- Produces cleaned count matrices, summary tables, and visualizations
- Generates a Quarto report

Write the directory tree and explain what goes in each folder:

```{r}
#| eval: false

# Fill in your project structure:
project_structure <- '
rnaseq-analysis/
  rnaseq-analysis.Rproj
  README.md
  .gitignore
  _targets.R
  data/
    raw/
      fastq/        # Raw FASTQ files (never modify)
    processed/
      counts/       # Gene count matrices from featureCounts/STAR
  R/
    functions/      # Reusable R functions (filter, normalize, plot)
    scripts/        # Numbered analysis scripts
  output/
    figures/        # Saved plots (PCA, heatmaps, volcano plots)
    tables/         # DEG tables, summary statistics
  docs/
    report.qmd      # Final Quarto report
  renv.lock         # Package versions
'
cat(project_structure)
```
::::

### Project Setup with the `fs` Package

::: {.callout-tip title="Why `fs` instead of base R?"}
The `fs` package provides a type-safe, cross-platform interface for file system operations. Compared to base R:

- `fs::dir_create()` vs `dir.create()`: `fs` throws an **error** instead of a silent warning when paths are invalid; no need to set `showWarnings = FALSE`.
- Paths returned by `fs` functions are class `fs_path` — they print with colour in the console and compose safely with `fs::path()`.
- **Unicode-aware**: handles non-ASCII filenames on all platforms.
- **Works identically on Windows, macOS, and Linux** — no backslash vs forward-slash surprises.
- `fs::dir_create()` accepts a **vector** of paths and creates them all in one call, including all missing parents (`recursive = TRUE` is the default).

In 2026, `fs` is the standard for modern R workflows and is used by `usethis`, `targets`, and most Bioconductor tooling.
:::

```{r}
#| eval: false

library(fs)

setup_project <- function(project_name) {
  # Create all directories in one vectorized call
  dirs <- fs::path(project_name, c(
    "data/raw/fastq",
    "data/processed/counts",
    "R/functions",
    "R/scripts",
    "output/figures",
    "output/tables",
    "docs"
  ))
  fs::dir_create(dirs)   # recursive = TRUE by default

  # Create placeholder files
  fs::file_create(fs::path(project_name, c(
    ".gitignore",
    "README.md",
    "_targets.R"
  )))

  # Write starter content
  writeLines(
    paste0("# ", project_name, "\n\nProject description here."),
    fs::path(project_name, "README.md")
  )

  writeLines(
    c(".Rhistory", ".RData", ".Rproj.user/", "renv/library/"),
    fs::path(project_name, ".gitignore")
  )

  message("Project '", project_name, "' created at: ",
          fs::path_abs(project_name))
  invisible(fs::path_abs(project_name))
}

# Test it:
# setup_project("rnaseq-analysis")
```

:::: {.callout-caution title="Your Turn — Bioinformatics Project Skeleton"}
Use `fs` to create the standard bioinformatics project structure below. Your function should also create an empty `.gitignore`, `README.md`, and `_targets.R` at the project root.

```{r}
#| eval: false

# Target structure:
# project-name/
#   data/{raw,processed}/
#   R/{functions,scripts}/
#   output/{figures,tables}/
#   docs/
#   .gitignore
#   README.md
#   _targets.R

create_bio_skeleton <- function(name) {
  # YOUR CODE HERE
  # Hint: pass a character vector to fs::dir_create()
}
```

::::

::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| eval: false

library(fs)

create_bio_skeleton <- function(name) {
  # Create all directories in one call
  fs::dir_create(fs::path(name, c(
    "data/raw",
    "data/processed",
    "R/functions",
    "R/scripts",
    "output/figures",
    "output/tables",
    "docs"
  )))

  # Create root-level files
  fs::file_create(fs::path(name, c(".gitignore", "README.md", "_targets.R")))

  # Seed the README
  writeLines(
    c(paste0("# ", name), "", "## Overview", "RNA-seq analysis project."),
    fs::path(name, "README.md")
  )

  message("Skeleton created: ", fs::path_abs(name))
  invisible(fs::path_abs(name))
}

# create_bio_skeleton("rnaseq-pilot")
```
:::

### `renv` for Reproducible Environments

`renv` creates a project-local package library:

```{r}
#| eval: false

# Initialize renv (do this once per project)
renv::init()

# Install CRAN packages (goes to project library, not global)
install.packages("dplyr")

# Install Bioconductor packages through renv
renv::install("bioc::airway")  # Example Bioconductor dataset package

# After installing or updating packages, snapshot:
renv::snapshot()
# Creates/updates renv.lock with exact package versions

# On a new machine (or teammate's machine):
renv::restore()
# Installs exact versions from renv.lock
```

The `renv.lock` file records every package and its version:

```{r}
#| eval: false

# Example renv.lock content (JSON):
'{
  "R": {"Version": "4.4.1"},
  "Packages": {
    "dplyr": {
      "Package": "dplyr",
      "Version": "1.1.4",
      "Source": "Repository",
      "Repository": "CRAN"
    },
    "airway": {
      "Package": "airway",
      "Version": "1.22.0",
      "Source": "Bioconductor"
    }
  }
}'
```

::: {.callout-tip title="Cloud environments for collaborative work"}
For collaborative work or when your laptop cannot handle the data (large BAM files, high-memory alignment), **Posit Cloud** or **GitHub Codespaces** give you a cloud R environment with `renv` pre-configured. Both support restoring from `renv.lock` automatically. This is especially useful for shared course projects where everyone needs an identical software environment.
:::

### Your Turn: renv Workflow

:::: {.callout-caution title="Your Turn — renv Workflow"}
Put the following `renv` operations in the correct order for each scenario:

**Scenario 1: Starting a new project**

```{r}
#| eval: false

# Order these steps (number them 1-4):
# ___ renv::snapshot()
# ___ renv::install("bioc::airway")
# ___ renv::init()
# ___ git commit renv.lock

# Answer:
# 1. renv::init()
# 2. renv::install("bioc::airway")
# 3. renv::snapshot()
# 4. git commit renv.lock
```

**Scenario 2: Joining an existing project**

```{r}
#| eval: false

# Order these steps (number them 1-3):
# ___ Open the .Rproj file
# ___ renv::restore()
# ___ Clone the Git repository

# Answer:
# 1. Clone the Git repository
# 2. Open the .Rproj file
# 3. renv::restore()
```
::::

:::: {.callout-caution title="Your Turn — renv with Bioconductor"}
Walk through the complete `renv` workflow for a Bioconductor package:

1. Run `renv::init()` in a new project.
2. Install the `airway` package: `renv::install("bioc::airway")`.
3. Snapshot the environment: `renv::snapshot()`.
4. Open `renv.lock` and find the `airway` entry — what fields does it have?
5. Simulate a teammate restoring the environment: `renv::restore()`.

**Question:** What would happen if your teammate runs `install.packages("airway")` instead of using `renv::restore()`?

::::

::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| eval: false

# 1. Initialize
renv::init()

# 2. Install from Bioconductor
renv::install("bioc::airway")

# 3. Snapshot
renv::snapshot()

# 4. Inspect the lock file
renv_lock <- jsonlite::fromJSON("renv.lock")
renv_lock$Packages$airway
# Returns: Package, Version, Source = "Bioconductor", Repository

# 5. Restore
renv::restore()

# If the teammate uses install.packages("airway"):
# - install.packages() cannot find Bioconductor packages on CRAN.
# - The correct approach is renv::restore(), which reads the Source field
#   and routes the install to the right repository automatically.
```
:::

::: {.callout-warning title="Common Mistake"}
Forgetting to snapshot after installing a package:

```{r}
#| eval: false

# You install a package:
install.packages("scales")

# You forget to snapshot
# You commit and push

# Your teammate runs renv::restore()
# scales is NOT installed because it is not in renv.lock

# Fix: always snapshot after installing
install.packages("scales")
renv::snapshot()  # Update the lock file
```
:::

### Pipeline Orchestration with {targets}

For multi-step analyses where some steps take a long time (alignment, normalization), re-running everything from scratch on every change is wasteful. The `targets` package solves this with a **dependency graph**: it only re-runs the steps whose inputs have changed.

```{r}
#| eval: false

# _targets.R — place this in your project root

library(targets)

# Tell targets which packages every target needs
tar_option_set(packages = c("dplyr", "readr"))

# Source your custom functions
tar_source("R/functions/")

# Define the pipeline as a list of tar_target() calls
list(
  # Track the raw counts file; re-run downstream if it changes
  tar_target(counts_file, "data/raw/gene_counts.csv", format = "file"),

  # Read counts into R
  tar_target(counts, read_csv(counts_file)),

  # Filter low-expressed genes (custom function in R/functions/)
  tar_target(filtered, filter_expressed(counts, min_cpm = 1)),

  # Compute per-gene summary table
  tar_target(summary_tbl, summarize_per_gene(filtered))
)
```

Run the pipeline and inspect it:

```{r}
#| eval: false

# Run all outdated targets (skips up-to-date ones)
targets::tar_make()

# Visualise the dependency graph in the browser
targets::tar_visnetwork()

# Load a specific target result into your session
targets::tar_load(filtered)
```

::: {.callout-important title="Why targets matters for bioinformatics"}
Alignment and counting can take hours per sample. If you change only the filtering threshold in `filter_expressed()`, `targets` detects that `counts_file` and `counts` are unchanged and **only re-runs `filtered` and `summary_tbl`**. This is the standard approach for reproducible, efficient bioinformatics pipelines. Widely-used Bioconductor workflows (e.g., `rnaseqGene`) are being ported to `targets`-based structures for exactly this reason.
:::

:::: {.callout-caution title="Your Turn — targets Mini-Pipeline"}
Convert the following 3-step analysis script into a `_targets.R` pipeline. Then answer: which targets re-run if you change the `min_cpm` threshold from 1 to 2?

```{r}
#| eval: false

# Current linear script (to convert):
counts     <- readr::read_csv("data/raw/gene_counts.csv")
filtered   <- filter_expressed(counts, min_cpm = 1)
summary_tbl <- summarize_per_gene(filtered)
```

::::

::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| eval: false

# _targets.R
library(targets)
tar_option_set(packages = c("dplyr", "readr"))
tar_source("R/functions/")

list(
  tar_target(counts_file, "data/raw/gene_counts.csv", format = "file"),
  tar_target(counts,       read_csv(counts_file)),
  tar_target(filtered,     filter_expressed(counts, min_cpm = 1)),
  tar_target(summary_tbl,  summarize_per_gene(filtered))
)

# If you change min_cpm = 1 to min_cpm = 2:
# - counts_file: up to date (file unchanged)
# - counts:      up to date (file unchanged)
# - filtered:    OUTDATED — the function call changed
# - summary_tbl: OUTDATED — its input (filtered) changed
# targets re-runs only filtered and summary_tbl.
```
:::

### README Template

Every project needs a README:

```{r}
#| eval: false

readme_template <- '
# Project Name

## Overview
One or two sentences describing the project and the biological question.

## Requirements
- R 4.4+
- Run `renv::restore()` to install dependencies

## Directory Structure
- `data/raw/fastq/`       - Raw FASTQ files (do not modify)
- `data/processed/counts/` - Gene count matrices
- `R/functions/`           - Reusable R functions
- `R/scripts/`             - Numbered analysis scripts
- `output/figures/`        - Saved plots
- `output/tables/`         - Summary tables
- `docs/`                  - Quarto reports

## Usage
1. Open `project-name.Rproj` in Positron
2. Run `renv::restore()` to install packages
3. Run `targets::tar_make()` to execute the full pipeline
4. Render `docs/report.qmd` for the final report

## Data Sources
Describe where data comes from and any access requirements.
'
cat(readme_template)
```

### Your Turn: Write a README

:::: {.callout-caution title="Your Turn — Write a README"}
Write a README for an RNA-seq project comparing gene expression in liver vs. kidney tissue. Include all sections from the template above, customized for this project.

```{r}
#| eval: false

my_readme <- '
# Your project README here
'
cat(my_readme)
```
::::

### .gitignore for R Projects

```{r}
#| eval: false

gitignore_content <- '
# R specific
.Rhistory
.RData
.Rproj.user

# renv (keep lock file, ignore library)
renv/library/
renv/staging/
renv/local/

# Generated output (reproducible from code)
output/figures/*.png
output/tables/*.csv
_site/

# Large bioinformatics data files
data/raw/fastq/*.fastq.gz
data/raw/fastq/*.bam

# OS files
.DS_Store
Thumbs.db
'
cat(gitignore_content)
```

Key decisions: commit `renv.lock` (so others can restore packages). Do not commit `renv/library/` (too large, platform-specific). Do not commit large FASTQ or BAM files (share via institutional storage or SRA accession number).

::: {.callout-tip title="Challenge — Complete Project Setup with `fs`"}
Create a complete bioinformatics project setup function that:

1. Creates the directory structure using `fs::dir_create()`
2. Writes a README.md with `fs::path()`
3. Writes a .gitignore
4. Creates a starter `_targets.R`

```{r}
#| eval: false

library(fs)

setup_project <- function(project_name) {
  # Create directories — fs::dir_create() is recursive by default
  dirs <- fs::path(project_name, c(
    "data/raw/fastq",
    "data/processed/counts",
    "R/functions",
    "R/scripts",
    "output/figures",
    "output/tables",
    "docs"
  ))
  fs::dir_create(dirs)

  # Create and populate README
  writeLines(
    paste0("# ", project_name, "\n\nProject description here."),
    fs::path(project_name, "README.md")
  )

  # Create and populate .gitignore
  writeLines(
    c(".Rhistory", ".RData", ".Rproj.user/", "renv/library/",
      "data/raw/fastq/*.fastq.gz"),
    fs::path(project_name, ".gitignore")
  )

  # Create starter _targets.R
  writeLines(
    c('library(targets)',
      'tar_option_set(packages = c("dplyr", "readr"))',
      'tar_source("R/functions/")',
      'list(',
      '  tar_target(counts_file, "data/raw/gene_counts.csv", format = "file"),',
      '  tar_target(counts, readr::read_csv(counts_file))',
      ')'),
    fs::path(project_name, "_targets.R")
  )

  message("Project '", project_name, "' created.")
  invisible(fs::path_abs(project_name))
}

# Test it:
# setup_project("rnaseq-analysis")
```
:::

## Part 2: Debugging and Profiling

### Reading Error Messages

Error messages follow a pattern: what went wrong and where.

```{r}
#| error: true

# Error 1: object not found
mean(nonexistent_variable)
```

```{r}
#| error: true

# Error 2: wrong argument type
log("text")
```

```{r}
#| error: true

# Error 3: subscript out of bounds
x <- c(1, 2, 3)
x[5] <- 10  # This works (extends vector)
x[[5]]      # This fails if x has only 4 elements now
```

Read the error message first. It often tells you exactly what to fix.

### `traceback()` for Call Stacks

When errors occur inside nested function calls:

```{r}
#| error: true

inner <- function(x) {
  log(x)  # Fails if x is character
}

middle <- function(x) {
  inner(paste(x))  # Converts to character
}

outer <- function(x) {
  middle(x)
}

outer(5)
```

After an error, `traceback()` shows which function called which:

```{r}
#| eval: false

traceback()
# 3: inner(paste(x))
# 2: middle(x)
# 1: outer(5)
# Read bottom to top: outer called middle called inner
```

### `browser()` for Interactive Debugging

Insert `browser()` to pause execution and inspect variables:

```{r}
#| eval: false

filter_expressed <- function(counts, min_cpm = 1) {
  browser()  # Execution pauses here
  # Compute CPM: divide each count by column sum * 1e6
  cpm_mat <- sweep(counts, 2, colSums(counts), "/") * 1e6
  # Keep genes where mean CPM across samples exceeds threshold
  keep <- rowMeans(cpm_mat) > min_cpm
  counts[keep, ]
}

# Simulated count matrix (rows = genes, columns = samples)
set.seed(1)
sim_counts <- matrix(rpois(500, lambda = 50), nrow = 50,
                     dimnames = list(paste0("gene_", 1:50),
                                     paste0("sample_", 1:10)))

# When this runs, you drop into debug mode:
filter_expressed(sim_counts, min_cpm = 5)

# In debug mode, you can:
# Type variable names to see their values: cpm_mat, keep
# n  - execute next line
# c  - continue (run to end or next browser())
# Q  - quit debugging
# ls() - list objects in current scope
```

### Your Turn: Find the Bug in `apply_qc_filter()`

:::: {.callout-caution title="Your Turn — Find the Bug"}
This function is supposed to return samples with RIN score **at or above** the threshold, but it silently drops borderline samples. Find and fix the bug using `browser()`.

```{r}
samples_metadata <- data.frame(
  sample_id = paste0("S", 1:6),
  tissue    = c("liver", "kidney", "liver", "kidney", "liver", "kidney"),
  condition = c("control", "control", "treated", "treated", "control", "treated"),
  rin_score = c(8.5, 7.9, 8.0, 9.1, 6.5, 8.0)
)

# Buggy function — off-by-one in the comparison operator:
apply_qc_filter <- function(samples, rin_threshold = 8.0) {
  samples[samples$rin_score > rin_threshold, ]  # Bug: should be >=
}

# This drops samples with rin_score == 8.0 (S3 and S6)
apply_qc_filter(samples_metadata, rin_threshold = 8.0)
```

Use `browser()` to step through. What do you see in `samples$rin_score > 8.0` for S3?

Fix the function:

```{r}
#| code-fold: true

# Fixed version:
flag_high_quality <- function(samples, rin_threshold = 8.0) {
  threshold <- as.numeric(rin_threshold)
  if (is.na(threshold)) {
    stop("rin_threshold must be numeric, got: ", rin_threshold)
  }
  samples[samples$rin_score >= threshold, ]  # >= keeps borderline samples
}

flag_high_quality(samples_metadata, rin_threshold = 8.0)
# Now includes S3 and S6 (rin_score == 8.0)
```
::::

### Positron Debugger

Positron provides a visual debugger:

1. **Set breakpoint**: Click the line number gutter (red dot appears)
2. **Source file**: `Ctrl+Shift+Enter` to source with debug support
3. **Call function**: Execution pauses at breakpoint
4. **Debug toolbar**: Step Over (F10), Step Into (F11), Continue (F5)
5. **Variables pane**: Shows all objects in current scope

Conditional breakpoints: right-click a breakpoint, enter a condition like `i == 42`. The debugger only pauses when the condition is true.

:::: {.callout-caution title="Your Turn — Debugging in Positron"}
The function below processes samples per tissue group. Set a breakpoint on the `tissue_subset <- ...` line. Use the Variables panel to inspect `tissue_subset` when the loop reaches `tissue == "kidney"`.

```{r}
#| eval: false

process_by_tissue <- function(samples) {
  tissues <- unique(samples$tissue)
  results <- lapply(tissues, function(tissue) {
    tissue_subset <- samples[samples$tissue == tissue, ]  # Set breakpoint here
    data.frame(
      tissue       = tissue,
      n_samples    = nrow(tissue_subset),
      mean_rin     = mean(tissue_subset$rin_score),
      pct_hq       = mean(tissue_subset$rin_score >= 8.0) * 100
    )
  })
  do.call(rbind, results)
}

process_by_tissue(samples_metadata)
```

::::

::: {.callout-tip collapse="true" title="Solution"}
Steps:

1. Open the script containing `process_by_tissue` in Positron.
2. Click the gutter to the left of the `tissue_subset <- ...` line — a red dot appears.
3. Source the file (`Ctrl+Shift+Enter`) to register the function.
4. Call `process_by_tissue(samples_metadata)` in the console.
5. Execution pauses at the breakpoint on the first iteration (`tissue == "liver"`).
6. In the **Variables** panel (top-right in the default layout), expand `tissue_subset`. You should see 3 rows (S1, S5, and the control liver sample).
7. Press **Continue** (F5) to advance to the next iteration.
8. On the second pause, `tissue_subset` shows 3 kidney rows. Verify `mean_rin` would be `(7.9 + 9.1 + 8.0) / 3 = 8.33`.

The Variables panel updates live with each step — no `print()` statements needed.
:::

### Your Turn: Debugging Practice

:::: {.callout-caution title="Your Turn — Weighted Mean Bugs"}
This function is supposed to compute the weighted mean, but has multiple bugs. Use `browser()` or Positron breakpoints to find and fix each one.

```{r}
#| error: true

# Buggy function (3 bugs):
weighted_avg <- function(values, weights) {
  # Bug 1: no length check
  # Bug 2: no NA handling
  # Bug 3: division issue
  total <- sum(values * weights)
  n <- length(values)
  total / n  # Should divide by sum of weights, not count
}

# Test cases:
weighted_avg(c(80, 90, 70), c(0.3, 0.5, 0.2))  # Should be 81
weighted_avg(c(80, NA, 70), c(0.3, 0.5, 0.2))   # Should handle NA
weighted_avg(c(80, 90), c(0.3, 0.5, 0.2))        # Should error: different lengths
```

Fix all three bugs:

```{r}
#| code-fold: true

weighted_avg_fixed <- function(values, weights) {
  # Bug 1 fix: check lengths match
  if (length(values) != length(weights)) {
    stop("values and weights must have same length")
  }

  # Bug 2 fix: handle NAs
  complete <- !is.na(values) & !is.na(weights)
  values  <- values[complete]
  weights <- weights[complete]

  # Bug 3 fix: divide by sum of weights
  sum(values * weights) / sum(weights)
}

weighted_avg_fixed(c(80, 90, 70), c(0.3, 0.5, 0.2))
weighted_avg_fixed(c(80, NA, 70), c(0.3, 0.5, 0.2))
```
::::

### Performance Timing

```{r}
# Time a single expression:
system.time({
  result <- sapply(1:100000, \(x) sqrt(x))
})
```

```{r}
# Compare approaches:
system.time({
  # Approach 1: loop with growing vector
  out <- c()
  for (i in 1:50000) out <- c(out, sqrt(i))
})

system.time({
  # Approach 2: pre-allocated vector
  out <- numeric(50000)
  for (i in 1:50000) out[i] <- sqrt(i)
})

system.time({
  # Approach 3: vectorized
  out <- sqrt(1:50000)
})
```

Approach 3 (vectorized) is fastest. Approach 1 (growing vector) is slowest because R copies the entire vector on each append.

### Your Turn: Performance Comparison

:::: {.callout-caution title="Your Turn — Row Means Timing"}
Compare three ways to compute row means for a matrix. Time each approach.

```{r}
# Create test data
set.seed(42)
mat <- matrix(rnorm(100000), nrow = 1000, ncol = 100)

# Approach 1: for loop
system.time({
  means1 <- numeric(nrow(mat))
  for (i in 1:nrow(mat)) {
    means1[i] <- mean(mat[i, ])
  }
})

# Approach 2: apply
system.time({
  means2 <- apply(mat, 1, mean)
})

# Approach 3: rowMeans (built-in vectorized)
system.time({
  means3 <- rowMeans(mat)
})

# Verify all give same result:
all.equal(means1, means2)
all.equal(means2, means3)
```
::::

### Profiling with `profvis`

`system.time()` tells you *how long* something takes; `profvis` tells you *where* the time is spent inside a function — which calls dominate the flame graph.

::: {.callout-note title="Reading a profvis flame graph"}
In the flame graph, the x-axis is cumulative time. Wide bars = slow calls. The bottom layer is the outermost function; each layer above is a nested call. Look for the **widest bar near the top** — that is your bottleneck.
:::

Below we profile a realistic bioinformatics scenario: reading a gene counts file, filtering low-expressed genes, and computing per-gene mean expression. We compare a slow `for`-loop version against a vectorized `rowMeans()` version.

```{r}
#| eval: false

library(profvis)

# ------------------------------------------------------------------
# Simulate a gene counts matrix (rows = genes, columns = samples)
# ------------------------------------------------------------------
set.seed(42)
n_genes   <- 5000
n_samples <- 20
sim_counts <- matrix(
  rpois(n_genes * n_samples, lambda = 100),
  nrow     = n_genes,
  dimnames = list(
    paste0("ENSG", sprintf("%05d", seq_len(n_genes))),
    paste0("sample_", seq_len(n_samples))
  )
)

# ------------------------------------------------------------------
# Slow version: for-loop over every gene
# ------------------------------------------------------------------
compute_gene_means_loop <- function(counts) {
  gene_means <- numeric(nrow(counts))
  for (i in seq_len(nrow(counts))) {
    gene_means[i] <- mean(counts[i, ])  # re-indexes each row individually
  }
  names(gene_means) <- rownames(counts)
  gene_means
}

# ------------------------------------------------------------------
# Fast version: vectorized rowMeans()
# ------------------------------------------------------------------
compute_gene_means_vec <- function(counts) {
  rowMeans(counts)  # one BLAS call across all rows
}

# ------------------------------------------------------------------
# Filter function used by both versions
# ------------------------------------------------------------------
filter_expressed <- function(counts, min_cpm = 1) {
  lib_sizes <- colSums(counts)                          # total counts per sample
  cpm_mat   <- sweep(counts, 2, lib_sizes, "/") * 1e6  # counts per million
  keep      <- rowMeans(cpm_mat) > min_cpm
  counts[keep, , drop = FALSE]
}

# ------------------------------------------------------------------
# Profile the slow pipeline
# ------------------------------------------------------------------
profvis({
  filtered_slow  <- filter_expressed(sim_counts, min_cpm = 1)
  gene_means_slow <- compute_gene_means_loop(filtered_slow)
})

# ------------------------------------------------------------------
# Profile the fast pipeline — run AFTER examining the flame graph above
# ------------------------------------------------------------------
profvis({
  filtered_fast  <- filter_expressed(sim_counts, min_cpm = 1)
  gene_means_fast <- compute_gene_means_vec(filtered_fast)
})

# ------------------------------------------------------------------
# Verify identical results
# ------------------------------------------------------------------
all.equal(gene_means_slow, gene_means_fast)  # TRUE

# ------------------------------------------------------------------
# Timing comparison
# ------------------------------------------------------------------
system.time(compute_gene_means_loop(sim_counts))  # seconds
system.time(compute_gene_means_vec(sim_counts))   # milliseconds
```

**What to look for in the flame graph:**

- In the **slow** profile, `compute_gene_means_loop` occupies most of the x-axis width. Inside it, repeated calls to `mean()` and index operations `[i, ]` each appear as thin stacked bars — but there are 5,000 of them. The `filter_expressed` call is a thin sliver.
- In the **fast** profile, `compute_gene_means_vec` barely registers. The call is one wide bar with a single `rowMeans` call; the bulk of time shifts into `filter_expressed`'s `sweep()` call, which is already vectorized.
- **Interpretation**: the loop is the bottleneck. Replacing it with `rowMeans()` moves the bottleneck to I/O and CPM calculation — both of which are already fast.

### Your Turn: Profile and Optimize

:::: {.callout-caution title="Your Turn — Profile and Optimize"}
The function below computes per-gene mean expression using a `for` loop.

1. Run `profvis({ slow_means <- gene_mean_loop(sim_counts) })` and note which call takes the most time.
2. Rewrite using `rowMeans()`.
3. Compare timings with `system.time()`.

```{r}
#| eval: false

set.seed(99)
sim_counts <- matrix(rpois(2000 * 30, lambda = 80), nrow = 2000,
                     dimnames = list(paste0("gene_", 1:2000),
                                     paste0("s", 1:30)))

gene_mean_loop <- function(counts) {
  result <- numeric(nrow(counts))
  for (i in seq_len(nrow(counts))) {
    result[i] <- mean(counts[i, ])
  }
  result
}

# YOUR CODE: profile, then write gene_mean_fast() using rowMeans()
```

::::

::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| eval: false

library(profvis)

# Profile the slow version
profvis({
  slow_out <- gene_mean_loop(sim_counts)
})
# Flame graph: gene_mean_loop is the widest bar.
# Inside it, the loop body (mean + indexing) repeats 2000 times.

# Fast version
gene_mean_fast <- function(counts) {
  rowMeans(counts)
}

# Timing comparison
system.time(gene_mean_loop(sim_counts))  # ~0.05–0.3 s
system.time(gene_mean_fast(sim_counts)) # <0.002 s

# Verify results match
all.equal(gene_mean_loop(sim_counts), gene_mean_fast(sim_counts))
```
:::

::: {.callout-warning title="Common Mistake"}
Growing vectors inside loops:

```{r}
#| eval: false

# Slow: copies entire vector on each iteration
results <- c()
for (i in 1:10000) {
  results <- c(results, compute_something(i))
}

# Fast: pre-allocate, then fill
results <- numeric(10000)
for (i in 1:10000) {
  results[i] <- compute_something(i)
}

# Fastest: vectorize if possible
results <- compute_something(1:10000)
```

R must allocate new memory and copy all existing elements each time you append with `c()`. Pre-allocation avoids this.
:::

## Part 3: Code Style

### The Tidyverse Style Guide

The standard R style reference: <https://style.tidyverse.org/>

```{r}
#| eval: false

# Naming: snake_case for everything
day_one          # variable
filter_expressed # function (verb phrase)
is_valid         # logical function (predicate)

# Spacing: around operators and after commas
x <- 5
y <- x + 1
z <- c(1, 2, 3)
mtcars[mtcars$cyl == 4, ]
mean(x, na.rm = TRUE)

# Line length: 80 characters max
# Indentation: 2 spaces
# Assignment: <- (not =)
```

### Native Pipe vs `%>%`

::: {.callout-note title="Native Pipe `|>` vs magrittr `%>%`"}
Since R 4.1, the **native pipe** `|>` is built into base R — no package required.

**Key differences from `magrittr`'s `%>%`:**

| Feature | `\|>` (native) | `%>%` (magrittr) |
|---------|--------------|----------------|
| Requires package | No | Yes (`magrittr` / `dplyr`) |
| Speed | Faster — zero overhead | Small overhead per call |
| Placeholder | `_` (R 4.2+, named arg only) | `.` (anywhere) |
| Lambda shorthand | `\(x) x + 1` | `~ .x + 1` |

**In 2026, prefer `|>` for all new code.** You will still see `%>%` in older codebases, tutorials, and Bioconductor vignettes — both work and can be mixed within a project if needed.

Example with placeholder:
```r
# Native pipe: placeholder must be used as a named argument
mtcars |> lm(mpg ~ cyl, data = _)

# magrittr: . can go anywhere
mtcars %>% lm(mpg ~ cyl, data = .)
```
:::

### Your Turn: Fix the Style

:::: {.callout-caution title="Your Turn — Style Cleanup"}
The script below has mixed `=`/`<-`, inconsistent spacing, `dplyr::filter` mixed with `subset()`, and lines over 100 characters. Run `styler::style_text()` on it, then fix the remaining issues that `styler` cannot auto-correct.

```{r}
#| eval: false

# Ugly script — fix it:
Process_Counts<-function(counts,threshold=1,RemoveZero=T){
x=colSums(counts);cpm=sweep(counts,2,x,"/")*1e6
keep=subset(as.data.frame(rowMeans(cpm)),rowMeans(cpm)>threshold)
if(RemoveZero==TRUE){counts=counts[rowSums(counts)>0,]}
return(counts[rownames(keep),])}

# Run styler on a string version:
styler::style_text('Process_Counts<-function(counts,threshold=1,RemoveZero=T){x=colSums(counts)}')
```

::::

::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| eval: false

library(styler)
library(lintr)

# First: auto-format with styler (fixes spacing, <- vs =, indentation)
styler::style_text('
Process_Counts<-function(counts,threshold=1,RemoveZero=T){
x=colSums(counts);cpm=sweep(counts,2,x,"/")*1e6
keep=subset(as.data.frame(rowMeans(cpm)),rowMeans(cpm)>threshold)
if(RemoveZero==TRUE){counts=counts[rowSums(counts)>0,]}
return(counts[rownames(keep),])}
')

# Then: apply lintr to find remaining issues
# e.g. T instead of TRUE, PascalCase name, subset() vs dplyr::filter()

# Fully corrected version:
filter_expressed <- function(counts, min_cpm = 1, remove_zero = TRUE) {
  if (remove_zero) {
    counts <- counts[rowSums(counts) > 0, ]
  }

  lib_sizes <- colSums(counts)
  cpm_mat   <- sweep(counts, 2, lib_sizes, "/") * 1e6
  keep      <- rowMeans(cpm_mat) > min_cpm

  counts[keep, , drop = FALSE]
}
```
:::

### Your Turn: Lint and Style

:::: {.callout-caution title="Your Turn — Lint and Style"}
Identify all style violations in this code block. Then write the corrected version.

```{r}
#| eval: false

# Find all violations:
Process_Data <- function(x,threshold=0.5,RemoveNA=T){
  if(RemoveNA==TRUE){x=x[!is.na(x)]}
  x_scaled = (x-min(x))/(max(x)-min(x))
  result = x_scaled[x_scaled>threshold]
  return(result)
}
```

Violations to find:

1. Function name uses PascalCase (should be snake_case)
2. Parameter `RemoveNA` uses PascalCase
3. `T` instead of `TRUE`
4. No spaces around `=`, `-`, `/`, `>`
5. Using `=` for assignment instead of `<-`
6. No spaces after commas
7. No space after `if`
8. Everything crammed onto minimal lines

```{r}
#| code-fold: true

# Corrected:
process_data <- function(x, threshold = 0.5, remove_na = TRUE) {
  if (remove_na) {
    x <- x[!is.na(x)]
  }

  x_scaled <- (x - min(x)) / (max(x) - min(x))
  result <- x_scaled[x_scaled > threshold]

  result
}

process_data(c(1, 5, 3, NA, 8, 2), threshold = 0.4)
```
::::

### `lintr`: Finding Style Problems

```{r}
#| eval: false

library(lintr)

# Check a single file:
lint("R/analysis.R")

# Check all R files in a directory:
lint_dir("R/")

# Check a string of code:
lint_text("x<-1+2")
# Returns:
# Line 1, Column 2: Use <-, not =, for assignment.
# Line 1, Column 2: Put spaces around all infix operators.
```

`lintr` integrates with Positron to show warnings inline as you type.

### `styler`: Auto-formatting

```{r}
#| eval: false

library(styler)

# Format a file:
style_file("R/messy_code.R")

# Format all files in a directory:
style_dir("R/")

# Format a string:
style_text("x<-1+2\nif(x>2){print('hi')}")
# Returns properly formatted code
```

`styler` fixes spacing, indentation, and line breaks automatically. Use it before committing code.

### Commenting Best Practices

```{r}
#| eval: false

# BAD: narrates the code (the code already says this)
# Calculate the mean of rin_score
mean_rin <- mean(samples_metadata$rin_score)

# GOOD: explains a decision
# Using median instead of mean because RIN score distribution
# is left-skewed for low-quality samples (see QC report, Fig 2)
center <- median(samples_metadata$rin_score)

# BAD: outdated comment that contradicts code
# Remove rows with missing values
clean <- samples_metadata[samples_metadata$rin_score > 0, ]  # Filters by score

# GOOD: section headers for navigation
# Load and validate sample metadata ----
metadata <- readr::read_tsv(here("data", "raw", "samples_metadata.tsv"))
stopifnot(nrow(metadata) > 0)
```

### Your Turn: Improve Comments

:::: {.callout-caution title="Your Turn — Improve Comments"}
Rewrite the comments in this code. Remove unnecessary comments, add comments that explain decisions, and add section headers.

```{r}
#| eval: false

# Original (bad comments):
library(dplyr)  # load dplyr

# read the data
data <- readr::read_tsv("samples_metadata.tsv")

# filter to liver
liver <- data |> filter(tissue == "liver")

# calculate mean
avg_rin <- mean(liver$rin_score, na.rm = TRUE)

# multiply by 0.9
projected <- avg_rin * 0.9

# print result
print(projected)
```

Write the improved version:

```{r}
#| code-fold: true
#| eval: false

# Setup ----
library(dplyr)
library(here)
library(readr)

# Import sample metadata ----
metadata <- read_tsv(here("data", "raw", "samples_metadata.tsv"))

# Liver tissue QIN projection ----
# Focus on liver samples because they show least batch effect (QC report, p.4)
liver <- metadata |> filter(tissue == "liver")
avg_rin <- mean(liver$rin_score, na.rm = TRUE)

# Apply 10% degradation correction based on storage time analysis
# (see methods in docs/report.qmd, section 2.3)
corrected_rin <- avg_rin * 0.9
corrected_rin
```
::::

### roxygen2 Documentation

Document functions with standardized comments:

```{r}
#' Filter a gene count matrix to remove low-expressed genes
#'
#' Computes counts per million (CPM) for each sample, then retains only
#' genes whose mean CPM across all samples exceeds \code{min_cpm}.
#'
#' @param counts A numeric matrix of raw read counts.
#'   Rows are genes (with rownames as gene IDs), columns are samples.
#' @param min_cpm Numeric. Minimum mean CPM to keep a gene. Default \code{1}.
#' @param min_samples Integer. Minimum number of samples that must exceed
#'   \code{min_cpm} (alternative filtering strategy). Default \code{3}.
#' @return A matrix with the same structure as \code{counts} but with
#'   low-expressed gene rows removed.
#' @examples
#' set.seed(1)
#' counts <- matrix(rpois(500, 50), nrow = 50,
#'                  dimnames = list(paste0("gene_", 1:50),
#'                                  paste0("s", 1:10)))
#' filter_expressed(counts, min_cpm = 1)
#' @export
filter_expressed <- function(counts, min_cpm = 1, min_samples = 3) {
  lib_sizes <- colSums(counts)
  cpm_mat   <- sweep(counts, 2, lib_sizes, "/") * 1e6
  keep      <- rowMeans(cpm_mat) > min_cpm
  counts[keep, , drop = FALSE]
}

# Demonstrate the function:
set.seed(1)
sim <- matrix(rpois(500, 50), nrow = 50,
              dimnames = list(paste0("gene_", 1:50), paste0("s", 1:10)))
dim(filter_expressed(sim, min_cpm = 1))
```

:::: {.callout-caution title="Your Turn — Document with roxygen2"}
Write full roxygen2 documentation for `filter_expressed(counts, min_cpm = 1, min_samples = 3)`. Include: title, one-paragraph description, `@param` for each argument, `@return`, `@examples`, and `@export`. After writing it, check what `?filter_expressed` would show.

```{r}
#| eval: false

# Write your documentation here:
#' <YOUR TITLE>
#'
#' <YOUR DESCRIPTION>
#'
#' @param counts ...
#' @param min_cpm ...
#' @param min_samples ...
#' @return ...
#' @examples
#' ...
#' @export
filter_expressed <- function(counts, min_cpm = 1, min_samples = 3) {
  lib_sizes <- colSums(counts)
  cpm_mat   <- sweep(counts, 2, lib_sizes, "/") * 1e6
  keep      <- rowMeans(cpm_mat) > min_cpm
  counts[keep, , drop = FALSE]
}
```

::::

::: {.callout-tip collapse="true" title="Solution"}
```{r}
#' Filter a gene count matrix to remove low-expressed genes
#'
#' Computes counts per million (CPM) normalised to each sample library size,
#' then retains only genes whose mean CPM across all samples exceeds
#' \code{min_cpm}. This is a standard pre-filtering step before differential
#' expression analysis (Robinson & Oshlack 2010).
#'
#' @param counts A numeric matrix of raw integer read counts.
#'   Rows are genes (rownames should be Ensembl IDs or gene symbols),
#'   columns are samples (colnames should be sample IDs).
#' @param min_cpm Numeric scalar. Genes with mean CPM strictly greater than
#'   this value are retained. Default \code{1}.
#' @param min_samples Integer scalar. Reserved for future use: minimum number
#'   of samples that must exceed \code{min_cpm} individually (alternative
#'   filtering strategy). Default \code{3}.
#' @return A numeric matrix with the same column structure as \code{counts}
#'   but with low-expressed gene rows removed. Row and column names are
#'   preserved.
#' @examples
#' set.seed(42)
#' counts <- matrix(rpois(500, 50), nrow = 50,
#'                  dimnames = list(paste0("ENSG", 1:50),
#'                                  paste0("sample_", 1:10)))
#' filtered <- filter_expressed(counts, min_cpm = 1)
#' nrow(filtered)  # fewer than 50
#' @export
filter_expressed <- function(counts, min_cpm = 1, min_samples = 3) {
  lib_sizes <- colSums(counts)
  cpm_mat   <- sweep(counts, 2, lib_sizes, "/") * 1e6
  keep      <- rowMeans(cpm_mat) > min_cpm
  counts[keep, , drop = FALSE]
}
```
:::

::: {.callout-warning title="Common Mistake"}
Using `T` and `F` instead of `TRUE` and `FALSE`:

```{r}
# T and F are just variables that happen to be TRUE and FALSE:
T
F

# But they can be overwritten:
T <- 42
T  # Now T is 42, not TRUE

# TRUE and FALSE are reserved words and cannot be overwritten:
# TRUE <- 42  # Error: invalid (do_set) left-hand side to assignment

# Always use TRUE and FALSE, never T and F
rm(T)  # Clean up
```
:::

### Your Turn: `fs` Path Manipulation

:::: {.callout-caution title="Your Turn — `fs` Path Manipulation"}
You have a vector of 100 FASTQ filenames following the naming convention `sample_001_R1.fastq.gz`. Use `fs::path_file()`, `fs::path_ext_remove()`, and `stringr::str_extract()` to build a tibble with columns `sample_id` and `read_direction`.

```{r}
#| eval: false

library(fs)
library(stringr)
library(tibble)

# Simulate 100 FASTQ filenames
fq_files <- c(
  outer(sprintf("sample_%03d", 1:50), c("_R1.fastq.gz", "_R2.fastq.gz"),
        paste0)
) |> sort()

head(fq_files, 6)

# YOUR CODE:
# 1. Extract just the filename (no directory) with fs::path_file()
# 2. Strip the .fastq.gz extension(s) with fs::path_ext_remove() twice
# 3. Extract sample_id and read_direction with str_extract()
# 4. Combine into a tibble
```

::::

::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| eval: false

library(fs)
library(stringr)
library(tibble)

fq_files <- c(
  outer(sprintf("sample_%03d", 1:50), c("_R1.fastq.gz", "_R2.fastq.gz"),
        paste0)
) |> sort()

fq_tibble <- tibble(
  filepath       = fq_files,
  # Strip directory component (here files have none, but good practice)
  filename       = fs::path_file(fq_files),
  # Remove .gz, then .fastq — two calls because path_ext_remove() removes one extension at a time
  stem           = fs::path_ext_remove(fs::path_ext_remove(filename)),
  sample_id      = str_extract(stem, "sample_\\d+"),
  read_direction = str_extract(stem, "R[12]$")
) |>
  dplyr::select(sample_id, read_direction, filepath)

head(fq_tibble)
```
:::

::: {.callout-tip title="Bonus Challenge — Full Project Initializer"}
Build a `setup_bioinformatics_project(name)` function that:

1. Creates the full directory skeleton with `fs::dir_create()`
2. Initializes `renv` (calls `renv::init()`)
3. Creates a `_targets.R` template
4. Writes a starter `README.md`
5. Returns the project path invisibly

```{r}
#| eval: false

library(fs)

setup_bioinformatics_project <- function(name) {
  # 1. Create directories
  fs::dir_create(fs::path(name, c(
    "data/raw/fastq", "data/processed/counts",
    "R/functions", "R/scripts",
    "output/figures", "output/tables",
    "docs"
  )))

  # 2. Initialize renv
  withr::with_dir(name, renv::init(bare = TRUE))

  # 3. Write _targets.R template
  writeLines(c(
    'library(targets)',
    'tar_option_set(packages = c("dplyr", "readr"))',
    'tar_source("R/functions/")',
    'list(',
    '  tar_target(counts_file, "data/raw/gene_counts.csv", format = "file"),',
    '  tar_target(counts,      readr::read_csv(counts_file))',
    ')'
  ), fs::path(name, "_targets.R"))

  # 4. Write README.md
  writeLines(c(
    paste0("# ", name),
    "",
    "## Overview",
    "RNA-seq analysis project.",
    "",
    "## Setup",
    "```r",
    "renv::restore()",
    "targets::tar_make()",
    "```"
  ), fs::path(name, "README.md"))

  # 5. Return invisibly
  invisible(fs::path_abs(name))
}

# setup_bioinformatics_project("pilot-rnaseq")
```
:::

::: {.callout-tip title="Challenge — Complete Styled Analysis Script"}
Write a complete, well-styled analysis script that:

1. Uses `here()` for all file paths
2. Has section headers
3. Defines at least two functions with roxygen2 documentation
4. Follows tidyverse style guide
5. Has meaningful comments (why, not what)
6. Would pass `lintr::lint()` with no warnings

The script should load `samples_metadata.tsv`, flag high-quality samples, and produce a summary by tissue.

```{r}
#| code-fold: true
#| eval: false

# RNA-seq Sample QC Summary
# Purpose: Identify high-quality samples and summarize by tissue type

# Setup ----
library(here)
library(dplyr)
library(readr)

# Functions ----

#' Flag samples meeting minimum RIN score threshold
#'
#' @param samples A data frame with a \code{rin_score} column.
#' @param rin_threshold Numeric. Minimum acceptable RIN score. Default \code{8.0}.
#' @return Filtered data frame containing only high-quality samples.
flag_high_quality <- function(samples, rin_threshold = 8.0) {
  samples[samples$rin_score >= rin_threshold, ]
}

#' Summarize sample quality metrics per tissue group
#'
#' @param samples A data frame with columns \code{tissue} and \code{rin_score}.
#' @return A data frame with per-tissue n, mean RIN, and % high-quality.
summarize_by_tissue <- function(samples) {
  samples |>
    group_by(tissue) |>
    summarise(
      n_samples = n(),
      mean_rin  = round(mean(rin_score, na.rm = TRUE), 2),
      pct_hq    = round(mean(rin_score >= 8.0) * 100, 1),
      .groups   = "drop"
    )
}

# Import ----
# read_tsv is more explicit than read.csv for tab-delimited files
metadata <- read_tsv(here("data", "raw", "samples_metadata.tsv"))

# QC filtering ----
# RIN >= 8.0 is the lab's minimum threshold for library prep (SOP-RNAseq-003)
hq_samples <- flag_high_quality(metadata, rin_threshold = 8.0)

# Summary ----
tissue_summary <- summarize_by_tissue(hq_samples)

cat("High-quality samples by tissue:\n")
print(tissue_summary)
```
:::

## Session Synthesis

### Connect the Dots

::: {.callout-important title="Synthesis Exercise"}
Set up a complete professional bioinformatics project from scratch.

Scenario: You are analyzing RNA-seq data from 20 samples (10 liver, 10 kidney) to identify differentially expressed genes.

Deliver:

1. Directory structure (bioinformatics-relevant subdirectories)
2. README.md content
3. .gitignore content
4. A well-styled `R/scripts/01-filter.R` script that:
   - Uses `here()` for paths
   - Loads `samples_metadata.tsv` and `gene_counts.csv`
   - Calls `filter_expressed()` and `flag_high_quality()`
   - Has roxygen2-documented helper functions
   - Follows tidyverse style guide with section headers
5. `renv` initialization and a `_targets.R` outline

```{r}
#| eval: false

# 1. Directory structure:
cat("
rnaseq-livkid/
  rnaseq-livkid.Rproj
  README.md
  .gitignore
  renv.lock
  _targets.R
  data/
    raw/
      fastq/          # Raw FASTQ (not committed to git)
    processed/
      counts/         # Gene count matrix from STAR/featureCounts
  R/
    functions/        # filter_expressed.R, plot_pca.R, ...
    scripts/          # 01-filter.R, 02-normalize.R, 03-de.R
  output/
    figures/          # PCA plots, volcano plots
    tables/           # DEG tables
  docs/
    report.qmd        # Final rendered report
")

# 2. README, .gitignore, scripts: write them here...

# 5. renv + targets setup:
# renv::init()
# renv::install(c("dplyr", "readr", "edgeR", "bioc::DESeq2"))
# renv::snapshot()
# targets::tar_make()
```
:::

### Key Takeaways

1. R Projects and `here()` make file paths portable across machines
2. The `fs` package provides type-safe, cross-platform file system operations — prefer it over `dir.create()` / `file.path()`
3. `renv::snapshot()` and `renv::restore()` lock package versions for reproducibility, including Bioconductor packages
4. `targets` orchestrates multi-step pipelines and only re-runs outdated steps — essential for long-running bioinformatics workflows
5. `traceback()`, `browser()`, and Positron breakpoints are your primary debugging tools
6. `profvis` reveals *where* time is spent; vectorized functions (`rowMeans()`, `sweep()`) are almost always faster than explicit loops
7. `lintr` finds style problems; `styler` auto-fixes them
8. Comment WHY (decisions and reasoning), not WHAT (the code already says what)
9. Use `|>` for new code; you will encounter `%>%` in older codebases — both work

### Prepare for Next Session

Next session covers version control and the capstone project:

- Git basics inside Positron
- Final project work session
- Course presentations and wrap-up

Before next session:

- Install Git: <https://git-scm.com/>
- Create a GitHub account if you do not have one
- Decide on your final project topic and dataset
- Set up a project directory following today's structure using `fs::dir_create()`
