---
title: "Session 4: Files, Errors, and Review"
subtitle: "R Programming Workbook"
author: "R Programming Course"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-tools: true
    code-overflow: wrap
    theme: cosmo
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-note title="Learning Objectives"}
By the end of this session, you will be able to:

1. Read CSV, TSV, and text files into R using `read.csv()` and `read.table()`
2. Write data frames to files with `write.csv()` and `saveRDS()`
3. Use `readLines()` and connections for non-tabular data
4. Handle errors with `tryCatch()` and `try()`
5. Write defensive functions using `stop()`, `warning()`, and `message()`
6. Distinguish between errors, warnings, and messages
7. Apply all Base R skills from Sessions 1-4 in integrated exercises
8. Choose the right data structure and tool for a given problem
:::

## Part 1: Reading and Writing Data

### Reading CSV Files

The most common data format you will encounter is CSV. R reads these with `read.csv()`.

```{r}
# Create a temporary CSV to work with
students <- data.frame(
  name = c("Alice", "Bob", "Carlos", "Diana", "Elena", "Frank"),
  age = c(21, 23, 22, 20, 24, 21),
  major = c("CS", "Math", "CS", "Stats", "CS", "Math"),
  gpa = c(3.8, 3.2, 3.5, 3.9, 3.1, 3.6),
  year = c(3, 4, 3, 2, 4, 3)
)

# Write it out so we can practice reading it back
write.csv(students, "students.csv", row.names = FALSE)

# Read it back
students_loaded <- read.csv("students.csv")
str(students_loaded)
```

`read.csv()` returns a data frame. Strings become character vectors by default in modern R (R 4.0+). In older R, they became factors.

### Key Arguments for read.csv()

```{r}
# stringsAsFactors: convert strings to factors?
students_fac <- read.csv("students.csv", stringsAsFactors = TRUE)
str(students_fac$major)  # Factor

# nrows: read only first N rows (useful for previewing large files)
preview <- read.csv("students.csv", nrows = 2)
preview

# skip: skip header rows
skipped <- read.csv("students.csv", skip = 1, header = FALSE)
head(skipped)
```

Notice what happened when we used `skip = 1` with `header = FALSE`. The column names became V1, V2, etc., and the first data row was kept.

### read.table() for Other Delimiters

```{r}
# Create a tab-separated file
write.table(students, "students.tsv", sep = "\t", row.names = FALSE)

# Read tab-separated
students_tab <- read.table("students.tsv", sep = "\t", header = TRUE)
head(students_tab)

# read.csv is just read.table with sep = "," and header = TRUE
identical(
  read.csv("students.csv"),
  read.table("students.csv", sep = ",", header = TRUE)
)
```

### Your Turn: Read Data with Options

```{r}
#| code-fold: true
#| code-summary: "Exercise 1: Read the first 3 students, converting major to a factor"

# Your code here:
# Use read.csv() with nrows and stringsAsFactors arguments


# Solution
first_three <- read.csv("students.csv", nrows = 3, stringsAsFactors = TRUE)
first_three
class(first_three$major)
```

### Reading Non-Tabular Data

Not all data comes in neat rows and columns. `readLines()` reads each line as a string.

```{r}
# Create a text file
writeLines(c(
  "# Student Report",
  "Generated: 2024-03-15",
  "",
  "Total students: 6",
  "Average GPA: 3.52",
  "CS majors: 3"
), "report.txt")

# Read all lines
lines <- readLines("report.txt")
lines
length(lines)

# Extract just the data lines (skip comments and blanks)
data_lines <- lines[!grepl("^#", lines) & nchar(lines) > 0]
data_lines
```

### RDS Format: R's Native Binary

CSV loses type information. RDS preserves everything.

```{r}
# Add a factor column
students$major_fac <- factor(students$major)

# Save as RDS
saveRDS(students, "students.rds")

# Load it back - types are preserved
students_rds <- readRDS("students.rds")
class(students_rds$major_fac)  # Still a factor
levels(students_rds$major_fac)
```

Compare: CSV stores text. RDS stores R objects exactly as they are in memory. Use RDS for intermediate results, CSV for sharing with other tools.

### Your Turn: Round-Trip Data

```{r}
#| code-fold: true
#| code-summary: "Exercise 2: Create a matrix, save as RDS, load back, verify it is still a matrix"

# 1. Create a 3x3 matrix with values 1-9
# 2. Save it as "my_matrix.rds"
# 3. Load it back into a new variable
# 4. Check its class


# Solution
m <- matrix(1:9, nrow = 3)
saveRDS(m, "my_matrix.rds")
m_loaded <- readRDS("my_matrix.rds")
class(m_loaded)
identical(m, m_loaded)
```

### Your Turn: Parse a Report File

```{r}
#| code-fold: true
#| code-summary: "Exercise 3: Extract numeric values from report.txt"

# Read report.txt and extract all numbers from lines containing ':'
# Hint: use readLines(), grep(), and gsub()


# Solution
lines <- readLines("report.txt")
data_lines <- grep(":", lines, value = TRUE)
# Remove everything except digits and dots
numbers <- as.numeric(gsub("[^0-9.]", "", data_lines))
numbers  # 6, 3.52, 3 (date numbers will appear too - that is OK)
```

::: {.callout-warning title="Common Mistake"}
Forgetting `header = TRUE` in `read.table()`:

```{r}
#| error: true
bad <- read.table("students.csv", sep = ",")
head(bad, 2)  # First row becomes data, columns are V1, V2...
```

`read.csv()` sets `header = TRUE` by default. `read.table()` sets it to `FALSE`. Always check.
:::

### Cleanup

```{r}
# Clean up temporary files
file.remove("students.csv", "students.tsv", "report.txt", 
            "students.rds", "my_matrix.rds")
```

::: {.callout-tip title="Challenge"}
Write a function `read_auto(filepath)` that detects the file extension and calls the right reader: `.csv` uses `read.csv()`, `.tsv` uses `read.table(..., sep="\t", header=TRUE)`, `.rds` uses `readRDS()`. For anything else, return `readLines()`.

```{r}
#| code-fold: true
#| code-summary: "Solution"

read_auto <- function(filepath) {
  ext <- tools::file_ext(filepath)
  switch(ext,
    csv = read.csv(filepath),
    tsv = read.table(filepath, sep = "\t", header = TRUE),
    rds = readRDS(filepath),
    readLines(filepath)
  )
}
```
:::

## Part 2: Error Handling

### The Three Signals: Errors, Warnings, Messages

R has three condition types. Each serves a different purpose.

```{r}
#| error: true

# Error: execution stops
log(-1)    # just a warning
log("a")   # this is an error
```

```{r}
# Warning: something is off, but R continued
as.numeric(c("1", "2", "abc"))
```

```{r}
# Message: informational, not a problem
message("Processing 100 records...")
```

Think of it this way: errors mean "I cannot do this," warnings mean "I did it but something looks wrong," and messages mean "here is some information."

### Producing Signals in Your Functions

```{r}
#| error: true

divide <- function(x, y) {
  if (!is.numeric(x) || !is.numeric(y)) {
    stop("Both arguments must be numeric")
  }
  if (any(y == 0)) {
    warning("Division by zero encountered; returning Inf")
  }
  message("Dividing ", length(x), " values")
  x / y
}

divide(10, 2)
divide(10, 0)
divide("a", 2)
```

`stop()` raises an error and halts execution. `warning()` flags an issue but continues. `message()` prints info to the console.

### Your Turn: Defensive Function

```{r}
#| code-fold: true
#| code-summary: "Exercise 4: Write safe_sqrt() that handles bad input"

# Write safe_sqrt(x) that:
# - Stops with an error if x is not numeric
# - Warns if any values are negative
# - Returns sqrt(abs(x)) for negative values
# - Sends a message with how many values it processed


# Solution
safe_sqrt <- function(x) {
  if (!is.numeric(x)) stop("x must be numeric")
  if (any(x < 0, na.rm = TRUE)) {
    warning("Negative values detected; using absolute value")
    x <- abs(x)
  }
  message("Processing ", length(x), " values")
  sqrt(x)
}

safe_sqrt(c(4, 9, -16, 25))
```

### tryCatch: Handling Errors Gracefully

`tryCatch()` lets you intercept conditions and decide what to do.

```{r}
# Without tryCatch: error kills execution
# log("abc")  # would stop everything

# With tryCatch: you control what happens
result <- tryCatch(
  log("abc"),
  error = \(e) {
    cat("Caught error:", conditionMessage(e), "\n")
    NA
  }
)
result
```

The structure is: `tryCatch(expr, error = handler, warning = handler, message = handler)`. Each handler receives the condition object `e`.

```{r}
# Handle warnings too
result <- tryCatch(
  as.numeric("abc"),
  warning = \(w) {
    cat("Warning caught:", conditionMessage(w), "\n")
    NA
  }
)
result
```

### try: The Simpler Alternative

`try()` is less flexible but simpler. It wraps an expression and returns either the result or an object of class `"try-error"`.

```{r}
result <- try(log("abc"), silent = TRUE)
class(result)

if (inherits(result, "try-error")) {
  cat("Something went wrong\n")
} else {
  cat("Result:", result, "\n")
}
```

### Your Turn: Safe File Reader

```{r}
#| code-fold: true
#| code-summary: "Exercise 5: Write a function that tries to read a CSV and returns NULL on failure"

# Write safe_read(filepath) that:
# - Tries to read the file with read.csv()
# - Returns the data frame if it works
# - Returns NULL and prints a message if the file does not exist


# Solution
safe_read <- function(filepath) {
  tryCatch(
    read.csv(filepath),
    error = \(e) {
      message("Could not read '", filepath, "': ", conditionMessage(e))
      NULL
    }
  )
}

safe_read("nonexistent_file.csv")
safe_read_result <- safe_read("nonexistent_file.csv")
is.null(safe_read_result)
```

### Processing Multiple Items with Error Handling

A common pattern: loop over items, catch errors per item, keep going.

```{r}
inputs <- list(4, "bad", 9, NULL, 16)

results <- lapply(inputs, \(x) {
  tryCatch(
    sqrt(as.numeric(x)),
    error = \(e) NA,
    warning = \(w) NA
  )
})

results
```

### Your Turn: Batch Processing

```{r}
#| code-fold: true
#| code-summary: "Exercise 6: Process a list of mixed values, returning converted numbers or NA"

# Given this list:
values <- list("42", "3.14", "abc", "100", "", "7.5")

# Convert each to numeric using tryCatch.
# Return NA for any that fail.
# Hint: as.numeric("abc") produces a warning, not an error


# Solution
converted <- sapply(values, \(v) {
  tryCatch(
    {
      result <- suppressWarnings(as.numeric(v))
      if (is.na(result) && nchar(v) > 0) {
        warning("Not a number")
      }
      result
    },
    warning = \(w) NA
  )
})
converted
```

::: {.callout-warning title="Common Mistake"}
Catching warnings swallows the result:

```{r}
# tryCatch with warning handler does NOT return the expression result
result <- tryCatch(
  as.numeric(c("1", "abc", "3")),
  warning = \(w) "caught a warning"
)
result  # "caught a warning", not the numeric vector

# If you want the result AND to handle the warning, use withCallingHandlers
result <- withCallingHandlers(
  as.numeric(c("1", "abc", "3")),
  warning = \(w) invokeRestart("muffleWarning")
)
result  # c(1, NA, 3) - the result is preserved
```

`tryCatch` exits on the first condition. `withCallingHandlers` handles the condition and continues.
:::

::: {.callout-tip title="Challenge"}
Write a function `retry(expr, n = 3)` that attempts to evaluate an expression up to `n` times. If it fails all attempts, return the last error message. Test it with an expression that randomly fails.

```{r}
#| code-fold: true
#| code-summary: "Solution"

retry <- function(f, n = 3) {
  last_error <- NULL
  for (i in seq_len(n)) {
    result <- tryCatch(
      { return(f()) },
      error = \(e) {
        message("Attempt ", i, " failed: ", conditionMessage(e))
        e
      }
    )
    last_error <- result
  }
  paste("All", n, "attempts failed. Last error:", conditionMessage(last_error))
}

# Test: function that fails ~50% of the time
set.seed(1)
retry(\() {
  if (runif(1) < 0.5) stop("Random failure!")
  "Success!"
})
```
:::

## Part 3: Base R Cumulative Review

This part brings together everything from Sessions 1-4. No new material. All practice.

### Review: Data Creation and Structure

```{r}
students <- data.frame(
  name = c("Alice", "Bob", "Carlos", "Diana", "Elena", "Frank"),
  age = c(21, 23, 22, 20, 24, 21),
  major = c("CS", "Math", "CS", "Stats", "CS", "Math"),
  gpa = c(3.8, 3.2, 3.5, 3.9, 3.1, 3.6),
  year = c(3, 4, 3, 2, 4, 3)
)
```

### Your Turn: Integrated Exercises

```{r}
#| code-fold: true
#| code-summary: "Exercise 7: Subset and summarize"

# 1. Get all CS majors with GPA above 3.4
# 2. Calculate their mean age
# 3. Return the result as a named list with elements: students (data frame), mean_age (number)


# Solution
cs_high <- students[students$major == "CS" & students$gpa > 3.4, ]
result <- list(
  students = cs_high,
  mean_age = mean(cs_high$age)
)
result
```

```{r}
#| code-fold: true
#| code-summary: "Exercise 8: Apply + function writing"

# Write a function gpa_category(gpa) that returns:
# "High" for gpa >= 3.7, "Mid" for gpa >= 3.3, "Low" otherwise
# Apply it to the students data frame to create a new column


# Solution
gpa_category <- function(gpa) {
  ifelse(gpa >= 3.7, "High",
    ifelse(gpa >= 3.3, "Mid", "Low"))
}

students$category <- gpa_category(students$gpa)
students[, c("name", "gpa", "category")]
```

```{r}
#| code-fold: true
#| code-summary: "Exercise 9: tapply + string formatting"

# Use tapply to find the mean GPA per major
# Format each result as: "Major: X.XX" using sprintf
# Return as a character vector


# Solution
means <- tapply(students$gpa, students$major, mean)
sprintf("%s: %.2f", names(means), means)
```

```{r}
#| code-fold: true
#| code-summary: "Exercise 10: Loops + error handling"

# Process this list of student updates:
updates <- list(
  list(name = "Alice", new_gpa = 3.9),
  list(name = "George", new_gpa = 3.5),  # does not exist
  list(name = "Bob", new_gpa = "high"),   # invalid GPA
  list(name = "Carlos", new_gpa = 3.7)
)

# Loop through updates. For each:
# - Check if student exists in students$name
# - Check if new_gpa is numeric
# - If valid, update the GPA
# - If invalid, store an error message
# Return a list of error messages (or empty if all succeeded)


# Solution
errors <- character(0)

for (update in updates) {
  tryCatch({
    nm <- update$name
    new_gpa <- update$new_gpa
    
    if (!nm %in% students$name) {
      stop(paste("Student not found:", nm))
    }
    if (!is.numeric(new_gpa)) {
      stop(paste("Invalid GPA for", nm, ":", new_gpa))
    }
    
    students$gpa[students$name == nm] <- new_gpa
    message("Updated ", nm, " to GPA ", new_gpa)
  },
  error = \(e) {
    errors <<- c(errors, conditionMessage(e))
  })
}

errors
students[, c("name", "gpa")]
```

```{r}
#| code-fold: true
#| code-summary: "Exercise 11: Plotting from computed data"

# 1. Compute mean GPA by major using tapply
# 2. Create a barplot of the results
# 3. Add a horizontal line at the overall mean GPA
# 4. Title it "Mean GPA by Major"


# Solution
major_means <- tapply(students$gpa, students$major, mean)

barplot(major_means,
  main = "Mean GPA by Major",
  ylab = "Mean GPA",
  col = c("steelblue", "coral", "seagreen"),
  ylim = c(0, 4)
)
abline(h = mean(students$gpa), lty = 2, col = "red")
legend("topright", "Overall Mean", lty = 2, col = "red", bty = "n")
```

```{r}
#| code-fold: true
#| code-summary: "Exercise 12: Full pipeline"

# Build a complete data processing pipeline:
# 1. Create a numeric vector of 100 random exam scores (runif, 0 to 100)
# 2. Write a function grade(score) that returns A/B/C/D/F
# 3. Apply it to all scores
# 4. Count how many of each grade
# 5. Save the results as an RDS file
# 6. Read it back and create a barplot

# Do all of this inside tryCatch so the whole pipeline is safe


# Solution
set.seed(123)

tryCatch({
  # Step 1
  scores <- round(runif(100, 0, 100), 1)
  
  # Step 2
  grade <- function(score) {
    if (!is.numeric(score)) stop("Score must be numeric")
    if (score >= 90) "A"
    else if (score >= 80) "B"
    else if (score >= 70) "C"
    else if (score >= 60) "D"
    else "F"
  }
  
  # Step 3
  grades <- vapply(scores, grade, character(1))
  
  # Step 4
  counts <- table(factor(grades, levels = c("A", "B", "C", "D", "F")))
  
  # Step 5
  saveRDS(counts, "grade_counts.rds")
  
  # Step 6
  loaded <- readRDS("grade_counts.rds")
  barplot(loaded,
    main = "Grade Distribution (n = 100)",
    col = c("darkgreen", "steelblue", "goldenrod", "orange", "firebrick"),
    ylab = "Count"
  )
  
  message("Pipeline completed successfully")
  
  # Cleanup
  file.remove("grade_counts.rds")
},
error = \(e) cat("Pipeline failed:", conditionMessage(e), "\n"))
```

### Think About It

You have now worked through all of Base R fundamentals. Consider these questions:

1. When would you choose a list over a data frame? When would a matrix be better than a data frame?
2. You need to apply a function to each column of a data frame. Which approach do you use: a for loop, `lapply()`, or `vapply()`? What are the tradeoffs?
3. Your script reads 50 CSV files. Some might be corrupted. How do you structure your code so one bad file does not crash the entire script?

## Session Synthesis

### Connect the Dots

This exercise combines file I/O, error handling, functions, apply, and data manipulation.

```{r}
#| code-fold: true
#| code-summary: "Synthesis Exercise: Student Records System"

# Build a small student records system:

# 1. Create 3 separate data frames representing 3 "class sections"
section_a <- data.frame(
  name = c("Alice", "Bob"), gpa = c(3.8, 3.2), section = "A"
)
section_b <- data.frame(
  name = c("Carlos", "Diana"), gpa = c(3.5, 3.9), section = "B"
)
section_c <- data.frame(
  name = c("Elena", "Frank"), gpa = c(3.1, 3.6), section = "C"
)

# 2. Save each as a CSV file
sections <- list(A = section_a, B = section_b, C = section_c)
lapply(names(sections), \(nm) {
  write.csv(sections[[nm]], paste0("section_", nm, ".csv"), row.names = FALSE)
})

# 3. Write a function that reads all section files and combines them
read_all_sections <- function(files) {
  all_data <- lapply(files, \(f) {
    tryCatch(
      read.csv(f),
      error = \(e) {
        warning("Failed to read ", f, ": ", conditionMessage(e))
        NULL
      }
    )
  })
  
  # Remove NULL entries (failed reads)
  all_data <- Filter(\(x) !is.null(x), all_data)
  
  if (length(all_data) == 0) stop("No files could be read")
  
  do.call(rbind, all_data)
}

# 4. Read all sections (include a fake file to test error handling)
files <- c("section_A.csv", "section_B.csv", "section_FAKE.csv", "section_C.csv")
combined <- read_all_sections(files)
combined

# 5. Add a grade category and summarize
combined$honor <- ifelse(combined$gpa >= 3.5, "Yes", "No")
summary_by_section <- tapply(combined$gpa, combined$section, \(x) {
  sprintf("n=%d, mean=%.2f", length(x), mean(x))
})
summary_by_section

# 6. Save final result
saveRDS(combined, "all_students.rds")

# 7. Cleanup
file.remove("section_A.csv", "section_B.csv", "section_C.csv", "all_students.rds")
```

### Key Takeaways

- `read.csv()` and `write.csv()` handle tabular data. `readRDS()` / `saveRDS()` preserve R objects exactly.
- R has three signal types: `stop()` for errors, `warning()` for concerns, `message()` for info.
- `tryCatch()` gives you full control over error handling. `try()` is simpler for quick checks.
- Defensive programming means checking inputs, handling failures, and never assuming data is clean.
- You now have all the Base R tools. Sessions 5+ build on this foundation with the Tidyverse.

### Prepare for Next Session

Session 5 introduces the Tidyverse. You will:

- Meet tibbles (an improved data frame)
- Learn the pipe operator `|>` for chaining operations
- Use `dplyr` for data manipulation that replaces many Base R patterns you learned

No preparation needed. Everything you learned in Sessions 1-4 carries forward.
