---
title: "Session 6: Reshaping and Importing"
subtitle: "R Programming Workbook"
author: "R Programming Course"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-tools: true
    code-overflow: wrap
    theme: cosmo
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-note title="Learning Objectives"}
After completing this session, you will be able to:

1. Reshape data from wide to long using `pivot_longer()`
2. Reshape data from long to wide using `pivot_wider()`
3. Split and combine columns with `separate_wider_*()` and `unite()`
4. Import CSV, Excel, and statistical software files using readr, readxl, and haven
5. Specify column types and diagnose parsing problems
6. Manipulate strings using stringr functions
7. Detect, replace, split, and combine strings with literal patterns
8. Combine reshaping, importing, and string tools in data pipelines
:::

```{r}
#| label: setup
library(tidyverse)

set.seed(42)
sales <- tibble(
  date = rep(seq(as.Date("2024-01-01"), as.Date("2024-12-31"), by = "day"), 3),
  region = rep(c("North", "South", "West"), each = 366),
  product = sample(c("Widget", "Gadget", "Doohickey"), 366 * 3, replace = TRUE),
  units = rpois(366 * 3, lambda = 50),
  revenue = round(runif(366 * 3, 10, 500), 2),
  returned = rbinom(366 * 3, 1, 0.05)
)
```

## Part 1: Reshaping with tidyr

### pivot_longer(): Wide to Long

When data has values spread across column names, you need to make it longer:

```{r}
library(janitor)
airquality |> clean_names() |> summarise(avg_temp = mean(temp), .by = month) |> mutate(
  month = case_when(
    month == 5 ~ "May",
    month == 6 ~ "June",
    month == 7 ~ "July",
    month == 8 ~ "August",
    .default = "September"
    )
  ) |> pivot_wider(names_from = month, values_from = avg_temp)

```

```{r}
# Wide format: one column per quarter
quarterly_wide <- tibble(
  region = c("North", "South", "West"),
  Q1 = c(12500, 11800, 14200),
  Q2 = c(13100, 12400, 15000),
  Q3 = c(14800, 13200, 16100),
  Q4 = c(15200, 14000, 16800)
)
quarterly_wide
```

```{r}
# Pivot to long: one row per region-quarter combination
quarterly_long <- quarterly_wide |>
  pivot_longer(
    cols = Q1:Q4,
    names_to = "quarter",
    values_to = "revenue"
  )
quarterly_long
```

The `cols` argument uses the same selection helpers as `select()`:

```{r}
# These all do the same thing
quarterly_wide |> pivot_longer(cols = -region, names_to = "quarter", values_to = "revenue")
quarterly_wide |> pivot_longer(cols = starts_with("Q"), names_to = "quarter", values_to = "revenue")
```

### pivot_wider(): Long to Wide

```{r}
quarterly_long |>
  pivot_wider(
    names_from = quarter,
    values_from = revenue
  )
```

When multiple values exist per cell, use `values_fn`:

```{r}
# Summarize sales into a region x product table
sales |>
  pivot_wider(
    id_cols = region,
    names_from = product,
    values_from = revenue,
    values_fn = sum
  )
```

### separate and unite

```{r}
# Split a combined column
events <- tibble(
  event = c("2024-01-15_North_Widget", "2024-02-20_South_Gadget"),
  count = c(100, 85)
)

events |>
  separate_wider_delim(
    event,
    delim = "_",
    names = c("date", "region", "product")
  )
```

```{r}
# Combine columns
tibble(year = c(2024, 2024), month = c(1, 6), day = c(15, 20)) |>
  unite("date_str", year, month, day, sep = "-")
```

::: {.callout-note title="Base R vs Tidyverse"}
**Base R reshape:** `reshape(df, direction = "long", ...)` — works but has confusing argument names.

**tidyr:** `pivot_longer()` and `pivot_wider()` — clear, consistent, and well-documented.
:::

### Your Turn: Reshaping

**Exercise 1.1:** Pivot this exam data to long format with columns `student`, `exam`, and `score`:

```{r}
exams <- tibble(
  student = c("Alice", "Bob", "Carlos"),
  midterm = c(88, 72, 95),
  final = c(91, 80, 88),
  project = c(85, 90, 92)
)
```

```{r}
exams_long <- exams |> 
  pivot_longer(
    cols = -student,
    names_to = "exams",
    values_to = "score")

exams_long    
```
```{r}
#| code-fold: true
#| code-summary: "Solution"
exams |>
  pivot_longer(
    cols = -student,
    names_to = "exam",
    values_to = "score"
  )
```

**Exercise 1.2:** Take the `sales` dataset, compute average revenue per region per product, then pivot so each product is a column.

```{r}
sales_summary <- sales |> summarise(avg_rev = mean(revenue), .by = c(product,region)) |> pivot_wider(
  names_from = product,
  values_from = avg_rev
)

sales_summary
```
```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  summarize(avg_revenue = round(mean(revenue), 2), .by = c(region, product)) |>
  pivot_wider(names_from = product, values_from = avg_revenue)
```

**Exercise 1.3:** Separate this column into `city` and `state`:


```{r}
locations <- tibble(
  id = 1:4,
  place = c("Portland, OR", "Austin, TX", "Denver, CO", "Boston, MA")
)
```

```{r}
locations |> separate_wider_delim(place, delim=",", names=c("city","state"))
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
locations |>
  separate_wider_delim(place, delim = ", ", names = c("city", "state"))
```

**Exercise 1.4:** Create monthly summary of sales (total revenue per month per region), then pivot so each month is a column. Hint: create a month column with `format(date, "%b")`.

```{r}
sales_sum <- sales |> separate_wider_delim(date, delim="-", names=c("year","month", "day")) |>summarise(total_revenue = sum(revenue), .by = c(month, region)) |> pivot_wider(names_from=month, values_from=total_revenue)
sales_sum
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  mutate(month = format(date, "%b")) |>
  summarize(total = round(sum(revenue), 0), .by = c(region, month)) |>
  pivot_wider(names_from = month, values_from = total)
```

::: {.callout-warning title="Common Mistake"}
Forgetting to aggregate before `pivot_wider()`:

```{r}
#| eval: false
# This creates list-columns because multiple values exist per cell
sales |>
  pivot_wider(names_from = region, values_from = revenue)
```

Fix: Either summarize first, or use `values_fn`:

```{r}
#| eval: false
sales |>
  pivot_wider(
    id_cols = date,
    names_from = region,
    values_from = revenue,
    values_fn = sum
  )
```
:::

::: {.callout-tip title="Challenge"}
Starting from the `exams` data, compute each student's average score, add a column indicating if they passed (average >= 80), pivot to wide format with students as columns and exams as rows, then add a "mean" row at the bottom.

```{r}
#| code-fold: true
#| code-summary: "Solution"
exams_long <- exams |>
  pivot_longer(-student, names_to = "exam", values_to = "score")

# Add mean row
means <- exams_long |>
  summarize(score = mean(score), .by = student) |>
  mutate(exam = "mean")

bind_rows(exams_long, means) |>
  pivot_wider(names_from = student, values_from = score)
```
:::

## Part 2: Data Import

### Reading CSV Files with readr

```{r}
# Create a temp file for demonstration
temp <- tempfile(fileext = ".csv")
write_csv(sales |> head(50), temp)

# Read it back
imported <- read_csv(temp)
imported |> head(3)
```

readr automatically detects column types and shows the result:

```{r}
# Explicit column types for safety
read_csv(temp, col_types = cols(
  date = col_date(),
  region = col_character(),
  product = col_character(),
  units = col_integer(),
  revenue = col_double(),
  returned = col_integer()
)) |> head(3)
```

### Handling Messy Input

```{r}
# Data with problems
messy <- "name,age,score\nAlice,21,95\nBob,twenty-three,88\nCarlos,22,A+\n"

result <- read_csv(I(messy), col_types = cols(
  name = col_character(),
  age = col_integer(),
  score = col_double()
))
result
```

```{r}
# Check what went wrong
problems(result)
```

### Your Turn: Data Import

**Exercise 2.1:** Read this inline CSV data and specify appropriate column types:

```{r}
raw <- "product,price,in_stock,launch_date
Widget,29.99,TRUE,2024-03-15
Gadget,49.50,FALSE,2024-06-01
Doohickey,15.00,TRUE,2024-01-10
"
```

```{r}
data_raw <- read_csv(I(raw), col_names = cols(
  product = col_character(),
  price = col_double(),
  in_stock = col_logical(),
  launch_date = col_date()
))

data_raw
```
```{r}
#| code-fold: true
#| code-summary: "Solution"
read_csv(I(raw), col_types = cols(
  product = col_character(),
  price = col_double(),
  in_stock = col_logical(),
  launch_date = col_date()
))
```

**Exercise 2.2:** This CSV has some issues. Read it, check for problems, and fix the data:

```{r}
bad_data <- "id,value,category
1,100,A
2,N/A,B
3,250,C
4,175,
5,,A
"
```

```{r}
bad_data_df <- read_csv(I(bad_data), na = c("", "NA", "N/A")) |> mutate(
  value = replace_na(value,0),
  category = replace_na(category,"Unknown")
)

bad_data_df
problems(bad_data_df)
spec(bad_data_df)
```
```{r}
#| code-fold: true
#| code-summary: "Solution"
# Read with na argument to handle "N/A"
read_csv(I(bad_data), na = c("", "NA", "N/A")) |>
  mutate(
    value = replace_na(value, 0),
    category = replace_na(category, "Unknown")
  )
```

**Exercise 2.3:** Write the `sales` summary (total revenue by region) to a temporary CSV file, then read it back and verify the data matches.

```{r}
sales_summary <- sales |> summarise(total_rev = sum(revenue), .by = region)
temp <-tempfile(fileext=".csv")
write_csv(sales_summary, temp) 

read_csv(temp)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
# Write
summary_data <- sales |>
  summarize(total_revenue = sum(revenue), .by = region)

temp_out <- tempfile(fileext = ".csv")
write_csv(summary_data, temp_out)

# Read back
read_csv(temp_out)
```

::: {.callout-warning title="Common Mistake"}
Using `read.csv()` (base R) instead of `read_csv()` (readr):

```{r}
# Base R: converts strings to factors (in older R), slower, returns data.frame
base_result <- read.csv(temp)
class(base_result)

# readr: returns tibble, shows types, faster
readr_result <- read_csv(temp, show_col_types = FALSE)
class(readr_result)
```

Use `read_csv()` (with underscore) in tidyverse workflows.
:::

::: {.callout-tip title="Challenge"}
Create a complete import-clean-reshape pipeline: Read the inline CSV below, parse dates, pivot to long format, and compute the average value per metric.

```{r}
multi_metric <- "date,temperature,humidity,pressure
2024-01-01,22.5,65,1013
2024-01-02,23.1,62,1015
2024-01-03,21.8,70,1010
2024-01-04,24.0,58,1012
"
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
read_csv(I(multi_metric), col_types = cols(date = col_date())) |>
  pivot_longer(
    cols = -date,
    names_to = "metric",
    values_to = "value"
  ) |>
  summarize(avg_value = mean(value), .by = metric)
```
:::

## Part 3: String Manipulation with stringr

### Core stringr Functions

```{r}
library(stringr)

fruits <- c("apple", "banana", "cherry", "date", "elderberry")

# Basic operations
str_length(fruits)
str_to_upper(fruits)
str_to_lower(fruits)
str_to_title(fruits)
str_sub(fruits, 1, 3)
str_pad(fruits, width = 12, side = "right", pad = ".")
```

### Detecting and Locating Text

`str_detect()` checks if a substring exists. `str_starts()` and `str_ends()` check prefixes and suffixes.

```{r}
# Detect a literal substring
str_detect(fruits, "an")       # contains "an"?
str_detect(fruits, "berry")    # contains "berry"?

# Check prefix and suffix
str_starts(fruits, "ch")       # starts with "ch"?
str_ends(fruits, "e")          # ends with "e"?

# Count occurrences of a literal substring
str_count(fruits, "r")         # how many "r"s?
str_count(c("banana", "aardvark"), "a")

# Locate position of a substring
str_locate(fruits, "rr")       # position of "rr"
```

### Replacing and Removing Text

```{r}
cities <- c("New York City", "Los Angeles City", "San Francisco Bay")

# Replace first occurrence
str_replace(cities, "City", "Metro")

# Replace all occurrences
messy <- c("one--two--three", "a--b")
str_replace_all(messy, "--", "-")

# Remove text (replace with nothing)
str_remove(cities, " City")
str_remove_all(c("a.b.c", "x.y"), ".")  # careful: removes all characters!
str_remove_all(c("a.b.c", "x.y"), fixed("."))  # correct: use fixed() for literal dots
```

Use `fixed()` when your pattern contains characters like `.`, `$`, `(`, `)`, `*` that have special meaning. We will learn about these special meanings in Session 14 when we cover regular expressions.

### Splitting and Extracting

```{r}
# Split strings by a delimiter
str_split("apple-banana-cherry", "-")

# str_split returns a list; simplify for a single string
str_split_1("apple-banana-cherry", "-")

# Extract words by position
sentences <- c("The quick brown fox", "R is great for data")
word(sentences, 1)       # first word
word(sentences, -1)      # last word
word(sentences, 2, 3)    # words 2 through 3
```

### Combining Strings

```{r}
# str_c (like paste0)
str_c("Hello", "World", sep = " ")

# str_glue (like Python f-strings)
name <- "Alice"
score <- 95
str_glue("{name} scored {score} out of 100")

# Vectorized
str_c("Student_", 1:5)
```

### Cleaning Strings

```{r}
# Trim leading/trailing whitespace
str_trim("  hello world  ")

# Squish: trim + collapse internal whitespace
str_squish("  too   many    spaces  ")

# Substring extraction by position
codes <- c("PRD-001", "PRD-042", "PRD-100")
str_sub(codes, 5, 7)  # characters 5 through 7
```

::: {.callout-note title="Base R vs Tidyverse"}
**Base R:** `paste()` concatenates strings. `grep("text", x)` returns indices. `grepl("text", x)` returns logical. `gsub("old", "new", x)` replaces all. Argument order varies.

**stringr:** `str_c()` concatenates. `str_detect(x, "text")` returns logical. `str_replace_all(x, "old", "new")` replaces all. String is always first, making it pipe-friendly.
:::

### Your Turn: String Manipulation

**Exercise 3.1:** Given this vector, extract just the numeric part from each string:

```{r}
codes <- c("PRD-001", "PRD-042", "PRD-100", "PRD-007")
```

Hint: Use `str_split_1()` or `str_sub()` to extract by position.

```{r}
split_data <- sapply(str_split(codes,"-"), \(x) x[2])
str_sub(codes, 5,7)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
# Method 1: split on "-" and take the second part
sapply(str_split(codes, "-"), \(x) x[2])

# Method 2: extract by position (characters 5-7)
str_sub(codes, 5, 7)
```

**Exercise 3.2:** Clean these messy names: trim whitespace, convert to title case, replace multiple spaces with a single space.

```{r}
messy_names <- c("  alice  smith  ", "BOB JONES", "  carlos   garcia")
```

```{r}
messy_names |> str_trim() |> str_squish() |> str_to_title()
```
```{r}
#| code-fold: true
#| code-summary: "Solution"
messy_names |>
  str_trim() |>
  str_squish() |>
  str_to_title()
```

**Exercise 3.3:** From the `sales` dataset, create a `label` column that combines region and product as "Region: Product" (e.g., "North: Widget").

```{r}
sales |> mutate(
  label = str_glue("{region}: {product}")
) |> distinct(label)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  mutate(label = str_glue("{region}: {product}")) |>
  distinct(label) |>
  arrange(label)
```

**Exercise 3.4:** Filter `sales` to only rows where the product name contains "gadget" (case-insensitive). Hint: convert to lowercase first, then detect.

```{r}
sales |> filter(str_detect(str_to_lower(product),"gadget"))
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  filter(str_detect(str_to_lower(product), "gadget")) |>
  count(product)
```

**Exercise 3.5:** Parse these log entries to extract the timestamp, level, and message using `str_sub()` and `str_split()`:

```{r}
logs <- c(
  "[2024-01-15 08:30:00] INFO: Server started",
  "[2024-01-15 08:31:15] WARN: High memory usage",
  "[2024-01-15 08:32:00] ERROR: Connection refused"
)
```

Hint: The timestamp always spans characters 2-20. Use `str_remove()` and `str_split_1()` for the rest.

```{r}
raw <- tibble(logs)
raw
cleaned_logs <- raw |> mutate(
  date = str_sub(logs,2,20),
  rest = str_sub(logs,23),
  level = word(rest,1) |> str_remove(":"),
  message = str_remove(rest, word(rest,1)) |> str_trim()
) |> select(-rest)
cleaned_logs
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
tibble(raw = logs) |>
  mutate(
    timestamp = str_sub(raw, 2, 20),
    rest = str_sub(raw, 23),
    level = word(rest, 1) |> str_remove(":"),
    message = str_remove(rest, word(rest, 1)) |> str_trim()
  ) |>
  select(-rest)
```

::: {.callout-warning title="Common Mistake"}
Forgetting `fixed()` when replacing special characters:

```{r}
str_remove_all("price: $10.99", ".")  # removes everything!
str_remove_all("price: $10.99", fixed("."))  # removes only dots
```

Characters like `.`, `$`, `(`, `)`, `*`, `+`, `?` have special meaning in pattern matching. When you want to match them literally, wrap the pattern in `fixed()`. We will learn what these special meanings are in Session 14.
:::

::: {.callout-tip title="Challenge"}
Parse these file paths into a tibble with columns for directory, filename, and extension:

```{r}
paths <- c(
  "data/raw/sales.csv",
  "output/figures/chart.png",
  "R/analysis.R",
  "docs/report.qmd"
)
```

Hint: Split on `"/"` for directory parts and `"."` for the extension.

```{r}
#| code-fold: true
#| code-summary: "Solution"
tibble(path = paths) |>
  mutate(
    parts = str_split(path, "/"),
    filename_ext = sapply(parts, \(x) x[length(x)]),
    directory = sapply(parts, \(x) str_c(x[-length(x)], collapse = "/")),
    filename = word(filename_ext, 1, sep = fixed(".")),
    extension = word(filename_ext, -1, sep = fixed("."))
  ) |>
  select(directory, filename, extension)
```
:::

## Session Synthesis

### Connect the Dots

Build a data cleaning pipeline that combines all three skills. Start with this messy data:

```{r}
messy_sales <- "date|rep_name|region_product|amount
2024-01-15|  Alice Smith  |North_Widget|$1,250.00
2024-01-16|BOB JONES|South_Gadget|$890.50
2024-01-17|  carlos garcia |West_Doohickey|$2,100.75
2024-01-18|DIANA LEE|North_Gadget|$445.00
"
```

Your pipeline should:

1. Read the pipe-delimited data
2. Clean the rep names (trim, title case)
3. Separate region_product into two columns
4. Remove the "$" and "," from amount and convert to numeric
5. Compute total and average amount per region

```{r}
read_delim(I(messy_sales), delim = "|") |>
  mutate(
    rep_name = rep_name |> str_trim() |> str_squish() |> str_to_title()
  ) |>
  separate_wider_delim(region_product, delim = "_", names = c("region", "product")) |> mutate(
    amount = amount |> str_remove_all(fixed("$")) |> str_remove_all(fixed(",")) |> as.numeric()
  ) |> summarise(
    total_amount = sum(amount),
    average_amount = mean(amount),
    .by = region
  )
  


```

```{r}
#| code-fold: true
#| code-summary: "Solution"
read_delim(I(messy_sales), delim = "|") |>
  mutate(
    rep_name = rep_name |> str_trim() |> str_squish() |> str_to_title()
  ) |>
  separate_wider_delim(region_product, delim = "_", names = c("region", "product")) |>
  mutate(
    amount = amount |> str_remove_all(fixed("$")) |> str_remove_all(",") |> as.numeric()
  ) |>
  summarize(
    total_amount = sum(amount),
    avg_amount = mean(amount),
    .by = region
  )
```

### Key Takeaways

1. `pivot_longer()` stacks columns into rows (wide to long). `pivot_wider()` spreads rows into columns (long to wide).
2. readr's `read_csv()` is the standard import tool. Always check `problems()` and specify `col_types` for safety.
3. stringr provides consistent, pipe-friendly string functions. Every function starts with `str_` and takes the string as the first argument.
4. These three skills combine into powerful data cleaning pipelines: import, reshape, clean strings.

### Prepare for Next Session

**Session 7: Specialized Data Types**

You will learn:

- forcats for factor handling (ordered categories)
- lubridate for dates and times
- purrr for functional programming with lists

Run `library(forcats)`, `library(lubridate)`, and `library(purrr)` to confirm they load.
