---
title: "Session 3: Patterns and Output"
subtitle: "R Programming Workbook"
author: "R Programming Course"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-tools: true
    code-overflow: wrap
    theme: cosmo
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-note title="Learning Objectives"}
By the end of this session, you will be able to:

**Part 1: Apply Family**

- Replace loops with `lapply`, `sapply`, `vapply`
- Summarize data by groups with `tapply`
- Iterate over multiple inputs with `mapply`

**Part 2: String Handling**

- Build strings with `paste()` and `sprintf()`
- Find and replace patterns with `grep()`, `gsub()`
- Split and clean strings

**Part 3: Base R Plotting**

- Create scatter plots, histograms, bar plots, and boxplots
- Customize with colors, labels, legends, and multi-panel layouts
- Save plots to files
:::

```{r}
students <- data.frame(
  name = c("Alice", "Bob", "Carlos", "Diana", "Elena", "Frank"),
  age = c(21, 23, 22, 20, 24, 21),
  major = c("CS", "Math", "CS", "Stats", "CS", "Math"),
  gpa = c(3.8, 3.2, 3.5, 3.9, 3.1, 3.6),
  year = c(3, 4, 3, 2, 4, 3)
)
```

## Part 1: The Apply Family

### 1.1 lapply: Returns a List

`lapply` applies a function to each element and always returns a list:

```{r}
lapply(1:4, \(x) x^2)
```

On a data frame, it iterates over columns:

```{r}
lapply(students[, c("age", "gpa", "year")], mean)
```

### 1.2 sapply: Simplifies the Output

```{r}
sapply(students[, c("age", "gpa", "year")], mean)
```

`sapply` tries to return a vector or matrix. Convenient but unpredictable.

### 1.3 vapply: Type-Safe

```{r}
vapply(students[, c("age", "gpa", "year")], mean, numeric(1))
```

You specify the expected output type. Use `vapply` in scripts and functions.

### 1.4 tapply: Group Summaries

```{r}
tapply(students$gpa, students$major, mean)
tapply(students$age, students$year, range)
```

### 1.5 mapply: Multiple Inputs

```{r}
mapply(\(name, gpa) paste(name, "has GPA", gpa),
       students$name, students$gpa)
```

### Your Turn: Exercise 1.1

Use `sapply` to compute the standard deviation of each numeric column in `students`.
```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false

sapply(students[,c("year", "gpa", "age")],\(x) round(sd(x), digits=2))
```
```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
sapply(students[, c("age", "gpa", "year")], sd)
```

### Your Turn: Exercise 1.2

Use `tapply` to find:

1. The maximum GPA in each major
2. The number of students in each year
```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
tapply(students$gpa, students$major, max)
tapply(students$name, students$year, length)
```
```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
tapply(students$gpa, students$major, max)
tapply(students$name, students$year, length)
```

### Your Turn: Exercise 1.3

Use `lapply` to apply a custom function to each numeric column that returns a named vector with `mean`, `min`, and `max`. Then convert the result to a data frame using `do.call(rbind, ...)`.
```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
results <- lapply(students[,c("year", "age", "gpa")], \(x) c(mean= mean(x),min=  min(x), max = max(x)))
do.call(rbind,results)

```

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
result <- lapply(students[, c("age", "gpa", "year")], \(x) {
  c(mean = mean(x), min = min(x), max = max(x))
})
do.call(rbind, result)
```

### Your Turn: Exercise 1.4

Use `mapply` to create labels like "Alice (CS, Year 3)" for each student.
```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
mapply(\(x,y,z) sprintf("%s (%s, Year %d)",x,y,z), students$name, students$major, students$year)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
mapply(\(n, m, y) sprintf("%s (%s, Year %d)", n, m, y),
       students$name, students$major, students$year)
```

::: {.callout-warning title="Common Mistake"}
**Using `sapply` when output length varies.**

```{r}
sapply(list(a = 1:3, b = 1:5), range)  # matrix (OK)
sapply(list(a = 1:3, b = 1:5), \(x) x[x > 2])  # list (surprise!)
```

When results have different lengths, `sapply` falls back to a list. Use `lapply` if you expect varied lengths.
:::

::: {.callout-tip title="Challenge"}
Write a function `column_report` that takes a data frame and uses `vapply` to return a character vector describing each column: "colname: type (n unique values)".

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
column_report <- function(df) {
  vapply(names(df), \(col) {
    sprintf("%s: %s (%d unique)", col, class(df[[col]]), length(unique(df[[col]])))
  }, character(1))
}
column_report(students)
```
:::

---

## Part 2: String Handling

### 2.1 Building Strings

```{r}
paste("Hello", "World")
paste("Hello", "World", sep = "-")
paste0("item_", 1:5)

fruits <- c("apple", "banana", "cherry")
paste(fruits, collapse = ", ")
```

`sep` goes between arguments. `collapse` joins vector elements into one string.

### 2.2 Formatted Strings with sprintf()

```{r}
sprintf("Name: %s, GPA: %.2f", "Alice", 3.812)
sprintf("Student %d of %d", 1:3, 6)
```

Format codes: `%s` = string, `%d` = integer, `%f` = float, `%.2f` = 2 decimals.

### 2.3 Pattern Matching

```{r}
emails <- c("alice@uni.edu", "bob@gmail.com", "carlos@uni.edu", "diana@yahoo.com")

grep("uni", emails)               # positions
grep("uni", emails, value = TRUE)  # values
grepl("uni", emails)               # TRUE/FALSE
```

### 2.4 Substitution

```{r}
x <- "The quick brown fox"
sub("quick", "slow", x)   # first match
gsub("o", "0", x)          # all matches

# Clean data
prices <- c("$10.99", "$25.50", "$7.00")
as.numeric(gsub("$", "", prices, fixed = TRUE))
```

### 2.5 Splitting Strings

```{r}
strsplit("2024-01-15", "-")[[1]]
strsplit(c("a,b,c", "d,e"), ",")
```

`strsplit` returns a list. Use `[[1]]` for a single string input.

### Your Turn: Exercise 2.1

Given this vector of full names:

```{r}
full_names <- c("Alice Smith", "Bob Jones", "Carlos Garcia", "Diana Chen")
```

1. Extract just the first names (split on space, take first element)
```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
sapply(strsplit(full_names, " "), \(x) x[1])
```
2. Create email addresses: firstname.lastname@university.edu (all lowercase)
```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false

parts <- sapply(tolower(full_names), \(x)strsplit(x, " "))
sapply(parts, \(x) {
  sprintf("%s.%s@university.edu", x[1],x[2])})

```

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
# 1
sapply(strsplit(full_names, " "), \(x) x[1])

# 2
parts <- strsplit(tolower(full_names), " ")
sapply(parts, \(x) paste0(x[1], ".", x[2], "@university.edu"))
```

### Your Turn: Exercise 2.2

Clean this messy data:

```{r}
messy <- c("  New York  ", "los angeles", "CHICAGO", "  san Francisco")
```

Produce: `"New York", "Los Angeles", "Chicago", "San Francisco"`
```{r}
cleaned <- trimws(messy)
cleaned <- tolower(cleaned)
sapply(strsplit(cleaned, " "), \(words) {
  paste(sapply(words, \(w) {
    paste0(toupper(substring(w, 1, 1)), substring(w, 2))
  }), collapse = " ")
})
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
clean <- trimws(messy)
clean <- tolower(clean)
# Capitalize first letter of each word
sapply(strsplit(clean, " "), \(words) {
  paste(sapply(words, \(w) {
    paste0(toupper(substring(w, 1, 1)), substring(w, 2))
  }), collapse = " ")
})
```

### Your Turn: Exercise 2.3

Given this log data:

```{r}
logs <- c("2024-01-15 ERROR: file not found",
          "2024-01-15 INFO: process started",
          "2024-01-16 ERROR: connection timeout",
          "2024-01-16 INFO: process complete",
          "2024-01-17 WARNING: low memory")
```

1. Find all ERROR logs using `grepl`
2. Extract just the dates from error logs
3. Replace "ERROR" with "ERR" in all logs
```{r}
errors <- logs[grepl("ERROR", logs)]
```
```{r}
sapply(logs[grepl("ERROR", logs)], \(x) {
  sapply(strsplit(x," "), \(y) y[1])
})
```
```{r}
substring(errors, 1,10)
logs <- gsub("ERROR", "ERR", logs)
logs
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
# 1
error_logs <- logs[grepl("ERROR", logs)]
error_logs

# 2
substring(error_logs, 1, 10)

# 3
gsub("ERROR", "ERR", logs)
```

::: {.callout-warning title="Common Mistake"}
**Forgetting that `gsub()` uses pattern matching by default.**

```{r}
gsub(".", "!", "Hello.")  # replaces everything (not just the dot)!
gsub(".", "!", "Hello.", fixed = TRUE)  # correct: use fixed = TRUE for literal text
```

By default, `gsub()` treats the pattern as a regular expression (which we will cover in Session 14). For now, add `fixed = TRUE` when you want to match exact text like `"."`, `"$"`, or `"("`.
:::

---

## Part 3: Base R Plotting

### 3.1 Scatter Plot

```{r}
#| fig-width: 7
#| fig-height: 5
plot(students$age, students$gpa,
     main = "Age vs GPA",
     xlab = "Age", ylab = "GPA",
     pch = 16, col = "steelblue", cex = 1.5)
```

Key arguments: `pch` = point shape, `col` = color, `cex` = size multiplier.

### 3.2 Line Plot

```{r}
#| fig-width: 7
#| fig-height: 5
x <- 1:50
y <- cumsum(rnorm(50))
plot(x, y, type = "l", col = "darkred", lwd = 2,
     main = "Random Walk", xlab = "Step", ylab = "Position")
abline(h = 0, lty = 2, col = "gray50")
```

### 3.3 Histogram

```{r}
#| fig-width: 7
#| fig-height: 5
set.seed(42)
data <- rnorm(500, mean = 100, sd = 15)
hist(data, breaks = 25, col = "lightblue", border = "black",
     main = "Test Scores", xlab = "Score")
abline(v = mean(data), col = "red", lwd = 2)
```

### 3.4 Bar Plot

```{r}
#| fig-width: 7
#| fig-height: 5
counts <- table(students$major)
barplot(counts, col = c("steelblue", "coral", "seagreen"),
        main = "Students by Major", ylab = "Count")
```

### 3.5 Boxplot

```{r}
#| fig-width: 7
#| fig-height: 5
boxplot(gpa ~ major, data = students,
        col = c("lightblue", "lightgreen", "lightyellow"),
        main = "GPA by Major", ylab = "GPA")
```

### 3.6 Multi-Panel Layouts

```{r}
#| fig-width: 10
#| fig-height: 5
par(mfrow = c(1, 2))
hist(students$gpa, col = "lightblue", main = "GPA Distribution")
barplot(table(students$year), col = "coral", main = "Students by Year")
par(mfrow = c(1, 1))
```

### 3.7 Adding Layers

```{r}
#| fig-width: 7
#| fig-height: 5
plot(students$age, students$gpa, type = "n",
     main = "Students by Major", xlab = "Age", ylab = "GPA")

colors <- c(CS = "red", Math = "blue", Stats = "green")
for (m in unique(students$major)) {
  sub <- students[students$major == m, ]
  points(sub$age, sub$gpa, col = colors[m], pch = 16, cex = 1.5)
}
legend("bottomright", legend = names(colors), col = colors, pch = 16)
```

### Your Turn: Exercise 3.1

Create a histogram of 1000 random values from a normal distribution with mean=0, sd=1. Add a vertical line at the mean. Use `breaks = 30` and a color of your choice.

```{r}
data <- rnorm(1000, mean= 0, sd= 1)
hist(data,col="red", breaks=30, main="Normal distribution", xlab="Generatedd numbers", border="black")
abline(v=mean(data), col="gray", lwd=2, lty = 2)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
set.seed(123)
data <- rnorm(1000)
hist(data, breaks = 30, col = "steelblue", border = "white",
     main = "Standard Normal", xlab = "Value")
abline(v = mean(data), col = "red", lwd = 2, lty = 2)
```

### Your Turn: Exercise 3.2

Create a 2x2 panel of plots using the `students` data:

1. Top left: scatter plot of age vs GPA
2. Top right: bar plot of major counts
3. Bottom left: boxplot of GPA by year
4. Bottom right: barplot of year counts

```{r}
par(mfrow = c(2,2), mar = c(4,4,3,2))
plot(students$age, students$gpa, col = "red", main="Age vs GPA", xlab="age", ylab="gpa", pch=16)
barplot(table(students$major), col="blue", main="Major distribution", xlab="Major", ylab="Count")
boxplot(gpa ~year, data = students, col="coral", xlab="Year", main="gpa by year")
barplot(table(students$year), main="Distribution of years", col="pink", xlab="years", ylab="count", 
border="purple")
par(mfrow=c(1,1))
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(students$age, students$gpa, pch = 16, col = "steelblue",
     main = "Age vs GPA", xlab = "Age", ylab = "GPA")
barplot(table(students$major), col = "coral", main = "Majors")
boxplot(gpa ~ year, data = students, col = "lightgreen", main = "GPA by Year")
barplot(table(students$year), col = "lightyellow", main = "Years")
par(mfrow = c(1, 1))
```

### Your Turn: Exercise 3.3

Create a scatter plot of `mtcars$wt` (weight) vs `mtcars$mpg` (fuel efficiency). Color the points by the number of cylinders (`mtcars$cyl`). Add a legend.

```{r}
colors <- c("4" = "blue", "6" = "green", "8" = "red")
plot(mtcars$wt, mtcars$mpg, pch = 16, col = colors[as.character(mtcars$cyl)])
legend("topright",legend = paste(names(colors), "cyl"), col = colors, pch = 16)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
colors <- c("4" = "blue", "6" = "green", "8" = "red")
plot(mtcars$wt, mtcars$mpg, pch = 16,
     col = colors[as.character(mtcars$cyl)],
     main = "Weight vs MPG by Cylinders",
     xlab = "Weight (1000 lbs)", ylab = "MPG")
legend("topright", legend = paste(names(colors), "cyl"),
       col = colors, pch = 16)
```

::: {.callout-tip title="Challenge"}
Create a "dashboard" with 4 plots about the built-in `iris` dataset:

1. Scatter: Sepal.Length vs Sepal.Width colored by Species
2. Boxplot: Petal.Length by Species
3. Histogram: Petal.Width
4. Bar plot: count of each Species

Include titles, colors, and legends where appropriate.

```{r}
iris
colors <- c("virginica" = "red", "versicolor" = "pink", "setosa" = "purple")
par(mfrow = c(2,2))
plot(iris$Sepal.Length, iris$Sepal.Width, col= colors[iris$Species], cex=1, pch= 16, main="length vs width", xlab = "length", ylab = "width")
boxplot(iris$Petal.Length ~ iris$Species, data = iris, col = "red")
hist(iris$Petal.Width, col= "red")
barplot(table(iris$Species), col="blue")
par(mfrow = c(1,1))


```

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# 1
cols <- c(setosa = "red", versicolor = "blue", virginica = "green")
plot(iris$Sepal.Length, iris$Sepal.Width, pch = 16,
     col = cols[iris$Species], main = "Sepal Dimensions",
     xlab = "Length", ylab = "Width")
legend("topright", legend = names(cols), col = cols, pch = 16, cex = 0.7)

# 2
boxplot(Petal.Length ~ Species, data = iris,
        col = c("lightcoral", "lightblue", "lightgreen"),
        main = "Petal Length by Species")

# 3
hist(iris$Petal.Width, breaks = 20, col = "steelblue",
     main = "Petal Width Distribution", xlab = "Width")

# 4
barplot(table(iris$Species), col = c("lightcoral", "lightblue", "lightgreen"),
        main = "Species Count")

par(mfrow = c(1, 1))
```
:::

---

## Session Synthesis

### Connect the Dots

Build a text analysis of the `students` data using skills from all three parts:

1. Use `tapply` to compute mean GPA by major
2. Use `sprintf` to create a formatted report string for each major: "Major: X | Avg GPA: X.XX | Count: X"
3. Create a two-panel plot: (a) bar plot of mean GPA by major, (b) scatter plot of age vs GPA colored by major with legend
4. Use `mapply` to create student labels: "Name (Major, GPA: X.X)"

```{r}
avg_gpa <- tapply(students$gpa, students$major, mean)
count <- tapply(students$gpa, students$major, length)

mapply(\(major, avg, n) {sprintf("Major: %s | Avg GPA: %.2f | Count: %d", major, avg, n)}, names(avg_gpa), avg_gpa, count)

color <- c(CS = "red", Math = "blue", Stats = "orange")

par(mfrow= c(1,2))
barplot(avg_gpa, major = "Average gpa", col = "blue")
plot(students$age, students$gpa, pch = 16, col = color[students$major], cex = 1)
par(mfrow=c(1,1))
mapply(\(name, major, gpa) {sprintf("%s: (%s, GPA: %2.f)", name, major,gpa)}, students$name, students$major, students$gpa)

```

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
# 1
avg_gpa <- tapply(students$gpa, students$major, mean)
counts <- tapply(students$name, students$major, length)

# 2
mapply(\(major, avg, n) {
  sprintf("Major: %s | Avg GPA: %.2f | Count: %d", major, avg, n)
}, names(avg_gpa), avg_gpa, counts)

# 3
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
barplot(avg_gpa, col = "steelblue", main = "Mean GPA by Major",
        ylab = "GPA", ylim = c(0, 4))
colors <- c(CS = "red", Math = "blue", Stats = "green")
plot(students$age, students$gpa, pch = 16,
     col = colors[students$major],
     main = "Age vs GPA", xlab = "Age", ylab = "GPA")
legend("bottomright", legend = names(colors), col = colors, pch = 16)
par(mfrow = c(1, 1))

# 4
mapply(\(n, m, g) sprintf("%s (%s, GPA: %.1f)", n, m, g),
       students$name, students$major, students$gpa)
```

### Key Takeaways

1. Use `vapply` in scripts (type-safe), `sapply` in console (convenient), `tapply` for groups.
2. `paste()` concatenates, `sprintf()` formats, `grep()`/`gsub()` find and replace patterns.
3. Base R plots build in layers: start with `plot()`, add with `points()`, `lines()`, `legend()`.
4. `par(mfrow)` creates multi-panel layouts.

### Prepare for Next Session

**Session 4: Files, Errors, and Review**

We will read and write data files, handle errors with `tryCatch`, and do a cumulative review of Sessions 1-3. Make sure all concepts from today are solid.
