---
title: "Session 1: Welcome to R"
subtitle: "R Programming Workbook"
author: "R Programming Course"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-tools: true
    code-overflow: wrap
    theme: cosmo
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-note title="Learning Objectives"}
By the end of this session, you will be able to:

**Part 1: R Environment**

- Navigate Positron IDE and use the R console
- Assign variables and check types
- Access help documentation for any function

**Part 2: Atomic Vectors**

- Create vectors with `c()`, `seq()`, and `rep()`
- Perform vectorized operations without loops
- Handle type coercion and missing values (`NA`)

**Part 3: Data Structures**

- Build and index matrices
- Create lists and extract elements with `$` and `[[`
- Construct and manipulate data frames
:::

## Part 1: R Environment Setup

### 1.1 The Console as Calculator

R evaluates expressions and prints the result:

```{r}
2 + 3
10 / 3
2^8
17 %% 5    # modulo (remainder)
17 %/% 5   # integer division
```

`%%` gives the remainder. `%/%` gives the quotient without the remainder. Both differ from the `/` operator used in Python for integer division.

### 1.2 Assignment

Use `<-` to assign values to names:

```{r}
x <- 42
greeting <- "Hello, R"
x
greeting
```

The shortcut for `<-` in Positron is Alt+- (Option+- on Mac). 

You can also use `=` for assignment, but `<-` is the R convention and what you will see in all R code.

### 1.3 Checking Types

```{r}
typeof(42)
typeof(42L)
typeof("hello")
typeof(TRUE)
```

Notice that `42` is a "double" (floating point), not an integer. To get an integer, append `L`: `42L`.

```{r}
class(42)
class(42L)
```

`typeof()` tells you internal storage. `class()` tells you the object-oriented class. For basic types they align, but they diverge for more complex objects.

### 1.4 Getting Help

```{r}
#| eval: false
?mean            # open help for mean()
help(mean)       # same thing
??regression     # search all help pages
example(mean)    # run the examples from help
```

Every function in R has a help page. The most useful section is **Examples** at the bottom.

### Your Turn: Exercise 1.1

Compute the following in R:

1. 15 raised to the power of 3
2. The remainder when 256 is divided by 7
3. Assign your name to a variable called `my_name` and your age to `my_age`
4. Use `typeof()` to check the type of each variable

```{r}
15^3
256 %% 7
my_name <- "Zsófi"
my_age <- 21
typeof(my_name)
typeof(my_age)
```

```{r}
#| code-fold: true
#| code-summary: "Hint"
#| eval: false
# 1
15^3
# 2
256 %% 7
# 3
my_name <- "Your Name"
my_age <- 22
# 4
typeof(my_name)
typeof(my_age)
```

### Your Turn: Exercise 1.2

Use `?sqrt` to open the help page for `sqrt()`. Then:

1. Compute the square root of 144
2. Compute the square root of -1. What happens?

```{r}
sqrt(144)
sqrt(-1) 
```

```{r}
#| code-fold: true
#| code-summary: "Hint"
#| eval: false
sqrt(144)
sqrt(-1)  # Returns NaN with a warning
```

::: {.callout-warning title="Common Mistake"}
**Forgetting that R is case-sensitive.**

```{r}
#| error: true
X <- 10
x  # This is a different variable
```

`X` and `x` are two different objects. R is case-sensitive for everything: variable names, function names, and arguments.
:::

---

## Part 2: Atomic Vectors

### 2.1 Creating Vectors

The `c()` function combines values into a vector:

```{r}
nums <- c(10, 20, 30, 40, 50)
nums

words <- c("apple", "banana", "cherry")
words

flags <- c(TRUE, FALSE, TRUE, TRUE)
flags
```

There are no scalars in R. Even a single number is a vector of length 1:

```{r}
x <- 5
is.vector(x)
length(x)
```

### 2.2 Sequences and Repetition

```{r}
# Colon operator for integer sequences
1:10

# seq() for more control
seq(0, 1, by = 0.25)
seq(1, 100, length.out = 5)

# rep() for repetition
rep(0, times = 5)
rep(c(1, 2, 3), times = 3)
rep(c(1, 2, 3), each = 3)
```

Notice the difference: `times` repeats the whole vector, `each` repeats each element before moving to the next.

### 2.3 Vectorized Arithmetic

Operations apply to every element at once:

```{r}
x <- c(2, 4, 6, 8, 10)
x * 3
x + 100
x^2
sqrt(x)
```

No loops needed. This is faster and more readable than looping.

### 2.4 Element-wise Operations

Two vectors of the same length get paired element by element:

```{r}
prices <- c(10, 20, 30)
quantities <- c(5, 3, 8)
prices * quantities
```

### 2.5 Recycling

When vectors have different lengths, the shorter one repeats:

```{r}
x <- c(1, 2, 3, 4, 5, 6)
y <- c(10, 20)
x + y
```

`y` recycles: `10, 20, 10, 20, 10, 20`. This works silently when the lengths divide evenly.

```{r}
x <- c(1, 2, 3, 4, 5)
y <- c(10, 20)
x + y
```

When they do not divide evenly, R gives a warning.

### 2.6 Type Coercion

Vectors can only hold one type. When you mix types, R coerces to the most general type:

```{r}
c(1, 2, "three")      # all become character
c(TRUE, FALSE, 1, 0)  # logicals become numeric
c(1L, 2.5)            # integer becomes double
```

The coercion hierarchy: logical → integer → double → character.

Explicit coercion with `as.*()`:

```{r}
as.numeric("42")
as.character(100)
as.logical(0)
as.integer(3.9)  # truncates, does not round
```

### 2.7 Missing Values

`NA` represents missing data:

```{r}
x <- c(10, 20, NA, 40, 50)
mean(x)            # NA propagates
mean(x, na.rm = TRUE)  # remove NAs first
is.na(x)           # check which are NA
sum(is.na(x))      # count NAs
```

NA propagation is a safety feature. It forces you to make a conscious choice about missing data.

### Your Turn: Exercise 2.1

Create a vector of the first 10 even numbers (2, 4, 6, ..., 20) using three different methods:

1. Using `c()` with all values typed out
2. Using the `seq()` function
3. Using the `:` operator with multiplication

```{r}
x <- c(2,4,6,8,10,12,14,16,18,20)
y <-  seq(2,20, by= 2)
z <- (1:10)*2

```
```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
# Method 1
c(2, 4, 6, 8, 10, 12, 14, 16, 18, 20)

# Method 2
seq(2, 20, by = 2)

# Method 3
(1:10) * 2
```

### Your Turn: Exercise 2.2

Given this temperature data:

```{r}
temps_f <- c(72, 68, 75, 80, NA, 78, 82)
```

1. Convert all temperatures to Celsius: C = (F - 32) * 5/9
2. Find the mean temperature in Celsius (handling the NA)
3. Count how many days were above 25 degrees Celsius

```{r}
(temps_f -32)*5/9
mean(temps_f, na.rm = TRUE)
sum(temps_f > 25, na.rm = TRUE)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
temps_c <- (temps_f - 32) * 5/9
temps_c

mean(temps_c, na.rm = TRUE)

sum(temps_c > 25, na.rm = TRUE)
```

### Your Turn: Exercise 2.3

```{r}
scores <- c(88, 92, 75, 95, 60, 84, 91, 77)
```

1. Find the highest and lowest scores
2. How many scores are above 80?
3. What proportion of scores are above 80?
4. Sort the scores in descending order

```{r}
max(scores)
min(scores)
sum(scores>80)/length(scores)*100
sort(scores, decreasing = TRUE)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
max(scores)
min(scores)

sum(scores > 80)
mean(scores > 80)

sort(scores, decreasing = TRUE)
```

::: {.callout-warning title="Common Mistake"}
**Using `==` to check for NA.**

```{r}
x <- c(1, NA, 3)
x == NA   # Does NOT work
is.na(x)  # Correct way
```

`NA == NA` returns `NA`, not `TRUE`. Always use `is.na()` to check for missing values.
:::

::: {.callout-tip title="Challenge"}
Create a named vector of five countries and their populations (in millions). Then:

1. Extract the population of the country with the highest population using logical indexing
2. Find which countries have populations above the mean
3. Calculate the total population

```{r}
countries <- c(London = 10, Paris = 20, Madrid = 5, Budapest = 2, NewYork = 25)
countries[countries == max(countries)]
countries[countries > mean(countries)]
sum(countries)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
pops <- c(USA = 331, India = 1408, Brazil = 214, Japan = 125, Germany = 83)

pops[pops == max(pops)]
pops[pops > mean(pops)]
sum(pops)
```
:::

---

## Part 3: Lists, Matrices, and Data Frames

### 3.1 Matrices

A matrix is a 2D structure with a single data type:

```{r}
m <- matrix(1:12, nrow = 3, ncol = 4)
m
```

By default, `matrix()` fills by column. Use `byrow = TRUE` to fill by row:

```{r}
m2 <- matrix(1:12, nrow = 3, ncol = 4, byrow = TRUE)
m2
```

### 3.2 Matrix Indexing

```{r}
m <- matrix(1:12, nrow = 3, ncol = 4)
m[1, 2]      # row 1, column 2
m[1, ]       # entire row 1
m[, 3]       # entire column 3
m[1:2, 3:4]  # submatrix
```

The syntax is `[row, col]`. Leave one blank to get the entire row or column.

### 3.3 Matrix Operations

```{r}
m <- matrix(1:4, nrow = 2)
m * 2         # element-wise multiplication
m %*% m       # matrix multiplication
t(m)          # transpose
colSums(m)
rowMeans(m)
```

`*` is element-wise. `%*%` is matrix multiplication. Do not confuse them.

### 3.4 Lists

Lists hold elements of any type and any length:

```{r}
student <- list(
  name = "Alice",
  age = 21,
  scores = c(90, 85, 92, 88),
  enrolled = TRUE
)
str(student)
```

### 3.5 Extracting from Lists

This is critical. Pay attention to the difference:

```{r}
student$name           # returns the character string
student[["age"]]       # returns the number
student["scores"]      # returns a LIST containing scores
student[[3]]           # returns the vector of scores
```

```{r}
class(student["scores"])   # list
class(student[["scores"]]) # numeric
```

`$` and `[[` extract the element itself. `[` returns a sub-list.

### 3.6 Data Frames

A data frame is a list of equal-length vectors. Each column can have a different type.

```{r}
students <- data.frame(
  name = c("Alice", "Bob", "Carlos", "Diana", "Elena", "Frank"),
  age = c(21, 23, 22, 20, 24, 21),
  major = c("CS", "Math", "CS", "Stats", "CS", "Math"),
  gpa = c(3.8, 3.2, 3.5, 3.9, 3.1, 3.6),
  year = c(3, 4, 3, 2, 4, 3)
)
students
```

We will use this `students` data frame throughout Sessions 1-4.

### 3.7 Exploring Data Frames

```{r}
str(students)
summary(students)
head(students, 3)
nrow(students)
ncol(students)
names(students)
```

`str()` is the single most useful function for understanding any R object.

### 3.8 Accessing Data Frame Elements

Columns:

```{r}
students$name
students[["gpa"]]
```

Rows:

```{r}
students[1, ]       # first row
students[1:3, ]     # first three rows
```

Filtering:

```{r}
students[students$major == "CS", ]
students[students$gpa > 3.5, ]
```

Selecting specific columns:

```{r}
students[students$gpa > 3.5, c("name", "gpa")]
```

### 3.9 Modifying Data Frames

```{r}
# Add a new column
students$honors <- students$gpa >= 3.5
students

# Modify an existing column
students$age <- students$age + 1
head(students, 3)
```

### Your Turn: Exercise 3.1

Create a 3x3 matrix containing the numbers 1 through 9. Then:

1. Extract the element in row 2, column 3
2. Extract the entire second row
3. Calculate the column sums
4. Multiply every element by 10

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
m <- matrix(1:9, nrow = 3)
m[2, 3]
m[2, ]
colSums(m)
m * 10
```

### Your Turn: Exercise 3.2

Create a list called `course` with the following elements:

- `title`: "R Programming"
- `sessions`: 15
- `topics`: a vector of "Base R", "Tidyverse", "Quarto"
- `active`: TRUE

Then:

1. Extract the title using `$`
2. Extract the number of sessions using `[[`
3. Get the second topic from the topics vector

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
course <- list(
  title = "R Programming",
  sessions = 15,
  topics = c("Base R", "Tidyverse", "Quarto"),
  active = TRUE
)

course$title
course[["sessions"]]
course$topics[2]
```

### Your Turn: Exercise 3.3

Using the `students` data frame:

1. Find all students with GPA above 3.5
2. Find all CS majors in year 3
3. What is the average GPA of Math majors?
4. Add a column called `status` that is "senior" if year is 4, otherwise "underclass"

```{r}
students <- data.frame(
  name = c("Alice", "Bob", "Carlos", "Diana", "Elena", "Frank"),
  age = c(21, 23, 22, 20, 24, 21),
  major = c("CS", "Math", "CS", "Stats", "CS", "Math"),
  gpa = c(3.8, 3.2, 3.5, 3.9, 3.1, 3.6),
  year = c(3, 4, 3, 2, 4, 3)
)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
# 1
students[students$gpa > 3.5, ]

# 2
students[students$major == "CS" & students$year == 3, ]

# 3
mean(students[students$major == "Math", "gpa"])

# 4
students$status <- ifelse(students$year == 4, "senior", "underclass")
students
```

::: {.callout-warning title="Common Mistake"}
**Using `[` instead of `[[` to extract a list element, then trying to use it.**

```{r}
#| error: true
student <- list(name = "Alice", gpa = 3.8)
student["gpa"] + 0.1  # ERROR: + doesn't work on lists
```

Fix: Use `[[` or `$`:

```{r}
student[["gpa"]] + 0.1
```
:::

::: {.callout-tip title="Challenge"}
Create a data frame called `products` with 5 products, each having a name, price, and quantity. Then:

1. Add a `total_value` column (price * quantity)
2. Find the product with the highest total value
3. Calculate the grand total of all inventory value

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
products <- data.frame(
  name = c("Laptop", "Mouse", "Keyboard", "Monitor", "Headset"),
  price = c(999, 25, 75, 400, 150),
  quantity = c(10, 200, 80, 30, 60)
)

products$total_value <- products$price * products$quantity
products[products$total_value == max(products$total_value), ]
sum(products$total_value)
```
:::

---

## Session Synthesis

### Connect the Dots

This exercise combines vectors, lists, and data frames.

You are tracking exam results for a class. Build the analysis step by step:

1. Create a vector of 20 random exam scores between 50 and 100
2. Calculate the mean, median, and standard deviation
3. Create a named list containing: the scores vector, the mean, the number of passing scores (>= 60), and a logical value for whether the class average is above 70
4. Create a data frame with columns: `student_id` (1:20), `score` (your scores), `passed` (TRUE/FALSE), and `grade` (A/B/C/F based on score ranges)
5. Find the number of students in each grade category

```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false
# 1
set.seed(123)
scores <- sample(50:100, 20, replace = TRUE)

# 2
mean(scores)
median(scores)
sd(scores)

# 3
results <- list(
  scores = scores,
  average = mean(scores),
  passing_count = sum(scores >= 60),
  class_passed = mean(scores) > 70
)
str(results)

# 4
exam_df <- data.frame(
  student_id = 1:20,
  score = scores,
  passed = scores >= 60,
  grade = ifelse(scores >= 90, "A",
          ifelse(scores >= 80, "B",
          ifelse(scores >= 70, "C", "F")))
)
head(exam_df)

# 5
table(exam_df$grade)
```

### Key Takeaways

1. Everything in R is a vector. A single number is a vector of length 1.
2. Vectorized operations replace loops. `x * 2` multiplies every element — no loop needed.
3. `[` keeps the container type. `[[` and `$` extract the contents. This matters most for lists.
4. Data frames are lists of equal-length columns. They are the primary structure for data analysis in R.
5. `str()` and `summary()` are your first tools for understanding any R object.

### Prepare for Next Session

**Session 2: Working with Data Structures**

We will cover:

- Advanced indexing and subsetting
- Control flow: if/else, for, while
- Writing your own functions

To prepare: make sure you can create the `students` data frame from memory and filter it by column values. We will build on it heavily.
