---
title: "Session 7: Specialized Data Types"
subtitle: "R Programming Workbook"
author: "R Programming Course"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-tools: true
    code-overflow: wrap
    theme: cosmo
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-note title="Learning Objectives"}
After completing this session, you will be able to:

1. Create and manipulate factors using forcats functions
2. Reorder factor levels for correct plotting and display
3. Parse dates from strings using lubridate
4. Extract date components and perform date arithmetic
5. Use map() and its typed variants to iterate over lists
6. Handle errors in iteration with safely() and possibly()
7. Work with nested data frames using nest() and map()
8. Combine factor, date, and list operations in data pipelines
:::

```{r}
#| label: setup
library(tidyverse)

set.seed(42)
sales <- tibble(
  date = rep(seq(as.Date("2024-01-01"), as.Date("2024-12-31"), by = "day"), 3),
  region = rep(c("North", "South", "West"), each = 366),
  product = sample(c("Widget", "Gadget", "Doohickey"), 366 * 3, replace = TRUE),
  units = rpois(366 * 3, lambda = 50),
  revenue = round(runif(366 * 3, 10, 500), 2),
  returned = rbinom(366 * 3, 1, 0.05)
)
```

## Part 1: Factors with forcats

### Creating Factors

```{r}
# Character vector — alphabetical order
sizes <- c("Large", "Small", "Medium", "Large", "Small")
sort(sizes)
```

```{r}
# Factor with meaningful order
sizes_f <- factor(sizes, levels = c("Small", "Medium", "Large"))
sort(sizes_f)
```

### Key forcats Functions

```{r}
library(forcats)

# fct_infreq: order by frequency
sales |>
  mutate(product = fct_infreq(product)) |>
  count(product)
```

```{r}
# fct_reorder: order by another variable
sales |>
  summarize(avg_rev = mean(revenue), .by = product) |>
  mutate(product = fct_reorder(product, avg_rev))
```

```{r}
# fct_lump_n: keep top n, collapse rest to "Other"
tibble(letter = sample(letters[1:10], 200, replace = TRUE,
                       prob = c(.2, .15, .12, .1, .1, .08, .08, .07, .05, .05))) |>
  mutate(letter = fct_lump_n(letter, n = 5)) |>
  count(letter, sort = TRUE)
```

```{r}
# fct_recode: rename levels
sales |>
  mutate(region = fct_recode(region,
    "Northern" = "North",
    "Southern" = "South",
    "Western" = "West"
  )) |>
  count(region)
```

::: {.callout-note title="Base R vs Tidyverse"}
**Base R:** `factor(x, levels = ...)` and `levels(x) <- ...` for direct manipulation.

**forcats:** Consistent `fct_*` functions that always return a factor and never silently drop levels.
:::

### Your Turn: Factors

**Exercise 1.1:** Create a factor from the vector `c("Freshman", "Junior", "Senior", "Sophomore", "Junior")` with levels in the correct academic order.

```{r}
levels <- factor(c("Freshman", "Junior", "Senior", "Sophomore", "Junior"),levels=c("Freshman", "Sophomore", "Junior", "Senior"))
sort(levels)

```

```{r}
#| code-fold: true
#| code-summary: "Solution"
year <- factor(
  c("Freshman", "Junior", "Senior", "Sophomore", "Junior"),
  levels = c("Freshman", "Sophomore", "Junior", "Senior")
)
sort(year)
```

**Exercise 1.2:** Using the `sales` dataset, reorder `region` by total revenue (highest first) and show the result.

```{r}
sales |> mutate(region = fct_reorder(region,revenue, .fun=sum, .desc=TRUE)) |> summarise(total_rev = sum(revenue),.by=region) 
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  mutate(region = fct_reorder(region, revenue, .fun = sum, .desc = TRUE)) |>
  summarize(total = sum(revenue), .by = region)
```

**Exercise 1.3:** Collapse the product column so that "Gadget" and "Doohickey" become "Non-Widget" and count the results.

```{r}
sales |> mutate(product = fct_collapse(product,
Widget = "Widget", 
`Non-Widget` = c("Gadget","Doohickey")
)) |> count(product)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  mutate(product = fct_collapse(product,
    Widget = "Widget",
    `Non-Widget` = c("Gadget", "Doohickey")
  )) |>
  count(product)
```

::: {.callout-warning title="Common Mistake"}
Assigning a value that isn't in the factor levels:

```{r}
f <- factor(c("A", "B", "C"), levels = c("A", "B", "C"))
f[1] <- "D"
f  # First element is now NA
```

Fix: Add the new level first with `fct_expand()`:

```{r}
f <- factor(c("A", "B", "C"), levels = c("A", "B", "C"))
f <- fct_expand(f, "D")
f[1] <- "D"
f
```
:::

::: {.callout-tip title="Challenge"}
From `sales`, create a factor `quarter` from the date column, ordered Q1 through Q4. Then compute average revenue per quarter per region, with regions ordered by their Q1 revenue.

```{r}
#| code-fold: true
#| code-summary: "Solution"
q1_order <- sales |>
  mutate(quarter = paste0("Q", quarter(date))) |>
  filter(quarter == "Q1") |>
  summarize(q1_rev = mean(revenue), .by = region) |>
  arrange(desc(q1_rev)) |>
  pull(region)

sales |>
  mutate(
    quarter = factor(paste0("Q", quarter(date)), levels = paste0("Q", 1:4)),
    region = factor(region, levels = q1_order)
  ) |>
  summarize(avg_revenue = round(mean(revenue), 2), .by = c(region, quarter)) |>
  arrange(region, quarter)
```
:::

## Part 2: Dates with lubridate

### Parsing Dates

```{r}
library(lubridate)

# Function name matches input order
ymd("2024-03-15")
mdy("March 15, 2024")
dmy("15/03/2024")
ymd_hms("2024-03-15 14:30:00")
```

### Extracting Components

```{r}
d <- ymd("2024-07-15")

year(d)
month(d)
month(d, label = TRUE)
day(d)
wday(d, label = TRUE)
quarter(d)
yday(d)
```

### Date Arithmetic

```{r}
d <- ymd("2024-03-15")

d + days(30)
d + months(2)
d - years(1)
```

```{r}
# Difference between dates
as.numeric(ymd("2024-12-31") - ymd("2024-01-01"))
```

### Rounding Dates

```{r}
dates <- ymd(c("2024-03-15", "2024-03-28", "2024-04-02"))

floor_date(dates, "month")    # start of month
ceiling_date(dates, "month")  # start of next month
floor_date(dates, "week")     # start of week
```

### Your Turn: Dates

**Exercise 2.1:** Parse these date strings into proper Date objects:

```{r}
date_strings <- c("01/15/2024", "02/28/2024", "12/25/2023")
```

```{r}

mdy(date_strings)
```
```{r}
#| code-fold: true
#| code-summary: "Solution"
mdy(date_strings)
```

**Exercise 2.2:** From the `sales` dataset, add columns for month name, day of week, and week number. Then find which day of the week has the highest average units sold.

```{r}
sales |> mutate(
  month_name = month(date,label=TRUE),
  day_of_week = wday(date, label=TRUE),
  week_number = week(date)
)|> summarise(average_unit_sold = mean(revenue), .by = day_of_week) |> arrange(desc(average_unit_sold))
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  mutate(
    month_name = month(date, label = TRUE),
    weekday = wday(date, label = TRUE),
    week_num = week(date)
  ) |>
  summarize(avg_units = mean(units), .by = weekday) |>
  arrange(desc(avg_units))
```

**Exercise 2.3:** Filter `sales` to the last 90 days of 2024 (October 2 through December 31).

```{r}
sales |> filter(date >= ymd("2024-10-02"))|> filter(date <= ymd("2024-12-31")) |>summarize(
    min_date = min(date),
    max_date = max(date),
    n_rows = n()
  )
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
cutoff <- ymd("2024-12-31") - days(90)
cutoff

sales |>
  filter(date >= cutoff) |>
  summarize(
    min_date = min(date),
    max_date = max(date),
    n_rows = n()
  )
```

**Exercise 2.4:** Group `sales` by week (using `floor_date()`) and compute weekly revenue totals per region.

```{r}
sales |> mutate(week = floor_date(date,"week")) |> summarise(total_rev = sum(revenue),.by=c(week,region))
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  mutate(week = floor_date(date, "week")) |>
  summarize(weekly_revenue = sum(revenue), .by = c(region, week)) |>
  arrange(region, week) |>
  head(10)
```

::: {.callout-warning title="Common Mistake"}
Month arithmetic near month boundaries:

```{r}
# January 31 + 1 month = ?
ymd("2024-01-31") + months(1)
```

This returns February 29 (2024 is a leap year). In non-leap years, it returns NA:

```{r}
ymd("2023-01-31") + months(1)
```

Fix: Use `%m+%` for rollback behavior:

```{r}
ymd("2023-01-31") %m+% months(1)  # rolls to Feb 28
```
:::

::: {.callout-tip title="Challenge"}
Compute the number of business days (Monday through Friday) in each month of 2024 for each region's sales data. Then find which month-region combination had the highest revenue per business day.

```{r}
business_days = c("Mon","Tue","Wed","Thu","Fri")
sales |> 
  mutate(
    week_day = wday(date,label=TRUE)%in% business_days, 
    month = month(date,label = TRUE)) |> 
  filter (week_day) |> 
  summarise(total_rev = sum(revenue), .by=c(region,month))
 ```

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  mutate(
    month = floor_date(date, "month"),
    is_weekday = !wday(date) %in% c(1, 7)
  ) |>
  filter(is_weekday) |>
  summarize(
    total_revenue = sum(revenue),
    n_days = n_distinct(date),
    rev_per_day = total_revenue / n_days,
    .by = c(region, month)
  ) |>
  slice_max(rev_per_day, n = 5)
```
:::

## Part 3: Functional Programming with purrr

### map() Basics

```{r}
library(purrr)

# map always returns a list
list(1, 4, 9, 16) |> map(sqrt)
```

```{r}
# Typed variants return vectors
list(1, 4, 9, 16) |> map_dbl(sqrt)
list("hello", "world") |> map_int(nchar)
c(1, -2, 3) |> map_lgl(\(x) x > 0)
```

### map() with Lambda

```{r}
# Custom operations
1:5 |> map_dbl(\(x) x^2 + 1)

# Multiple operations per element
list(c(1, 2, 3), c(10, 20), c(100)) |>
  map(\(x) tibble(mean = mean(x), n = length(x)))
```

### map2() and pmap()

```{r}
# Two parallel inputs
map2_chr(
  c("Alice", "Bob", "Carlos"),
  c(90, 85, 92),
  \(name, score) str_glue("{name}: {score}")
)
```

```{r}
# Multiple inputs from a data frame
params <- tibble(
  n = c(100, 200, 50),
  mean = c(0, 10, 5),
  sd = c(1, 2, 0.5)
)

pmap_dbl(params, \(n, mean, sd) mean(rnorm(n, mean, sd)))
```

### Error Handling

```{r}
# possibly: return default on error
safe_parse <- possibly(as.numeric, otherwise = NA_real_)
c("1", "two", "3", "four") |> map_dbl(safe_parse)
```

```{r}
# safely: capture both result and error
safe_log <- safely(log)
results <- list(10, -1, "oops") |> map(safe_log)
results |> map("result")
results |> map("error")
```

### Nested Data and map()

```{r}
# Nest by region
nested <- sales |> nest(.by = region)
nested
```

```{r}
# Compute per group using map
nested |>
  mutate(
    n_rows = map_int(data, nrow),
    avg_revenue = map_dbl(data, \(df) mean(df$revenue)),
    top_product = map_chr(data, \(df) {
      df |> count(product, sort = TRUE) |> slice(1) |> pull(product)
    })
  ) |>
  select(-data)
```

::: {.callout-note title="Base R vs Tidyverse"}
**Base R:** `lapply()` returns list, `sapply()` guesses output type (dangerous), `vapply()` requires type spec.

**purrr:** `map()` returns list, `map_dbl()/map_chr()/map_int()/map_lgl()` enforce return types. Safer and more consistent.
:::

### Your Turn: purrr

**Exercise 3.1:** Use `map_dbl()` to compute the standard deviation of each numeric column in `sales`.

```{r}
sales |> select(where(is.numeric)) |> map_dbl(sd)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  select(where(is.numeric)) |>
  map_dbl(sd)
```

**Exercise 3.2:** Use `map2_dbl()` to compute weighted scores. Students have scores and weights:


```{r}
scores <- list(c(85, 90, 78), c(92, 88, 95), c(70, 75, 80))
weights <- list(c(0.3, 0.4, 0.3), c(0.2, 0.3, 0.5), c(0.5, 0.3, 0.2))
```

```{r}
map2_dbl(scores, weights, weighted.mean)
```
```{r}
#| code-fold: true
#| code-summary: "Solution"
map2_dbl(scores, weights, weighted.mean)
```

**Exercise 3.3:** Nest `sales` by product, fit a linear model `revenue ~ units` for each, and extract the slope (coefficient for units).

```{r}
sales |> nest(.by= product) |> mutate(
  model = map(data, \(x) lm(revenue~units,data = x)),
  slope = map_dbl(model, \(m) coef(m)[["units"]])
)
```
```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  nest(.by = product) |>
  mutate(
    model = map(data, \(df) lm(revenue ~ units, data = df)),
    slope = map_dbl(model, \(m) coef(m)[["units"]])
  ) |>
  select(product, slope)
```

**Exercise 3.4:** Use `possibly()` to safely parse these mixed values to numeric, returning NA for failures:

```{r}
messy_values <- c("42", "3.14", "N/A", "100", "unknown", "7.5")
```

```{r}
messy_values |> map_dbl(possibly(as.numeric, otherwise=NA))
```
```{r}
#| code-fold: true
#| code-summary: "Solution"
safe_numeric <- possibly(as.numeric, otherwise = NA_real_)
messy_values |> map_dbl(safe_numeric)
```

**Exercise 3.5:** Use `walk()` to print a summary sentence for each region's total revenue:

```{r}
sales |> summarise(total_rev = sum(revenue),.by=region) |> pmap(list)|> walk(\(x) cat(str_glue("{x$region} : {x$total_rev}"),"\n"))
```
```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  summarize(total = sum(revenue), .by = region) |>
  pmap(list) |>
  walk(\(x) cat(str_glue("{x$region}: ${round(x$total, 0)}"), "\n"))
```

::: {.callout-warning title="Common Mistake"}
Using `map_dbl()` when the function returns something other than a single double:

```{r}
#| error: true
# This returns a vector, not a scalar
list(c(1, 2), c(3, 4)) |> map_dbl(range)
```

Fix: Use `map()` for complex return values:

```{r}
list(c(1, 2), c(3, 4)) |> map(range)
```
:::

**Think About It:** When is `nest() + map()` better than `.by` in `summarize()`? What kinds of operations require nesting that grouped summarize cannot handle?

::: {.callout-tip title="Challenge"}
Build a nested analysis: for each region-product combination, compute a rolling 7-day average of revenue. Use nest, map, and a custom function.

```{r}
#| code-fold: true
#| code-summary: "Solution"
rolling_avg <- function(df, window = 7) {
  df |>
    arrange(date) |>
    mutate(
      rolling_7d = map_dbl(
        seq_len(n()),
        \(i) mean(revenue[max(1, i - window + 1):i])
      )
    )
}

sales |>
  nest(.by = c(region, product)) |>
  mutate(data = map(data, rolling_avg)) |>
  unnest(data) |>
  select(date, region, product, revenue, rolling_7d) |>
  head(10)
```
:::

## Session Synthesis

### Connect the Dots

Build a comprehensive time-based analysis combining all three skills:

1. Parse dates and extract month, quarter, and day of week from `sales`
2. Create ordered factor for day of week (Monday first)
3. Nest by region and compute monthly summaries using map
4. Unnest and identify the best-performing month per region

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  mutate(
    month = floor_date(date, "month"),
    weekday = fct_relevel(
      wday(date, label = TRUE),
      "Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"
    ),
    quarter = paste0("Q", quarter(date))
  ) |>
  nest(.by = region) |>
  mutate(
    monthly = map(data, \(df) {
      df |>
        summarize(
          total_revenue = sum(revenue),
          avg_units = mean(units),
          n_returns = sum(returned),
          .by = month
        )
    })
  ) |>
  select(-data) |>
  unnest(monthly) |>
  mutate(month_label = month(month, label = TRUE)) |>
  slice_max(total_revenue, n = 1, by = region) |>
  select(region, month_label, total_revenue, avg_units, n_returns)
```

### Key Takeaways

1. forcats gives you control over factor ordering. Use `fct_reorder()` for data-driven ordering and `fct_lump_n()` to simplify many-level factors.
2. lubridate parses dates with `ymd()`/`mdy()`/`dmy()` and extracts parts with `year()`/`month()`/`wday()`. Use `floor_date()` for grouping by time periods.
3. purrr's `map()` family provides type-safe iteration. Use `nest() + map()` for per-group operations that go beyond simple summarize.
4. `safely()` and `possibly()` let you handle errors gracefully during iteration.

### Prepare for Next Session

**Session 8: Data Visualization with ggplot2**

You will learn:

- Grammar of graphics concepts
- Building plots layer by layer
- Facets, scales, themes, and customization

The factor ordering skills from today directly improve your ggplot2 output. Run `library(ggplot2)` to confirm it works.
