---
title: "Session 5: Enter the Tidyverse"
subtitle: "R Programming Workbook"
author: "R Programming Course"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-tools: true
    code-overflow: wrap
    theme: cosmo
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-note title="Learning Objectives"}
After completing this session, you will be able to:

1. Explain the design principles behind the tidyverse
2. Write pipe chains using the native |> operator
3. Create and work with tibbles, understanding how they differ from data frames
4. Use filter(), select(), mutate(), summarize(), and arrange() to manipulate data
5. Apply grouped operations using the .by argument
6. Use across() to apply functions to multiple columns
7. Combine datasets using different join types
8. Build multi-step data pipelines that chain verbs together
:::

```{r}
#| label: setup
library(tidyverse)

# Recurring datasets
students_tbl <- tibble(
  name = c("Alice", "Bob", "Carlos", "Diana", "Elena", "Frank"),
  age = c(21, 23, 22, 20, 24, 21),
  major = c("CS", "Math", "CS", "Stats", "CS", "Math"),
  gpa = c(3.8, 3.2, 3.5, 3.9, 3.1, 3.6),
  year = c(3, 4, 3, 2, 4, 3)
)

set.seed(42)
sales <- tibble(
  date = rep(seq(as.Date("2024-01-01"), as.Date("2024-12-31"), by = "day"), 3),
  region = rep(c("North", "South", "West"), each = 366),
  product = sample(c("Widget", "Gadget", "Doohickey"), 366 * 3, replace = TRUE),
  units = rpois(366 * 3, lambda = 50),
  revenue = round(runif(366 * 3, 10, 500), 2),
  returned = rbinom(366 * 3, 1, 0.05)
)
```

## Part 1: Pipes and Tibbles

### The Native Pipe |>

The pipe passes the left side as the first argument to the right side:

```{r}
# Without pipe
round(mean(c(10, 20, NA, 40), na.rm = TRUE), 1)

# With pipe — read left to right
c(10, 20, NA, 40) |> mean(na.rm = TRUE) |> round(1)
```

Both produce the same result. The pipe version reads as a sequence of steps.

When the input needs to go somewhere other than the first argument, use `_` as a placeholder:

```{r}
mtcars |> lm(mpg ~ wt, data = _) |> coef()
```

For more complex cases, use a lambda:

```{r}
c(3, 1, 4, 1, 5) |> (\(x) x[x > 2])()
```

### Tibbles

Tibbles are modern data frames with stricter behavior:

```{r}
# Print difference
students_tbl
```

Key differences:

```{r}
# Data frame allows partial matching
students_df <- as.data.frame(students_tbl)
students_df$na  # matches "name" — dangerous

# Tibble does not
students_tbl$na  # returns NULL with a warning
```

```{r}
# Subsetting difference
class(students_df[, "name"])    # vector
class(students_tbl[, "name"])   # tibble
```

::: {.callout-note title="Base R vs Tidyverse"}
**Data frames** allow partial column matching and silently drop dimensions. **Tibbles** prevent both, catching bugs early. You can convert freely with `as_tibble()` and `as.data.frame()`.
:::

### Your Turn: Pipe Practice

**Exercise 1.1:** Rewrite this nested code using pipes:

```{r}
#| eval: false
sort(unique(tolower(c("Apple", "BANANA", "apple", "Cherry", "banana"))))

c("Apple", "BANANA", "apple", "Cherry", "banana") |> tolower() |> unique() |> sort()
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
c("Apple", "BANANA", "apple", "Cherry", "banana") |>
  tolower() |>
  unique() |>
  sort()
```

**Exercise 1.2:** Use a pipe chain to: take the vector `c(4, 9, 16, 25, 36)`, compute the square root of each, then calculate the mean.

```{r}
#| code-fold: true
#| code-summary: "Solution"
c(4, 9, 16, 25, 36) |> sqrt() |> mean()
```

**Exercise 1.3:** Create a tibble called `courses` with three columns: `code` (character), `credits` (numeric), and `required` (logical). Include at least 4 rows.

```{r}
#| code-fold: true
#| code-summary: "Solution"
courses <- tibble(
  code = c("CS101", "MATH201", "STAT301", "CS401"),
  credits = c(3, 4, 3, 3),
  required = c(TRUE, TRUE, FALSE, TRUE)
)
courses
```

::: {.callout-warning title="Common Mistake"}
Using `%>%` instead of `|>`:

```{r}
#| eval: false
# Wrong (magrittr pipe — not used in this course)
x %>% mean()

# Right (native pipe)
x |> mean()
```

Both work if magrittr is loaded, but we use the native pipe exclusively. It requires no package and works everywhere in R 4.1+.
:::

### Your Turn: Tibble Exploration

**Exercise 1.4:** Convert `mtcars` to a tibble, move the row names into a column called `car`, and print it.

```{r}
#| code-fold: true
#| code-summary: "Solution"
mtcars |>
  rownames_to_column("car") |>
  as_tibble()
```

::: {.callout-tip title="Challenge"}
Write a single pipe chain that: takes `sales`, extracts the `revenue` column as a vector, computes the median, rounds to 2 decimal places, and wraps it in `paste0("Median revenue: $", ...)`.

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  pull(revenue) |>
  median() |>
  round(2) |>
  (\(x) paste0("Median revenue: $", x))()
```
:::

## Part 2: Core dplyr Verbs

### filter(): Keep Rows

```{r}
# Single condition
sales |> filter(region == "North") |> head(3)
```

```{r}
# Multiple conditions (AND)
sales |> filter(region == "North", units > 60) |> head(3)
```

```{r}
# OR condition
sales |> filter(region == "North" | product == "Gadget") |> head(3)
```

```{r}
# Set membership
sales |> filter(product %in% c("Widget", "Gadget")) |> head(3)
```

### select(): Choose Columns

```{r}
# By name
sales |> select(date, region, revenue) |> head(3)
```

```{r}
# Drop columns
sales |> select(-returned, -units) |> head(3)
```

```{r}
# Helper functions
sales |> select(starts_with("r")) |> head(3)
```

```{r}
# By type
sales |> select(where(is.numeric)) |> head(3)
```

### mutate(): Create Columns

```{r}
sales |>
  mutate(
    profit = revenue * 0.3,
    high_volume = units > 60
  ) |>
  head(5)
```

New columns can reference columns created in the same `mutate()`:

```{r}
sales |>
  mutate(
    revenue_per_unit = revenue / units,
    rpu_category = if_else(revenue_per_unit > 5, "high", "low")
  ) |>
  head(5)
```

### summarize(): Aggregate

```{r}
sales |>
  summarize(
    total_rev = sum(revenue),
    mean_units = mean(units),
    n_rows = n(),
    n_products = n_distinct(product)
  )
```

### arrange(): Sort

```{r}
# Ascending (default)
sales |> arrange(revenue) |> head(5)

# Descending
sales |> arrange(desc(revenue)) |> head(5)

# Multiple columns
sales |> arrange(region, desc(revenue)) |> head(5)
```

::: {.callout-note title="Base R vs Tidyverse"}
**Base R:** `students[students$gpa > 3.5, c("name", "gpa")]`

**dplyr:** `students_tbl |> filter(gpa > 3.5) |> select(name, gpa)`

Both work. dplyr reads more clearly when operations grow complex.
:::

### Your Turn: Core Verbs

**Exercise 2.1:** Filter `sales` to show only rows where `returned` is 1 (returned items) in the "West" region.

```{r}
sales |> filter(returned == 1, region == "West")
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  filter(returned == 1, region == "West")
```

**Exercise 2.2:** Select only `date`, `product`, and `units` from `sales`, then arrange by `units` descending. Show the top 10.

```{r}
sales |> select(c("date", "product", "units")) |> arrange(desc(units)) |> head(10)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  select(date, product, units) |>
  arrange(desc(units)) |>
  head(10)
```

**Exercise 2.3:** Add a column `revenue_category` to `sales` that is "high" when revenue > 300, "medium" when revenue is between 100 and 300, and "low" otherwise. Then count how many rows fall into each category.

```{r}
sales |> mutate(
  revenue_category = case_when(
    revenue>300 ~ "high",
    revenue>= 100 ~ "medium",
    .default = "low"
  )
) |> count(revenue_category)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  mutate(
    revenue_category = case_when(
      revenue > 300 ~ "high",
      revenue >= 100 ~ "medium",
      .default = "low"
    )
  ) |>
  count(revenue_category)
```

**Exercise 2.4:** Find the average revenue and total units sold for each product type.

```{r}
sales |> summarise(
  average_revenue = mean(revenue),
  total_units = sum(units),
  .by = product
)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  summarize(
    avg_revenue = mean(revenue),
    total_units = sum(units),
    .by = product
  )
```

::: {.callout-warning title="Common Mistake"}
Forgetting that `filter()` uses `==`, not `=`:

```{r}
#| error: true
sales |> filter(region = "North")
```

Fix: Use `==` for comparison.

```{r}
#| eval: false
sales |> filter(region == "North")
```
:::

**Exercise 2.5:** Build a pipeline that: filters sales to January 2024 only, selects date, region, and revenue, adds a column `day_of_week` using `weekdays(date)`, then sorts by revenue descending. Show the top 5.

```{r}
sales
sales |> filter(date >= "2024-01-01", date < "2024-02-01") |> select(date, region, revenue) |> mutate(day_of_week = wday(date, label = TRUE, abbr = FALSE, locale = "hu_HU.UTF-8")) |> arrange(desc(revenue)) |> head(5)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  filter(date >= "2024-01-01", date <= "2024-01-31") |>
  select(date, region, revenue) |>
  mutate(day_of_week = weekdays(date)) |>
  arrange(desc(revenue)) |>
  head(5)
```

::: {.callout-tip title="Challenge"}
Using `students_tbl`, write a pipeline that: adds a column `standing` ("Dean's List" if gpa >= 3.7, "Good" if gpa >= 3.0, "Probation" otherwise), selects name, major, gpa, and standing, and arranges by gpa descending.

```{r}
students_tbl |> mutate(
  standing = case_when(
    gpa >= 3.7 ~ "Dean's list",
    gpa >= 3.0 ~ "Good",
    .default = "Probation"
  )
) |> select(name,major,gpa,standing) |> arrange(desc(gpa))
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
students_tbl |>
  mutate(
    standing = case_when(
      gpa >= 3.7 ~ "Dean's List",
      gpa >= 3.0 ~ "Good",
      .default = "Probation"
    )
  ) |>
  select(name, major, gpa, standing) |>
  arrange(desc(gpa))
```
:::

## Part 3: Advanced dplyr

### Grouped Operations with .by

```{r}
# Summarize by group
sales |>
  summarize(
    avg_revenue = mean(revenue),
    max_units = max(units),
    .by = region
  )
```

```{r}
# Multiple grouping columns
sales |>
  summarize(
    total_revenue = sum(revenue),
    .by = c(region, product)
  )
```

```{r}
# .by works with mutate (adds group-level values without collapsing)
sales |>
  mutate(region_mean = mean(revenue), .by = region) |>
  select(date, region, revenue, region_mean) |>
  head(5)
```

::: {.callout-note title="Base R vs Tidyverse"}
**group_by approach:**
```{r}
#| eval: false
sales |>
  group_by(region) |>
  summarize(avg = mean(revenue)) |>
  ungroup()
```

**.by approach (preferred):**
```{r}
#| eval: false
sales |>
  summarize(avg = mean(revenue), .by = region)
```

Both produce the same result. `.by` never leaves leftover grouping.
:::

### across(): Multiple Columns

```{r}
# Apply one function to multiple columns
sales |>
  summarize(across(c(units, revenue), list(avg=mean,max=max),.names="{.col}({.fn})"), .by = region)
```

```{r}
# Named list of functions
sales |>
  summarize(
    across(c(units, revenue), list(avg = mean, sd = sd)),
    .by = region
  )
```

```{r}
# With lambda and column selection by type
sales |>
  summarize(
    across(where(is.numeric), \(x) round(mean(x), 1)),
    .by = region
  )
```

### Joins

```{r}
# Lookup table
regions <- tibble(
  region = c("North", "South", "West"),
  manager = c("Alice", "Bob", "Carlos"),
  target = c(50000, 45000, 55000)
)
```

```{r}
# left_join: keep all sales rows, add region info
sales |>
  left_join(regions, by = "region") |>
  head(5)
```

```{r}
# semi_join: keep sales rows that have a match (no new columns)
high_targets <- regions |> filter(target >= 50000)

high_targets
regions
sales

sales |>
  semi_join(high_targets, by = "region") |>
  count(region)
```

```{r}
# anti_join: keep sales rows that do NOT match
sales |>
  anti_join(high_targets, by = "region") |>
  count(region)
```

### Your Turn: Advanced dplyr

**Exercise 3.1:** Compute the total revenue and average units per region per product.

```{r}
sales |> summarise(
  total_revenue = sum(revenue),
  average_unit = mean(units),
  .by = c(region, product)
)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  summarize(
    total_revenue = sum(revenue),
    avg_units = mean(units),
    .by = c(region, product)
  )
```

**Exercise 3.2:** For each region, find the date with the highest total revenue (sum across all products on that date).

```{r}
sales |> summarise(daily_total = sum(revenue), .by=c(region,date)) |> slice_max(daily_total, by=region, n=1)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  summarize(daily_total = sum(revenue), .by = c(region, date)) |>
  slice_max(daily_total, n = 1, by = region)
```

**Exercise 3.3:** Use `across()` to compute the minimum and maximum of every numeric column in `sales`, grouped by region.

```{r}
sales |> summarise(across(where(is.numeric), list(min=min,max=max)), .by = region)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  summarize(
    across(where(is.numeric), list(min = min, max = max)),
    .by = region
  )
```

**Exercise 3.4:** Add a column `pct_of_region` that shows each row's revenue as a percentage of its region's total revenue.

```{r}
sales |> mutate(pct_of_region = revenue / sum(revenue)*100,.by=region)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  mutate(
    pct_of_region = revenue / sum(revenue) * 100,
    .by = region
  ) |>
  select(date, region, revenue, pct_of_region) |>
  head(5)
```

**Exercise 3.5:** Join `regions` to `sales`, then compute each region's total revenue as a percentage of its target.

```{r}

sales |> summarise(total_revenue = sum(revenue),.by = region) |> left_join(regions, by="region") |> mutate(
  pct_total = total_revenue / target *100
)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  summarize(total_revenue = sum(revenue), .by = region) |>
  left_join(regions, by = "region") |>
  mutate(pct_of_target = round(total_revenue / target * 100, 1)) |>
  select(region, manager, total_revenue, target, pct_of_target)
```

::: {.callout-warning title="Common Mistake"}
Forgetting `.by` returns ungrouped data. If you use `group_by()` instead, the result stays grouped:

```{r}
# This result is still grouped!
grouped_result <- sales |>
  group_by(region) |>
  summarize(total = sum(revenue))

# Check:
group_vars(grouped_result)
```

With `.by`, the result is always ungrouped:

```{r}
ungrouped_result <- sales |>
  summarize(total = sum(revenue), .by = region)

group_vars(ungrouped_result)
```
:::

**Think About It:** When would you choose `semi_join()` over `filter()` with `%in%`? What advantage does `anti_join()` give you over negated filtering?

::: {.callout-tip title="Challenge"}
Create a complete analysis pipeline:

1. Filter sales to Q1 2024 (January through March)
2. Join with the regions table
3. Compute total revenue and return rate (mean of returned) per region per product
4. Add a column showing each row's revenue as a percentage of the region target
5. Arrange by total revenue descending

```{r}
sales |> 
  filter(date>="2024-01-01", date <"2024-04-01") |> 
  left_join(regions, by="region") |>
  summarise(
    total_rev = sum(revenue),
    return_rate = mean(returned),
    target = first(target),
    .by = c(region,product)
  ) |>
  mutate(
    pct_total = total_rev / target *100
  )  |>
  arrange(desc(total_rev))  
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  filter(date >= "2024-01-01", date <= "2024-03-31") |>
  left_join(regions, by = "region") |>
  summarize(
    total_revenue = sum(revenue),
    return_rate = mean(returned),
    target = first(target),
    .by = c(region, product)
  ) |>
  mutate(pct_of_target = round(total_revenue / target * 100, 1)) |>
  arrange(desc(total_revenue))
```
:::

## Session Synthesis

### Connect the Dots

Build a "Regional Performance Report" pipeline that:

1. Starts with `sales`
2. Creates a `month` column from `date`
3. Computes monthly revenue totals per region
4. Joins with the `regions` table to get targets
5. Calculates a monthly pro-rated target (annual target / 12)
6. Flags months as "on_track" or "behind" based on whether monthly revenue met the pro-rated target
7. Summarizes: for each region, count how many months were on track vs behind

```{r}
sales |> 
  mutate(
    month = month(date, label = TRUE)
  ) |>
  summarise(
    monthly_rev = sum(revenue),
    .by = c(month,region)
  )  |>
  left_join(regions, by = "region")  |>
  mutate(
    pro_rated_target = target / 12,
    .by = c(month,region)
  ) |>
  mutate(
    flag = case_when(
      monthly_rev > pro_rated_target ~ "On track",
      .default = "Behind"
    )
  )  |>
  summarise(
    count = n(),
    .by = c(flag, region)
  )  


```

```{r}
#| code-fold: true
#| code-summary: "Solution"
sales |>
  mutate(month = format(date, "%Y-%m")) |>
  summarize(monthly_revenue = sum(revenue), .by = c(region, month)) |>
  left_join(regions, by = "region") |>
  mutate(
    monthly_target = target / 12,
    status = if_else(monthly_revenue >= monthly_target, "on_track", "behind")
  ) |>
  count(region, manager, status) |>
  arrange(region, status)
```

### Key Takeaways

1. The pipe `|>` passes left side as first argument to right side — read data pipelines top to bottom
2. Five core dplyr verbs handle most data manipulation: filter, select, mutate, summarize, arrange
3. Use `.by` for grouped operations — it never leaves leftover grouping
4. `across()` applies functions to multiple columns at once
5. Joins combine datasets: left_join adds columns, semi_join and anti_join filter rows

### Prepare for Next Session

**Session 6: Reshaping and Importing**

You will learn:

- tidyr for reshaping data (wide to long and back)
- readr, readxl, haven for importing different file formats
- stringr for string manipulation

Run `library(tidyverse)` and confirm it loads without errors.
