---
title: "Session 15: Version Control and Capstone"
subtitle: "R Programming Workbook"
author: "R Programming Course"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-tools: true
    code-overflow: wrap
    theme: cosmo
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-note title="Learning Objectives"}
By the end of this session, you will be able to:

1. Initialize a Git repository and make commits from the command line and Positron
2. Write clear commit messages that describe what changed and why (including conventional commit style)
3. Push code to GitHub and pull changes from collaborators
4. Create and merge branches for feature development using `git switch`
5. Resolve merge conflicts
6. Set up a `.gitignore` file for R projects (including bioinformatics-specific exclusions)
7. Use `git restore` to discard file-level changes safely
8. Automate report rendering with GitHub Actions
9. Structure a complete data analysis project using all course skills
10. Initialize `renv` with Bioconductor support (`bioc = TRUE`) for fully reproducible environments
11. Present your work to peers
:::

# Part 1: Version Control with Git

## Why Git?

Every data project benefits from version control. Git tracks every change you make, lets you undo mistakes, and enables collaboration.

Without Git, you end up with filenames like `analysis_v3_final_FIXED.R`. With Git, you have one file and a complete history.

## Git Setup

Before using Git, configure your identity. Run these commands once in the Positron terminal:

```{bash}
#| eval: false

git config --global user.name "Your Name"
git config --global user.email "your.email@example.com"
```

Verify your settings:

```{bash}
#| eval: false

git config --list
```

## Core Git Workflow

The basic Git cycle has four steps:

1. Edit files in your working directory
2. Stage the changes you want to save (`git add`)
3. Commit the staged changes with a message (`git commit`)
4. Push commits to a remote like GitHub (`git push`)

```{bash}
#| eval: false

# Initialize a new repository
git init

# Check what has changed
git status

# Stage a file
git add analysis.R

# Stage all changes
git add -A

# Commit with a descriptive message
git commit -m "Add initial data cleaning script"
```

## Example: Starting a Project

```{bash}
#| eval: false

# Create project directory
mkdir my-r-project
cd my-r-project
git init

# Create initial files
# (In practice, you would create these in Positron)
touch README.md
touch analysis.qmd

# First commit
git add -A
git commit -m "Initialize project with README and analysis template"
```

### Your Turn: First Repository

Create a new directory called `practice-repo`. Initialize it as a Git repository. Create a file called `hello.R` that prints "Hello, Git!" to the console. Stage and commit the file.

```{bash}
#| eval: false

# Your code here
mkdir practice-repo
cd practice-repo
git init

# Create hello.R (use Positron or echo command)
echo 'cat("Hello, Git!\n")' > hello.R

# Stage and commit
git add hello.R
git commit -m "Add hello script"
```

## Writing Good Commit Messages

A commit message should complete the sentence: "This commit will..."

Good examples:

- "Filter outliers from revenue data using 3 SD threshold"
- "Add monthly trend visualization with ggplot2"
- "Fix NA handling in summary statistics function"
- "Restructure project to use here package for paths"

Bad examples:

- "Update"
- "Fix stuff"
- "WIP"
- "Changes"

### Your Turn: Judge These Messages

For each commit message, decide if it is good or bad. If bad, rewrite it.

1. "Updated code"
2. "Add unit tests for data validation functions"
3. "asdf"
4. "Refactor pivot_longer call to handle multiple value columns"
5. "Fixed the thing"

```{r}
#| code-fold: true
#| eval: false

# Suggested answers:
# 1. Bad -> "Update filter thresholds in cleaning pipeline"
# 2. Good
# 3. Bad -> Describe what you actually changed
# 4. Good
# 5. Bad -> "Fix off-by-one error in date range filter"
```

::: {.callout-caution title="Your Turn — Conventional Commits Practice"}
Given these 5 commit message drafts, rewrite each one in **conventional commit** style with a proper type prefix and scope. Conventional commit format: `type(scope): short description`

Common types: `feat` (new feature), `fix` (bug fix), `docs` (documentation), `chore` (tooling/maintenance), `refactor`, `test`, `style`.

Draft messages to rewrite:

1. `fix typo`
2. `WIP`
3. `did some stuff`
4. `add filter function`
5. `update`
:::

::: {.callout-tip collapse="true" title="Solution"}
Corrected conventional commit messages:

1. `docs(README): fix typo in installation instructions`
   - "fix typo" lacks scope and context. Specify what file and what kind of typo.

2. `feat(analysis): add preliminary PCA step for sample QC`
   - "WIP" is never a valid commit message. Describe the actual state of the work, even if incomplete.

3. `refactor(cleaning): extract sample filtering into helper function`
   - "did some stuff" is too vague. Name the change type and the affected component.

4. `feat(qc): add filter_samples() with min_reads and max_pct_mt thresholds`
   - "add filter function" is acceptable but undersells the detail. Add the scope and name the function.

5. `chore(renv): update renv.lock after adding airway package`
   - "update" has no type, no scope, and no description. Always say what was updated and why.

The pattern is `type(scope): imperative verb + object`. Keep it under 72 characters.
:::

## Viewing History

```{bash}
#| eval: false

# Full log
git log

# Compact one-line view
git log --oneline

# See which files changed in each commit
git log --oneline --stat

# See the diff for a specific commit
git show abc1234
```

## Connecting to GitHub

```{bash}
#| eval: false

# 1. Create a new repo on github.com (do NOT add README)

# 2. Add the remote
git remote add origin https://github.com/username/my-project.git

# 3. Push your local commits
git push -u origin main

# 4. Future pushes
git push
```

### Your Turn: Push to GitHub

1. Create a new repository on GitHub (no README, no .gitignore)
2. Connect your `practice-repo` to the GitHub remote
3. Push your commit
4. Verify the code appears on GitHub

```{bash}
#| eval: false

# Your commands here
cd practice-repo
git remote add origin https://github.com/YOUR_USERNAME/practice-repo.git
git push -u origin main
```

## Branching and Merging

Branches let you develop features without affecting the main codebase.

::: {.callout-note title="Modern Git: switch and restore"}
Modern Git (>= 2.23) splits the overloaded `git checkout` into two clearer commands: `git switch` for branches, `git restore` for files. The old `git checkout` still works, but `switch`/`restore` is what you'll see in modern tutorials and what reduces confusion.

| Old command | New command | What it does |
|---|---|---|
| `git checkout -b <branch>` | `git switch -c <branch>` | Create and switch to a new branch |
| `git checkout <branch>` | `git switch <branch>` | Switch to an existing branch |
| `git checkout -- <file>` | `git restore <file>` | Discard working-directory changes to a file |
| `git checkout HEAD -- <file>` | `git restore --source=HEAD <file>` | Restore a file from a specific commit |
:::

```{bash}
#| eval: false

# Create and switch to a new branch
git switch -c add-visualization

# Work on the feature
echo 'library(ggplot2)' > plot.R
git add plot.R
git commit -m "Add visualization script"

# Switch back to main
git switch main

# Merge the feature branch
git merge add-visualization

# Delete the branch (no longer needed)
git branch -d add-visualization
```

### Your Turn: Feature Branch

1. Create a branch called `add-analysis`
2. Create a file called `analysis.R` with some R code
3. Commit it on the branch
4. Switch back to main
5. Verify `analysis.R` does not exist on main
6. Merge the branch
7. Verify `analysis.R` now exists on main

```{bash}
#| eval: false

# Your code here
git switch -c add-analysis
echo 'summary(mtcars)' > analysis.R
git add analysis.R
git commit -m "Add exploratory analysis of mtcars"

git switch main
ls  # analysis.R should not be here

git merge add-analysis
ls  # analysis.R should now be here

git branch -d add-analysis
```

::: {.callout-caution title="Your Turn — Branch Strategy for Parallel Features"}
You are working on an RNA-seq project. Three tasks need to happen in parallel:

**(a)** Adding a QC report
**(b)** Updating from DESeq2 1.40 to DESeq2 1.42
**(c)** Writing the README

Sketch a branch plan. Answer these questions:

1. Which branch does each task branch off from?
2. In what order should they be merged back, and why?
3. Are there any dependencies between the tasks — i.e., should (b) be merged before (a) tests against it?

Draw an ASCII `git log --graph` sketch of the resulting history.
:::

::: {.callout-tip collapse="true" title="Solution"}
All three branches should branch off from `main` independently, since they touch different parts of the project.

```
* e3f1a2c (main) Merge branch 'feat/readme'
|\
| * b4d9f01 (feat/readme) docs(README): add usage and reproduction steps
|/
* 7c3a881 Merge branch 'feat/qc-report'
|\
| * 2a1d4e5 (feat/qc-report) feat(qc): add per-sample QC report with plotly
|/
* 9f0c3b2 Merge branch 'chore/deseq2-update'
|\
| * 1e8b7d4 (chore/deseq2-update) chore(deps): update DESeq2 to 1.42, renv::snapshot()
|/
* 0a2f6c1 Add initial project structure
```

**Recommended merge order:**

1. Merge `chore/deseq2-update` **first**. The QC report likely runs DESeq2, so the QC branch should be tested against the updated version after the update lands on `main`. Merging the update first prevents a second conflict when the QC branch is eventually merged.
2. Merge `feat/qc-report` second, rebasing onto updated `main` if needed.
3. Merge `feat/readme` last — it is purely documentation and has no code dependencies.

**Key principle:** Merge infrastructure/dependency changes before feature branches that depend on them. Branch off `main` often; rebase onto the latest `main` before merging to minimize conflicts.
:::

## Handling Merge Conflicts

Conflicts happen when two branches change the same line. Git marks the conflict in the file:

```
<<<<<<< HEAD
threshold <- 0.05
=======
threshold <- 0.01
>>>>>>> feature-branch
```

To resolve:

1. Open the file
2. Choose the correct version (or combine them)
3. Remove the `<<<<<<<`, `=======`, and `>>>>>>>` markers
4. Stage and commit

```{bash}
#| eval: false

# After fixing the conflict:
git add analysis.R
git commit -m "Resolve conflict: use 0.01 threshold per team discussion"
```

::: {.callout-warning title="Common Mistake"}
Committing a file that still contains conflict markers (`<<<<<<<`, `=======`, `>>>>>>>`). Your code will not run. Always search for these markers before committing a merge resolution.
:::

::: {.callout-caution title="Your Turn — Resolve a Merge Conflict in an Analysis Script"}
Two branches have made competing edits to the same `filter_samples()` function.

**Branch `feat/read-depth-filter`** has:

```r
filter_samples <- function(counts, metadata) {
  keep <- colSums(counts) >= 10e6   # min 10 M reads
  counts[, keep]
}
```

**Branch `feat/mt-filter`** has:

```r
filter_samples <- function(counts, metadata) {
  keep <- metadata$pct_mt <= 0.1    # max 10 % mitochondrial
  counts[, keep]
}
```

When you merge `feat/mt-filter` into `feat/read-depth-filter`, Git cannot resolve this automatically because both branches edited the same lines.

**Task:** Manually resolve the conflict so that the final function applies **both** filters. Write the resolved function below.
:::

::: {.callout-tip collapse="true" title="Solution"}
The merged function keeps both conditions. Apply them in sequence (read-depth first, then MT percentage, which is a common QC ordering):

```r
filter_samples <- function(counts, metadata,
                           min_reads   = 10e6,
                           max_pct_mt  = 0.1) {
  # Filter 1: minimum library size
  keep_reads <- colSums(counts) >= min_reads

  # Filter 2: maximum mitochondrial fraction
  keep_mt    <- metadata$pct_mt <= max_pct_mt

  keep <- keep_reads & keep_mt

  message(sprintf(
    "Retained %d / %d samples after QC filters.",
    sum(keep), length(keep)
  ))

  counts[, keep]
}
```

After editing the file to remove all conflict markers, stage and commit:

```bash
git add R/qc.R
git commit -m "fix(qc): merge read-depth and MT filters into filter_samples()"
```

Making the thresholds arguments (with defaults) is better practice than hardcoding them — it makes the function reusable for different experiments.
:::

## The .gitignore File

Create a `.gitignore` file to exclude files that should not be tracked:

```{bash}
#| eval: false

# .gitignore for R projects
.Rhistory
.RData
.Rproj.user/
renv/library/
renv/staging/

# Generated output
*.html
*.pdf
_site/

# Large data files
data/raw/*.csv

# OS files
.DS_Store
Thumbs.db
```

Place `.gitignore` in the project root and commit it as your first or second commit.

### Your Turn: Write a .gitignore

Write a `.gitignore` for a project that:

- Uses renv for package management
- Renders Quarto documents to HTML
- Has a `data/raw/` folder with large CSV files that should not be tracked
- Has a `data/clean/` folder with processed RDS files that should be tracked

```{bash}
#| eval: false

# Your .gitignore content:
.Rhistory
.RData
.Rproj.user/

# renv
renv/library/
renv/staging/

# Rendered output
*.html
_site/
_freeze/

# Large raw data (but NOT clean data)
data/raw/*.csv

# OS files
.DS_Store
Thumbs.db
```

::: {.callout-caution title="Your Turn — Bioinformatics .gitignore"}
Write a `.gitignore` for an RNA-seq project that:

- Excludes raw FASTQ files (too large for Git)
- Excludes `renv/library/` (reproducible via `renv::restore()`)
- Excludes `.Rproj.user/` (IDE state)
- Excludes rendered HTML output
- Excludes `_targets/objects/` (large cached targets)
- **Includes** `_targets/meta/meta` (the targets dependency graph — small and important)
- **Includes** small example data files in `data/example/`

Write the file with comments explaining each section.
:::

::: {.callout-tip collapse="true" title="Solution"}
```bash
# ─── R / RStudio ─────────────────────────────────────────────────────────────
.Rhistory
.RData
.Rproj.user/

# ─── renv: exclude the installed library, keep the lockfile ──────────────────
renv/library/
renv/staging/
# renv.lock is TRACKED — it is what makes the project reproducible

# ─── Quarto rendered output ───────────────────────────────────────────────────
*.html
*.pdf
_site/
_freeze/

# ─── Raw sequencing data (too large for Git; use a data repository) ───────────
*.fastq
*.fastq.gz
*.fq
*.fq.gz
*.bam
*.bam.bai
data/raw/

# ─── targets pipeline cache ───────────────────────────────────────────────────
_targets/objects/     # large cached R objects — do NOT commit
_targets/scratch/
!_targets/meta/       # the dependency graph is small; ALWAYS commit it
!_targets/meta/meta

# ─── Example data (small, committed for reproducibility) ──────────────────────
# data/example/ is intentionally NOT excluded
# If your raw data folder pattern catches too much, be explicit:
!data/example/

# ─── OS and editor noise ──────────────────────────────────────────────────────
.DS_Store
Thumbs.db
*.swp
```

Key points:
- `renv.lock` is always tracked; `renv/library/` never is.
- The `!` prefix negates a previous exclusion pattern — order matters. Put `!_targets/meta/` after `_targets/`.
- FASTQ / BAM files belong in a dedicated data repository (e.g., SRA, Zenodo, or institutional storage), not in Git.
- `_targets/meta/meta` is the targets metadata file. Committing it means collaborators see the same pipeline dependency graph.
:::

::: {.callout-caution title="Your Turn — Atomic Commits Practice"}
You have been working on an RNA-seq project and your working directory now has three unrelated changes ready to be staged:

1. You added a new `normalize_counts()` function to `R/normalize.R`
2. You fixed a typo in `README.md` ("reproduciblity" -> "reproducibility")
3. You updated `.gitignore` to exclude `*.bam` files

**Task:** Use `git add -p` (patch mode) to stage and commit these as **three separate atomic commits** with proper conventional commit messages.

Write the three `git commit -m` commands you would use.
:::

::: {.callout-tip collapse="true" title="Solution"}
```bash
# Stage only the normalize.R changes (use git add -p and select only those hunks)
git add -p R/normalize.R
git commit -m "feat(normalization): add normalize_counts() using DESeq2 size factors"

# Stage only the README change
git add -p README.md
git commit -m "docs(README): fix typo in reproducibility section"

# Stage only the .gitignore change
git add .gitignore
git commit -m "chore(gitignore): exclude BAM alignment files"
```

`git add -p` enters interactive patch mode. Git shows each changed "hunk" and asks `Stage this hunk? [y/n/q/a/d/?]`. Answer `y` for hunks you want in the current commit, `n` for hunks that belong in a later commit. This is the cleanest way to split logically unrelated edits that happen to be in the same file.

**Why atomic commits matter:**
- Each commit does exactly one thing — it is easy to revert with `git revert`
- `git bisect` can pinpoint which commit introduced a bug
- Code review is easier when reviewers can evaluate each change independently
:::

## Git in Positron

Positron has a built-in Source Control pane:

- **Stage**: Click the `+` next to changed files
- **Commit**: Type a message in the text box and click Commit
- **Push/Pull**: Use the arrows in the Source Control header
- **Branch**: Click the branch name in the bottom status bar
- **Diff**: Click a changed file to see what changed

You can also open the integrated terminal (Ctrl+` or Cmd+`) to run Git commands directly.

::: {.callout-tip title="Challenge"}
Set up a complete project with Git:

1. Create a new Positron project with Git enabled
2. Add a `.gitignore` and commit it
3. Create an `analysis.qmd` file with a basic Quarto document
4. Commit the initial analysis
5. Create a branch using `git switch -c`, add a ggplot2 visualization, commit on the branch
6. Merge the branch back to main
7. Push everything to GitHub

This exercise combines everything from Part 1.
:::

## Automating with GitHub Actions

GitHub Actions runs workflows automatically on every `git push`. For a Quarto report backed by `renv`, this means every push re-renders your report in a clean environment — catching any "works on my machine" issues immediately.

### Minimal render workflow

Create the file `.github/workflows/render.yml` in your repository:

```{bash}
#| eval: false
mkdir -p .github/workflows
```

```yaml
# .github/workflows/render.yml
on: [push]

jobs:
  render:
    runs-on: ubuntu-latest
    steps:
      - uses: actions/checkout@v4

      - uses: r-lib/actions/setup-r@v2

      - uses: quarto-dev/quarto-actions/setup@v2

      - uses: r-lib/actions/setup-renv@v2

      - run: quarto render report.qmd
```

The four `uses:` steps handle everything: checking out the repository, installing R, installing Quarto, and restoring your exact package versions from `renv.lock`. The final `run:` step renders your report.

::: {.callout-note title="Why this matters"}
Every `git push` automatically rebuilds your Quarto report. This is how modern bioinformatics labs ensure reports stay in sync with code. If a collaborator's change breaks the render, the workflow fails and GitHub sends a notification — before anyone notices the HTML is stale.
:::

Commit the workflow file:

```{bash}
#| eval: false
git add .github/workflows/render.yml
git commit -m "ci: add GitHub Actions workflow to render report on push"
git push
```

After pushing, go to the **Actions** tab on your GitHub repository to watch the workflow run.

::: {.callout-caution title="Your Turn — GitHub Actions Debugging"}
The following `render.yml` causes a workflow failure. The error message on GitHub is:

```
Error in loadNamespace(name) : there is no package called 'DESeq2'
Calls: ... :: quarto render -> source -> ...
Execution halted
```

Here is the broken workflow:

```yaml
on: [push]

jobs:
  render:
    runs-on: ubuntu-latest
    steps:
      - uses: actions/checkout@v4
      - uses: r-lib/actions/setup-r@v2
      - uses: quarto-dev/quarto-actions/setup@v2
      - run: quarto render report.qmd
```

**Task:** Identify what is missing and write the corrected `render.yml`.
:::

::: {.callout-tip collapse="true" title="Solution"}
The missing step is `r-lib/actions/setup-renv@v2`. Without it, the runner has a bare R installation with no packages. The `renv.lock` file is present in the repository but nothing reads it.

Corrected workflow:

```yaml
on: [push]

jobs:
  render:
    runs-on: ubuntu-latest
    steps:
      - uses: actions/checkout@v4

      - uses: r-lib/actions/setup-r@v2

      - uses: quarto-dev/quarto-actions/setup@v2

      - uses: r-lib/actions/setup-renv@v2    # <-- this was missing

      - run: quarto render report.qmd
```

**Why the dependency chain matters:**

1. `checkout` — makes your code (including `renv.lock`) available
2. `setup-r` — installs R
3. `setup-quarto` — installs Quarto (can come before or after R)
4. `setup-renv` — reads `renv.lock`, calls `renv::restore()`, installs all packages including `DESeq2`
5. `quarto render` — now all packages exist

If `setup-renv` is missing, step 5 fails immediately because R has no packages beyond base.
:::

# Part 2: Capstone Project

## Project Specification

Your final project demonstrates the skills you have built over 15 sessions. Choose a dataset and produce a complete, reproducible analysis.

**Requirements:**

1. An R project using `here` for file paths
2. At least one data import and cleaning step
3. At least three ggplot2 visualizations
4. At least one summary table
5. A Quarto document that tells the story of your analysis
6. A Git repository with meaningful commit history
7. A README.md that describes the project

## Project Structure Template

```{r}
#| eval: false

# Recommended project layout
structure <- '
my-project/
  .gitignore
  README.md
  renv.lock
  my-project.Rproj
  .github/
    workflows/
      render.yml
  R/
    01-import.R
    02-clean.R
    03-analyze.R
    04-visualize.R
  data/
    raw/         # Original data (possibly gitignored)
    clean/       # Processed data
  output/
    figures/     # Saved plots
    tables/      # Saved tables
  report.qmd    # Main Quarto document
'
```

## Capstone Setup: Reproducible Environment

Before writing any analysis code, set up `renv` so your package environment is locked and reproducible.

```{r}
#| eval: false

# Initialize renv with Bioconductor support
renv::init(bioc = TRUE)
```

::: {.callout-important title="Why bioc = TRUE?"}
Pass `bioc = TRUE` so `renv` recognizes Bioconductor packages and pins their version compatibly with your R version. Without this, installing `DESeq2` or `limma` later will produce confusing version mismatches. Bioconductor has its own release cycle tied to R versions; `renv` needs to know about this to generate a correct `renv.lock`.
:::

For faster Bioconductor + CRAN installs, use `pak::pkg_install()`. It downloads packages in parallel and has smarter dependency resolution than either `install.packages()` or `BiocManager::install()`.

```{r}
#| eval: false

# Install pak if not already available
install.packages("pak")

# Install a CRAN package
pak::pkg_install("dplyr")

# Install a Bioconductor package
pak::pkg_install("bioc::DESeq2")

# Install multiple packages at once (runs in parallel)
pak::pkg_install(c("ggplot2", "here", "bioc::airway", "bioc::DESeq2"))
```

::: {.callout-note title="pak vs install.packages"}
`pak::pkg_install()` is the modern replacement for both `install.packages()` and `BiocManager::install()` combined. It is typically 2-10x faster because it downloads binary packages in parallel and only reinstalls what has changed. Use `pak` for interactive installs; `renv` handles restoration in automated contexts.
:::

## Step 1: Choose a Dataset

Pick a dataset that interests you. Some options:

```{r}
#| eval: false

# Built-in datasets
data(diamonds, package = "ggplot2")   # 54,000 diamonds with price, cut, etc.
data(storms, package = "dplyr")       # Atlantic storms from 1975-2021
data(starwars, package = "dplyr")     # Star Wars characters
data(gapminder, package = "gapminder") # Country-level economic data

# Or use your own data (CSV, Excel, API, web scraping)
```

::: {.callout-tip title="Recommended bioinformatics datasets"}
If you are taking this course as part of a bioinformatics programme, consider using a biological dataset so your capstone demonstrates domain-relevant skills.

**`airway` (Bioconductor) — recommended**

RNA-seq read counts from human airway smooth muscle cells (Himes et al. 2014). Eight samples across two conditions (treated/untreated with dexamethasone). This is the canonical teaching dataset for DESeq2.

```r
pak::pkg_install("bioc::airway")
library(airway)
data(airway)
airway   # SummarizedExperiment with 63,677 genes x 8 samples
```

**`palmerpenguins` — clean tabular alternative**

If you prefer a tidy data frame without Bioconductor dependencies, `palmerpenguins` is a well-documented alternative to `iris`. Three penguin species, body measurements, and island data.

```r
pak::pkg_install("palmerpenguins")
library(palmerpenguins)
glimpse(penguins)
```

**GEO public dataset — advanced option**

Download a small, pre-processed DESeq2 result table from NCBI GEO (Gene Expression Omnibus). For example, GSE52778 accompanies the `airway` package. Use the `GEOquery` Bioconductor package:

```r
pak::pkg_install("bioc::GEOquery")
library(GEOquery)
gse <- getGEO("GSE52778", GSEMatrix = TRUE)
```

You are free to use any dataset you find interesting. The bioinformatics options are nudges, not requirements.
:::

### Your Turn: Pick and Explore

Choose your dataset and run an initial exploration:

```{r}
#| eval: false

# Load your data
library(tidyverse)

# Replace with your chosen dataset
my_data <- diamonds

# Explore
glimpse(my_data)
summary(my_data)

# How many rows and columns?
dim(my_data)

# Any missing values?
colSums(is.na(my_data))
```

::: {.callout-caution title="Your Turn — Capstone Bioinformatics Path (airway)"}
This is a mini-version of the capstone using the `airway` dataset. Complete each step.

**Step 1:** Load the data and extract a count matrix and sample metadata.

**Step 2:** Compute total read counts per sample (column sums of the count matrix).

**Step 3:** Plot total counts per sample as a bar chart with `ggplot2`. Color bars by treatment condition (`dex` column in the metadata).

**Step 4:** Write a parameterized Quarto report where the parameter `min_count` controls which genes are included in the plot (genes with row mean >= `min_count`).

**Step 5:** Render the report with `quarto render report.qmd -P min_count:10`.
:::

::: {.callout-tip collapse="true" title="Solution"}
```r
# Step 1: Load data
library(airway)
library(SummarizedExperiment)
library(tidyverse)

data(airway)
counts   <- assay(airway, "counts")        # matrix: genes x samples
metadata <- as.data.frame(colData(airway)) # sample-level metadata

# Step 2: Total counts per sample
total_counts <- colSums(counts)
metadata$total_counts <- total_counts

# Step 3: Bar chart coloured by treatment
ggplot(metadata, aes(x = rownames(metadata), y = total_counts / 1e6,
                     fill = dex)) +
  geom_col() +
  scale_fill_manual(values = c("untrt" = "steelblue", "trt" = "tomato"),
                    labels = c("untrt" = "Untreated", "trt" = "Dexamethasone")) +
  labs(title = "Total Sequencing Depth per Sample",
       x = "Sample", y = "Total Counts (millions)",
       fill = "Treatment") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

For the parameterized report, add to the YAML header:

```yaml
params:
  min_count: 10
```

Then filter in a code chunk:

```r
# Filter genes by minimum average count
keep <- rowMeans(counts) >= params$min_count
counts_filtered <- counts[keep, ]
cat(sprintf("Retained %d / %d genes with mean count >= %d\n",
            nrow(counts_filtered), nrow(counts), params$min_count))
```

Render from the terminal:

```bash
quarto render report.qmd -P min_count:10
quarto render report.qmd -P min_count:100
```
:::

## Step 2: Clean and Transform

Apply the tidyverse skills from Sessions 5-7:

```{r}
#| eval: false

library(tidyverse)

# Example cleaning pipeline
clean_data <- my_data |>
  # Remove rows with missing key variables
  filter(!is.na(price), !is.na(carat)) |>
  # Create new variables
  mutate(
    price_per_carat = price / carat,
    size_category = case_when(
      carat < 0.5  ~ "Small",
      carat < 1.0  ~ "Medium",
      carat < 2.0  ~ "Large",
      .default = "Very Large"
    )
  ) |>
  # Remove extreme outliers
  filter(price_per_carat < quantile(price_per_carat, 0.99))
```

### Your Turn: Clean Your Data

Write a cleaning pipeline for your chosen dataset. Include at least:

- One `filter()` step
- One `mutate()` step
- One type conversion or recoding step

```{r}
#| eval: false

# Your cleaning pipeline here
```

## Step 3: Summarize

Create summary statistics using dplyr:

```{r}
#| eval: false

# Example summaries
summary_by_cut <- clean_data |>
  summarize(
    n = n(),
    mean_price = mean(price),
    median_price = median(price),
    mean_carat = mean(carat),
    .by = cut
  ) |>
  arrange(desc(mean_price))

summary_by_cut
```

### Your Turn: Create Summaries

Produce at least two summary tables for your data:

```{r}
#| eval: false

# Summary 1: grouped summary

# Summary 2: cross-tabulation or different grouping
```

## Step 4: Visualize

Create at least three visualizations using ggplot2 (Sessions 8-9):

```{r}
#| eval: false

library(ggplot2)

# Visualization 1: Distribution
ggplot(clean_data, aes(x = price)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  scale_x_log10(labels = scales::dollar) +
  labs(title = "Distribution of Diamond Prices",
       x = "Price (log scale)", y = "Count") +
  theme_minimal()

# Visualization 2: Relationship
ggplot(clean_data, aes(x = carat, y = price, color = cut)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_continuous(labels = scales::dollar) +
  labs(title = "Price vs Carat by Cut Quality",
       x = "Carat", y = "Price", color = "Cut") +
  theme_minimal()

# Visualization 3: Comparison
ggplot(clean_data, aes(x = cut, y = price_per_carat, fill = cut)) +
  geom_boxplot(show.legend = FALSE) +
  scale_y_continuous(labels = scales::dollar) +
  labs(title = "Price per Carat by Cut Quality",
       x = "Cut", y = "Price per Carat") +
  theme_minimal() +
  coord_flip()
```

### Your Turn: Create Your Visualizations

Build three plots for your dataset. Each should show something different:

1. A distribution (histogram, density, or bar chart)
2. A relationship (scatter plot, line plot, or heatmap)
3. A comparison (box plot, faceted plot, or grouped bar chart)

```{r}
#| eval: false

# Visualization 1: Distribution

# Visualization 2: Relationship

# Visualization 3: Comparison
```

## Step 5: Write Your Quarto Report

Create a `report.qmd` that tells the story. Use the skills from Sessions 11-13:

```{r}
#| eval: false

# Template for report.qmd:
report_template <- '
---
title: "Your Project Title"
author: "Your Name"
date: today
format:
  html:
    toc: true
    code-fold: true
    theme: cosmo
execute:
  echo: true
  warning: false
---

## Introduction

What question are you exploring? Why does it matter?

## Data

Where does the data come from? What does each variable represent?

## Cleaning and Preparation

What steps did you take to prepare the data?

## Analysis

What patterns did you find?

## Visualizations

Show and explain your three plots.

## Findings

What did you learn? What surprised you?

## Conclusion

Summary of results and possible next steps.

## Reproducibility

```{r}
sessioninfo::session_info()
```
'
```

### Your Turn: Draft Your Report

Open a new Quarto document and write the first draft of your report. Include:

- A title and author
- An introduction explaining your question
- At least one code chunk that loads and cleans data
- At least one visualization with interpretation

## Step 6: Version Control Your Project

Initialize Git and make meaningful commits as you work:

```{bash}
#| eval: false

# Initialize
cd my-project
git init

# First commit: project structure
git add .gitignore README.md
git commit -m "Initialize project structure with gitignore and README"

# Second commit: data import
git add R/01-import.R
git commit -m "Add data import script for diamonds dataset"

# Third commit: cleaning
git add R/02-clean.R
git commit -m "Add cleaning pipeline: filter outliers, create size categories"

# Continue committing as you work...
```

## Step 7: Snapshot Your Environment

::: {.callout-important title="Non-optional reproducibility artifacts"}
These two artifacts are what makes your analysis reproducible. Without them, your collaborator (or your future self) cannot recreate your results.
:::

**Step 7a — Snapshot**

After installing all packages needed for your analysis, take a snapshot:

```{r}
#| eval: false

renv::snapshot()
```

Verify that `renv.lock` exists and commit it:

```{bash}
#| eval: false

# Confirm the lockfile was created or updated
ls renv.lock

git add renv.lock
git commit -m "chore(renv): snapshot package environment after adding airway"
```

**Step 7b — Session info**

Add a final code chunk at the end of your `report.qmd`:

```{r}
#| eval: false

# Prefer sessioninfo over base sessionInfo() for cleaner output
# install if needed: pak::pkg_install("sessioninfo")
sessioninfo::session_info()
```

This chunk records the exact R version, platform, and package versions used when the report was last rendered. Commit the rendered output and the source together.

::: {.callout-caution title="Your Turn — renv Round-Trip"}
Simulate the full reproducibility round-trip that a collaborator would experience.

**On your machine:**

1. Install the `airway` Bioconductor package using `pak::pkg_install("bioc::airway")`
2. Run `renv::snapshot()` and confirm `renv.lock` lists `airway`
3. Commit `renv.lock` and push to GitHub

**Simulating a clean clone:**

4. Create a new temporary directory and clone the repository (or copy without the `renv/library/` folder)
5. Open R in that directory and run `renv::restore()`
6. Confirm that `airway` is now installed and `packageVersion("airway")` matches the version in `renv.lock`

Write the commands for each step and describe what output you expect from `renv::restore()`.
:::

::: {.callout-tip collapse="true" title="Solution"}
```r
# --- On your machine ---

# Step 1: install airway
pak::pkg_install("bioc::airway")
library(airway)  # verify it loads

# Step 2: snapshot
renv::snapshot()
# Expected output:
# The following package(s) will be updated in the lockfile:
# - airway   [* -> 1.42.0]
# Do you want to proceed? [y/N]: y

# Verify airway appears in renv.lock
readLines("renv.lock") |> grep("airway", x = _, value = TRUE)
# Should show: "Package": "airway",

# Step 3: commit and push
# (in terminal)
# git add renv.lock
# git commit -m "chore(renv): add airway to lockfile"
# git push
```

```r
# --- Simulating a clean clone ---

# Step 4: in a fresh R session with no renv library:
setwd("/path/to/cloned/repo")

# Step 5: restore
renv::restore()
# Expected output:
# The following packages will be installed:
#   airway       [1.42.0]
#   ... (all other locked packages)
# Do you want to proceed? [y/N]: y
# Installing airway [1.42.0] ...
# OK (installed binary)

# Step 6: verify
library(airway)
packageVersion("airway")
# Should match renv.lock exactly, e.g., '1.42.0'
```

The key guarantee: `renv::restore()` reads `renv.lock` and installs exactly those versions — not the latest available on CRAN or Bioconductor. This is what makes your analysis reproducible months or years later.
:::

::: {.callout-caution title="Your Turn — pak vs install.packages Benchmark"}
Compare installation speed between `install.packages()` and `pak::pkg_install()` for a small set of packages.

**Task:** Use `system.time()` to time installing `dplyr`, `ggplot2`, and `airway` using both methods. (Remove the packages first so they are not already cached, or run in a fresh R session / `renv` sandbox.)

Write the benchmark code and predict: which method will be faster, and why?
:::

::: {.callout-tip collapse="true" title="Solution"}
```r
# Benchmark install.packages (CRAN only; airway requires BiocManager separately)
t1 <- system.time({
  install.packages(c("dplyr", "ggplot2"), repos = "https://cloud.r-project.org")
  BiocManager::install("airway", ask = FALSE)
})

# Benchmark pak (handles CRAN and Bioconductor in one call)
t2 <- system.time({
  pak::pkg_install(c("dplyr", "ggplot2", "bioc::airway"))
})

cat("install.packages + BiocManager:", t1["elapsed"], "seconds\n")
cat("pak::pkg_install:              ", t2["elapsed"], "seconds\n")
cat("Speedup factor:", round(t1["elapsed"] / t2["elapsed"], 1), "x\n")
```

**Why `pak` is typically 2-10x faster:**

1. **Parallel downloads** — `pak` downloads multiple packages simultaneously. `install.packages()` downloads one at a time.
2. **Binary-first** — `pak` prefers pre-compiled binaries (RSPM / Posit Package Manager) over source compilation, especially on Linux.
3. **Smarter dependency resolution** — `pak` computes the full dependency graph before downloading anything, avoiding redundant work.
4. **Unified interface** — no need to call `BiocManager::install()` separately; `pak` handles both CRAN and Bioconductor packages in one pass.

In practice, the speedup is most dramatic for Bioconductor packages, which have deep dependency trees.
:::

::: {.callout-tip title="Challenge"}
**Full capstone project:**

Build a complete analysis that uses skills from every phase of the course:

**Base R (Sessions 1-4):**

- Use `tryCatch()` for error handling in data import
- Write at least one custom function

**Tidyverse (Sessions 5-10):**

- Full dplyr pipeline with `filter()`, `mutate()`, `summarize()`
- At least one pivot operation
- At least one string or date manipulation
- Three ggplot2 visualizations with custom themes

**Quarto (Sessions 11-13):**

- Parameterized report (e.g., parameter for which region to analyze)
- Cross-references for figures
- A table of contents

**Professional Practices (Sessions 14-15):**

- `here` package for all file paths
- `renv::init(bioc = TRUE)` for dependency management
- Clean code following the tidyverse style guide
- Git repository with 10+ meaningful commits using `git switch`/`git restore`
- README.md documenting the project
- GitHub Actions workflow that renders the report on every push
- `renv::snapshot()` committed and verified
- `sessioninfo::session_info()` chunk at the end of the report

This is the full capstone. Work on it incrementally, committing as you go.
:::

::: {.callout-caution title="Your Turn — Complete Reproducible Mini-Project (Bonus Challenge)"}
Set up a complete reproducible mini-project from scratch. This integrates everything from Sessions 14-15.

Work through the following checklist, running each command in order. Write down the exact commands you use.

- [ ] `mkdir rnaseq-mini && cd rnaseq-mini && git init`
- [ ] Create `.gitignore` (use the bioinformatics template from the earlier exercise)
- [ ] `git add .gitignore && git commit -m "chore: add bioinformatics gitignore"`
- [ ] Open R: `renv::init(bioc = TRUE)`
- [ ] `pak::pkg_install(c("bioc::airway", "ggplot2", "here", "sessioninfo"))`
- [ ] Write `analysis.R`: load `airway`, compute total counts, save a bar chart to `output/figures/`
- [ ] Write `report.qmd`: source `analysis.R`, display the chart, call `sessioninfo::session_info()`
- [ ] `renv::snapshot()`
- [ ] `git add renv.lock R/ report.qmd && git commit -m "feat: add airway analysis and report"`
- [ ] Create `.github/workflows/render.yml` using the template from this workbook
- [ ] `git add .github/ && git commit -m "ci: add GitHub Actions render workflow"`
- [ ] Push to GitHub: `git push -u origin main`
- [ ] Navigate to the Actions tab on GitHub and confirm the workflow passes
:::

::: {.callout-tip collapse="true" title="Solution"}
Step-by-step commands with expected output:

```bash
# Terminal
mkdir rnaseq-mini
cd rnaseq-mini
git init
# Initialized empty Git repository in .../rnaseq-mini/.git/
```

Create `.gitignore` (contents from the bioinformatics exercise above), then:

```bash
git add .gitignore
git commit -m "chore: add bioinformatics gitignore"
```

```r
# R session
renv::init(bioc = TRUE)
# * Initializing project ...
# * Discovering package dependencies ...
# * Copying packages into the cache ...
# * Snapshot written to renv.lock

pak::pkg_install(c("bioc::airway", "ggplot2", "here", "sessioninfo"))
```

`analysis.R`:

```r
library(airway)
library(ggplot2)
library(here)
library(SummarizedExperiment)

data(airway)
counts   <- assay(airway, "counts")
metadata <- as.data.frame(colData(airway))
metadata$total_counts <- colSums(counts)
metadata$sample       <- rownames(metadata)

p <- ggplot(metadata, aes(x = sample, y = total_counts / 1e6, fill = dex)) +
  geom_col() +
  labs(title = "Sequencing depth per sample",
       x = "Sample", y = "Counts (M)", fill = "Treatment") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

dir.create(here("output", "figures"), recursive = TRUE, showWarnings = FALSE)
ggsave(here("output", "figures", "depth.png"), p, width = 7, height = 4)
```

`report.qmd` final chunk:

````
```{r}
sessioninfo::session_info()
```
````

```r
renv::snapshot()
```

```bash
git add renv.lock analysis.R report.qmd output/
git commit -m "feat: add airway analysis and Quarto report"

git add .github/
git commit -m "ci: add GitHub Actions render workflow"

git remote add origin https://github.com/YOUR_USERNAME/rnaseq-mini.git
git push -u origin main
```

After pushing, go to `github.com/YOUR_USERNAME/rnaseq-mini/actions`. You should see a workflow run in progress. A green checkmark means the report rendered successfully in a clean cloud environment — proof that your project is fully reproducible.
:::

# Part 3: Course Review and Reflection

## Skills Checklist

Rate your confidence (1 = need review, 5 = confident) on each skill:

```{r}
#| eval: false

skills <- tibble::tibble(
  phase = c(rep("Base R", 7), rep("Tidyverse", 8),
            rep("Quarto", 5), rep("Professional", 7)),
  skill = c(
    # Base R
    "Create and manipulate vectors, lists, data frames",
    "Subset data with numeric, logical, and name indexing",
    "Write if/else, for loops, and while loops",
    "Write functions with default arguments and validation",
    "Use lapply, sapply, vapply, tapply",
    "Handle strings with paste, grep, gsub, sprintf",
    "Create base R plots: scatter, histogram, boxplot, bar",
    # Tidyverse
    "Use the pipe |> to chain operations",
    "filter, select, mutate, summarize, arrange with dplyr",
    "Reshape data with pivot_longer and pivot_wider",
    "Import data with readr and readxl",
    "Manipulate strings with stringr",
    "Work with dates using lubridate",
    "Use map functions from purrr",
    "Build layered ggplot2 visualizations with custom themes",
    # Quarto
    "Write Quarto documents with YAML, markdown, and code chunks",
    "Render to HTML, PDF, and Word",
    "Create Reveal.js presentations",
    "Build dashboards with value boxes and layouts",
    "Publish to Quarto Pub or GitHub Pages",
    # Professional
    "Organize projects with here and renv (bioc = TRUE)",
    "Debug with browser(), traceback(), and Positron tools",
    "Follow the tidyverse style guide",
    "Use git switch / git restore for branching and file ops",
    "Write conventional commit messages",
    "Snapshot and restore environments with renv",
    "Automate rendering with GitHub Actions"
  ),
  confidence = NA_integer_
)

# Fill in your confidence ratings:
# skills$confidence <- c(...)
```

### Your Turn: Identify Growth Areas

1. Which three skills improved the most during this course?
2. Which two skills do you want to practice more?
3. What is one project idea where you could apply these skills?

## Think About It

Reflect on the full course:

- What was the most surprising thing you learned about R?
- How does R compare to other languages you know for data work?
- What would you do differently if you started the course over?
- What is the first real project you want to tackle with R?

## Resources for Continued Learning

**Books (free online):**

- R for Data Science (2e): <https://r4ds.hadley.nz>
- Advanced R: <https://adv-r.hadley.nz>
- ggplot2 Book: <https://ggplot2-book.org>
- Quarto documentation: <https://quarto.org/docs/guide/>

**Bioinformatics R:**

- Bioconductor workflows: <https://bioconductor.org/packages/release/workflows/>
- DESeq2 vignette: <https://bioconductor.org/packages/DESeq2>
- Orchestrating Single-Cell Analysis: <https://bioconductor.org/books/release/OSCA/>

**Next steps to explore:**

- Shiny: interactive web applications
- tidymodels: machine learning in R
- R packages: build and distribute your own code
- gganimate: animated visualizations
- targets: reproducible pipeline management (pairs naturally with `renv`)

**Community:**

- Posit Community Forum: <https://forum.posit.co>
- R-bloggers: <https://www.r-bloggers.com>
- Bioconductor Support: <https://support.bioconductor.org>

# Session Synthesis

### Connect the Dots: Complete Project Setup

This exercise ties together Git, project structure, and all course skills. Work through it step by step.

```{bash}
#| eval: false

# 1. Create and initialize the project
mkdir course-capstone
cd course-capstone
git init

# 2. Create .gitignore
cat > .gitignore << 'EOF'
.Rhistory
.RData
.Rproj.user/
renv/library/
renv/staging/
*.html
_site/
.DS_Store
EOF

git add .gitignore
git commit -m "Add .gitignore for R project"

# 3. Create README
cat > README.md << 'EOF'
# Course Capstone Project

Analysis of [your dataset] using R.

## Structure
- `R/` - Analysis scripts
- `data/` - Raw and processed data
- `output/` - Figures and tables
- `report.qmd` - Final report

## Reproduction
1. Open the .Rproj file in Positron
2. Run `renv::restore()` to install packages
3. Render `report.qmd`
EOF

git add README.md
git commit -m "Add README with project description and reproduction steps"

# 4. Create directory structure
mkdir -p R data/raw data/clean output/figures output/tables

# 5. Continue building from here...
```

```{r}
#| eval: false

# In R, set up the project:
library(tidyverse)
library(here)

# Initialize renv with Bioconductor support
renv::init(bioc = TRUE)

# Import data
my_data <- read_csv(here("data", "raw", "my_data.csv"))

# Clean
clean_data <- my_data |>
  filter(!is.na(key_variable)) |>
  mutate(
    category = case_when(
      value < 10 ~ "Low",
      value < 50 ~ "Medium",
      .default = "High"
    )
  )

# Save cleaned version
write_rds(clean_data, here("data", "clean", "clean_data.rds"))

# Summarize
summary_table <- clean_data |>
  summarize(
    n = n(),
    mean_val = mean(value),
    sd_val = sd(value),
    .by = category
  )

# Visualize
p1 <- ggplot(clean_data, aes(x = value, fill = category)) +
  geom_histogram(bins = 30, color = "white") +
  labs(title = "Distribution of Values by Category") +
  theme_minimal()

ggsave(here("output", "figures", "distribution.png"), p1,
       width = 8, height = 5)

# Snapshot the environment
renv::snapshot()

# Session info at end of report
sessioninfo::session_info()
```

## Key Takeaways

1. Git tracks every change to your project. Commit early, commit often, write clear messages. Use `git switch` for branches and `git restore` for file-level changes.
2. Branches let you experiment without breaking the main codebase. A clear branch strategy prevents painful conflicts.
3. A well-structured project with Git, `renv::init(bioc = TRUE)`, `renv::snapshot()`, and a Quarto report is fully reproducible by anyone.
4. GitHub Actions closes the loop: every push re-renders your report in a clean environment, proving reproducibility continuously.
5. You now have a complete R toolkit: data wrangling, visualization, reporting, and professional practices.

## Course Complete

You started this course knowing programming but not R. Over 15 sessions, you built skills in:

- Base R fundamentals
- Tidyverse data wrangling and visualization
- Quarto for reproducible communication
- Professional development practices including Git, renv, and CI/CD

Take these skills and apply them to problems that matter to you. The best way to keep learning R is to keep using R.
