---
title: "Session 9: Advanced Tidyverse"
subtitle: "R Programming Workbook"
author: "R Programming Course"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-tools: true
    code-overflow: wrap
    theme: cosmo
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-note title="Learning Objectives"}
By the end of this session you will be able to:

1. Use the embrace operator `{{ }}` to write functions that accept column names
2. Apply the walrus operator `:=` for dynamic column naming
3. Use `.data` and `.env` pronouns for string-based column access
4. Create function factories that return specialized functions
5. Apply `reduce()` and `accumulate()` for sequential operations
6. Parse JSON data with jsonlite
7. Build and perform HTTP requests with httr2
8. Handle nested API responses with purrr
9. Implement error handling and rate limiting for API workflows
:::

```{r}
#| label: setup
library(dplyr)
library(purrr)
library(jsonlite)
library(tidyr)
library(ggplot2)

set.seed(42)
sales <- tibble(
  date = rep(seq(as.Date("2024-01-01"), as.Date("2024-12-31"), by = "day"), 3),
  region = rep(c("North", "South", "West"), each = 366),
  product = sample(c("Widget", "Gadget", "Doohickey"), 366 * 3, replace = TRUE),
  units = rpois(366 * 3, lambda = 50),
  revenue = round(runif(366 * 3, 10, 500), 2),
  returned = rbinom(366 * 3, 1, 0.05)
)
```

## Part 1: Tidy Evaluation

### The Problem: Column Names in Functions

When you write a function that calls dplyr verbs, you cannot pass column names like regular variables.

```{r}
# This does NOT work as expected
bad_mean <- function(data, col) {
  data |> summarize(result = mean(col))
}

# Try it -- you get an error or wrong result
# bad_mean(sales, revenue)
```

dplyr uses **data masking**: it treats bare names (like `revenue`) as columns, not as R variables. Inside your function, `col` is a symbol, not the column itself.

### The Embrace Operator {{ }}

Wrap function arguments in `{{ }}` to pass them through to dplyr:

```{r}
good_mean <- function(data, col) {
  data |> summarize(result = mean({{ col }}))
}

good_mean(sales, revenue)
good_mean(sales, units)
```

`{{ }}` tells dplyr: "evaluate this argument the way the caller intended."

### Embrace in Multiple Contexts

```{r}
# Works in filter
filter_above <- function(data, col, threshold) {
  data |> filter({{ col }} > threshold)
}

sales |> filter_above(revenue, 400) |> head(5)
```

```{r}
# Works in mutate
add_scaled <- function(data, col) {
  data |> mutate(scaled = ({{ col }} - mean({{ col }})) / sd({{ col }}))
}

sales |> add_scaled(revenue) |> select(revenue, scaled) |> head(5)
```

```{r}
# Works in .by
grouped_summary <- function(data, group_col, value_col) {
  data |>
    summarize(
      mean_val = mean({{ value_col }}),
      n = n(),
      .by = {{ group_col }}
    )
}

sales |> grouped_summary(region, revenue)
```

### Dynamic Column Names with :=

Use the walrus operator `:=` with glue syntax for dynamic output names:

```{r}
named_summary <- function(data, group_col, value_col) {
  data |>
    summarize(
      "mean_{{ value_col }}" := mean({{ value_col }}),
      "sd_{{ value_col }}" := sd({{ value_col }}),
      .by = {{ group_col }}
    )
}

sales |> named_summary(region, revenue)
sales |> named_summary(product, units)
```

### .data and .env Pronouns

When column names are strings (not bare symbols), use `.data[[ ]]`:

```{r}
col_name <- "revenue"
sales |> summarize(avg = mean(.data[[col_name]]), .by = region)
```

Use `.env` to refer to variables from the surrounding environment inside dplyr:

```{r}
threshold_val <- 300
sales |>
  filter(.data[["revenue"]] > .env[["threshold_val"]]) |>
  nrow()
```

::: {.callout-note title="Base R vs Tidyverse"}
In base R, you access columns with `df[[col_name]]` where `col_name` is a string. No special syntax needed because base R does not use data masking. Tidy evaluation is the price you pay for the convenience of typing `revenue` instead of `df[["revenue"]]` everywhere.
:::

### Your Turn: Exercise 1

Write a function `count_above` that takes a data frame, a column, and a threshold, and returns the number of rows where the column exceeds the threshold.

```{r}
#| code-fold: true
#| code-summary: "Solution"
count_above <- function(data, col, threshold) {
  data |>
    filter({{ col }} > threshold) |>
    nrow()
}

count_above(sales, revenue, 400)
count_above(sales, units, 60)
```

### Your Turn: Exercise 2

Write a function `top_n_by_group` that takes a data frame, a grouping column, a value column, and n, then returns the top n rows by value within each group.

```{r}
#| code-fold: true
#| code-summary: "Solution"
top_n_by_group <- function(data, group_col, value_col, n = 3) {
  data |>
    slice_max({{ value_col }}, n = n, by = {{ group_col }})
}

sales |> top_n_by_group(region, revenue, n = 2) |> head(6)
```

### Your Turn: Exercise 3

Write a function `quick_plot` that takes a data frame, x column, y column, and color column, then produces a scatter plot.

```{r}
#| code-fold: true
#| code-summary: "Solution"
quick_plot <- function(data, x_col, y_col, color_col) {
  ggplot(data, aes(x = {{ x_col }}, y = {{ y_col }}, color = {{ color_col }})) +
    geom_point(alpha = 0.3) +
    theme_minimal()
}

sales |> quick_plot(units, revenue, region)
```

::: {.callout-warning title="Common Mistake"}
Forgetting `{{ }}` inside `aes()`:

```r
# WRONG -- plots nothing useful
quick_plot <- function(data, x_col, y_col) {
  ggplot(data, aes(x = x_col, y = y_col)) + geom_point()
}
```

ggplot2 also uses data masking. Always embrace column arguments inside `aes()`.
:::

### Your Turn: Exercise 4

Write a function `summarize_string_col` that accepts a string column name (not a bare name) and computes the mean of that column.

```{r}
#| code-fold: true
#| code-summary: "Solution"
summarize_string_col <- function(data, col_string) {
  data |>
    summarize(mean_val = mean(.data[[col_string]], na.rm = TRUE))
}

summarize_string_col(sales, "revenue")
summarize_string_col(sales, "units")
```

::: {.callout-tip title="Challenge"}
Write a function `multi_summarize` that accepts a data frame, a grouping column, and a character vector of column names. It should return the mean of each named column grouped by the grouping column. (Hint: use `across(all_of(...))`)

```{r}
#| code-fold: true
#| code-summary: "Solution"
multi_summarize <- function(data, group_col, col_names) {
  data |>
    summarize(
      across(all_of(col_names), \(x) mean(x, na.rm = TRUE)),
      .by = {{ group_col }}
    )
}

multi_summarize(sales, region, c("units", "revenue"))
```
:::

## Part 2: Functional Programming Patterns

### Function Factories

A function that returns a function:

```{r}
multiplier <- function(factor) {
  \(x) x * factor
}

double <- multiplier(2)
triple <- multiplier(3)

double(10)
triple(10)
```

The returned function "remembers" the value of `factor` from when it was created. This is a closure.

### Practical Factory: Threshold Checker

```{r}
make_checker <- function(lower, upper) {
  \(x) x >= lower & x <= upper
}

valid_revenue <- make_checker(10, 500)
valid_units <- make_checker(0, 100)

sales |>
  filter(valid_revenue(revenue), valid_units(units)) |>
  nrow()
```

### Function Operators

Functions that take a function and return a modified version:

```{r}
with_logging <- function(fn) {
  \(...) {
    cat("Input:", ..., "\n")
    result <- fn(...)
    cat("Output:", result, "\n")
    result
  }
}

logged_sum <- with_logging(sum)
logged_sum(1, 2, 3)
```

### safely() and possibly()

```{r}
safe_divide <- safely(\(x, y) x / y)

safe_divide(10, 2)
safe_divide(10, 0)
```

```{r}
maybe_divide <- possibly(\(x, y) x / y, otherwise = NA_real_)

map_dbl(c(2, 0, 5, NA), \(d) maybe_divide(10, d))
```

### reduce() and accumulate()

`reduce()` applies a 2-argument function across a list from left to right:

```{r}
reduce(1:5, `+`)     # ((((1+2)+3)+4)+5) = 15
reduce(1:5, `*`)     # 1*2*3*4*5 = 120
```

```{r}
# Merge multiple data frames
df1 <- tibble(id = 1:3, a = c("x", "y", "z"))
df2 <- tibble(id = 2:4, b = c(10, 20, 30))
df3 <- tibble(id = 1:4, c = c(TRUE, FALSE, TRUE, FALSE))

reduce(list(df1, df2, df3), inner_join, by = "id")
```

`accumulate()` returns all intermediate results:

```{r}
# Running total
transactions <- c(100, -30, 50, -10, 80)
accumulate(transactions, `+`)
```

### Your Turn: Exercise 5

Create a function factory `make_normalizer` that takes a vector and returns a function that normalizes new values to the 0-1 range using the original vector's min and max.

```{r}
#| code-fold: true
#| code-summary: "Solution"
make_normalizer <- function(reference) {
  rng <- range(reference, na.rm = TRUE)
  \(x) (x - rng[1]) / (rng[2] - rng[1])
}

norm_revenue <- make_normalizer(sales$revenue)
norm_revenue(c(100, 250, 500))
```

### Your Turn: Exercise 6

Use `reduce()` to find the intersection of these three vectors:

```{r}
v1 <- c(1, 2, 3, 4, 5)
v2 <- c(2, 4, 6, 8)
v3 <- c(3, 4, 5, 6)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
reduce(list(v1, v2, v3), intersect)
```

### Your Turn: Exercise 7

Use `accumulate()` to compute a running product of `c(2, 3, 4, 5)`.

```{r}
#| code-fold: true
#| code-summary: "Solution"
accumulate(c(2, 3, 4, 5), `*`)
```

### Your Turn: Exercise 8

Write a function operator `time_it` that wraps any function and prints how long it took to execute.

```{r}
#| code-fold: true
#| code-summary: "Solution"
time_it <- function(fn) {
  \(...) {
    start <- Sys.time()
    result <- fn(...)
    elapsed <- Sys.time() - start
    cat("Elapsed:", format(elapsed), "\n")
    result
  }
}

timed_mean <- time_it(mean)
timed_mean(rnorm(1e6))
```

::: {.callout-warning title="Common Mistake"}
Forgetting that `reduce()` needs a binary function (exactly 2 arguments):

```r
# WRONG
reduce(1:5, sqrt)  # sqrt takes 1 argument, reduce needs 2
```

The function passed to `reduce()` must take `(accumulated, next_element)`.
:::

::: {.callout-tip title="Challenge"}
Create a list of 5 transformation functions (e.g., add 10, multiply by 2, square, subtract 5, take sqrt). Use `reduce()` to compose them into a single function, then apply it to the number 3.

```{r}
#| code-fold: true
#| code-summary: "Solution"
transforms <- list(
  \(x) x + 10,
  \(x) x * 2,
  \(x) x^2,
  \(x) x - 5,
  \(x) sqrt(x)
)

composed <- reduce(transforms, \(f, g) \(x) g(f(x)))
composed(3)

# Verify step by step:
# 3 -> 13 -> 26 -> 676 -> 671 -> sqrt(671)
sqrt(671)
```
:::

## Part 3: APIs and JSON

### JSON Basics with jsonlite

```{r}
# Parse JSON string to R
json_str <- '{"name": "Alice", "age": 21, "scores": [90, 85, 78]}'
parsed <- fromJSON(json_str)
parsed
str(parsed)
```

```{r}
# Convert R object to JSON
my_list <- list(
  product = "Widget",
  price = 29.99,
  tags = c("sale", "popular")
)
toJSON(my_list, auto_unbox = TRUE, pretty = TRUE)
```

`auto_unbox = TRUE` prevents single values from becoming JSON arrays.

### Nested JSON to Data Frames

```{r}
json_array <- '[
  {"name": "Widget", "price": 29.99, "in_stock": true},
  {"name": "Gadget", "price": 49.99, "in_stock": false},
  {"name": "Doohickey", "price": 9.99, "in_stock": true}
]'

products <- fromJSON(json_array)
products
class(products)
```

An array of uniform JSON objects becomes a data frame automatically.

### Deeply Nested JSON

```{r}
nested_json <- '[
  {"name": "Alice", "address": {"city": "NY", "zip": "10001"}},
  {"name": "Bob", "address": {"city": "LA", "zip": "90001"}}
]'

# flatten = TRUE pulls nested fields up
fromJSON(nested_json, flatten = TRUE)
```

### httr2 Request Building

```{r}
#| eval: false
library(httr2)

# Build a request piece by piece
req <- request("https://httpbin.org/get") |>
  req_headers(Accept = "application/json") |>
  req_url_query(search = "widgets", page = 1)

# Inspect before sending
req
```

httr2 separates request construction from execution. This makes it easy to inspect and debug.

### Performing and Parsing

```{r}
#| eval: false
resp <- req |> req_perform()

# Status code
resp |> resp_status()

# Body as parsed JSON (list)
data <- resp |> resp_body_json()

# Body as text
text <- resp |> resp_body_string()
```

### API Key Security

```{r}
#| eval: false
# Store keys in .Renviron (never in scripts)
# MY_API_KEY=your_key_here

# Read at runtime
api_key <- Sys.getenv("MY_API_KEY")

req <- request("https://api.example.com/data") |>
  req_auth_bearer_token(api_key)
```

::: {.callout-warning title="Common Mistake"}
Hard-coding API keys in your script:

```r
# NEVER do this
req <- request("https://api.com") |>
  req_headers(Authorization = "Bearer sk-abc123secret")
```

If you share or commit this file, your key is exposed. Always use environment variables.
:::

### Your Turn: Exercise 9

Parse this JSON string into an R object and extract the second student's GPA:

```{r}
student_json <- '[
  {"name": "Alice", "gpa": 3.8, "major": "CS"},
  {"name": "Bob", "gpa": 3.2, "major": "Math"},
  {"name": "Carlos", "gpa": 3.5, "major": "CS"}
]'
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
students <- fromJSON(student_json)
students$gpa[2]
```

### Your Turn: Exercise 10

Convert this R data frame to a JSON string:

```{r}
orders <- tibble(
  id = 1:3,
  item = c("laptop", "mouse", "keyboard"),
  price = c(999.99, 29.99, 79.99)
)
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
toJSON(orders, pretty = TRUE)
```

### Your Turn: Exercise 11

Parse this nested JSON and flatten it into a single-level data frame:

```{r}
nested <- '[
  {"product": "Widget", "details": {"weight": 1.2, "color": "red"}},
  {"product": "Gadget", "details": {"weight": 0.8, "color": "blue"}}
]'
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
fromJSON(nested, flatten = TRUE)
```

### Your Turn: Exercise 12

Write a function `parse_api_response` that takes a JSON string representing an API response with a "data" field containing an array of records, and returns just the data as a tibble.

```{r}
test_response <- '{"status": "ok", "count": 2, "data": [{"id": 1, "value": 42}, {"id": 2, "value": 99}]}'
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
parse_api_response <- function(json_str) {
  parsed <- fromJSON(json_str)
  as_tibble(parsed$data)
}

parse_api_response(test_response)
```

### Think About It

You need to call an API 1,000 times (once per item in your dataset). The API allows 100 requests per minute. How would you structure this in R? What tools from today's session would you use?

::: {.callout-tip title="Challenge"}
Write a function `json_to_flat_tibble` that takes arbitrarily nested JSON (as a string) and returns a flat tibble. Handle the case where the JSON might be a single object or an array of objects.

```{r}
#| code-fold: true
#| code-summary: "Solution"
json_to_flat_tibble <- function(json_str) {
  parsed <- fromJSON(json_str, flatten = TRUE)
  if (is.data.frame(parsed)) {
    as_tibble(parsed)
  } else if (is.list(parsed)) {
    as_tibble(as.data.frame(parsed, stringsAsFactors = FALSE))
  } else {
    stop("Cannot convert to tibble")
  }
}

# Test with array
json_to_flat_tibble('[{"a": 1, "b": {"c": 2}}, {"a": 3, "b": {"c": 4}}]')

# Test with single object
json_to_flat_tibble('{"name": "test", "info": {"x": 1, "y": 2}}')
```
:::

## Session Synthesis

### Connect the Dots

Build a complete, reusable data analysis toolkit.

1. Write a function `explore_numeric` that accepts a data frame and a column name (bare), and returns a tibble with count, mean, median, sd, min, max. Use `{{ }}`.

2. Write a function factory `make_outlier_detector` that takes a reference vector and returns a function that flags values more than 2 SDs from the mean.

3. Parse this simulated API response and combine the results:

```{r}
api_pages <- list(
  '{"page": 1, "data": [{"id": 1, "score": 85}, {"id": 2, "score": 92}]}',
  '{"page": 2, "data": [{"id": 3, "score": 78}, {"id": 4, "score": 95}]}',
  '{"page": 3, "data": [{"id": 5, "score": 88}, {"id": 6, "score": 91}]}'
)
```

4. Use `map()` and `reduce()` to parse all pages and combine them into a single tibble.

```{r}
#| code-fold: true
#| code-summary: "Solution"
# 1. Explore numeric column
explore_numeric <- function(data, col) {
  data |>
    summarize(
      count = n(),
      mean = mean({{ col }}, na.rm = TRUE),
      median = median({{ col }}, na.rm = TRUE),
      sd = sd({{ col }}, na.rm = TRUE),
      min = min({{ col }}, na.rm = TRUE),
      max = max({{ col }}, na.rm = TRUE)
    )
}

explore_numeric(sales, revenue)

# 2. Outlier detector factory
make_outlier_detector <- function(reference) {
  mu <- mean(reference, na.rm = TRUE)
  sigma <- sd(reference, na.rm = TRUE)
  \(x) abs(x - mu) > 2 * sigma
}

is_outlier <- make_outlier_detector(sales$revenue)
sum(is_outlier(sales$revenue))

# 3 & 4. Parse and combine API pages
all_scores <- map(api_pages, \(page_json) {
  fromJSON(page_json)$data
}) |>
  reduce(bind_rows) |>
  as_tibble()

all_scores
```

### Key Takeaways

- `{{ }}` handles 90% of tidy evaluation needs. Use it whenever you pass column names to tidyverse functions.
- Function factories and operators let you build reusable, composable tools.
- `reduce()` and `accumulate()` solve sequential combination problems.
- jsonlite converts between R and JSON. httr2 handles HTTP requests.
- Always handle API errors and respect rate limits.

### Prepare for Next Session

Session 10 covers web scraping with rvest, a cumulative tidyverse practice session, and a full end-to-end data pipeline project. Review your dplyr, tidyr, ggplot2, and purrr skills -- you will use all of them.
